## The bias {#sec-dyn-bias}

**Question.** If last year's employment explains this year's, how wrong is
the within estimate of that persistence, and does more data fix it?

**Intuition.** Demeaning subtracts each firm's average over its sample
years, and that average contains every shock the firm had, including the
ones that built last year's employment. So the demeaned lag and the demeaned
error are correlated through those shocks, and negatively: the within
estimate of γ is too small, by about $(1 + \gamma)/(T - 1)$ [@nickell1981;
@croissant2019, p. 165]. The bias is of order $1/T$, so more firms do not
help and more years do [@baltagi2021, p. 188; @pesaran2015, p. 679]. Pooled
OLS without firm effects errs the other way: the firm effect sits in the
error and in the lag, so γ is pushed up [@croissant2019, p. 164]. The truth
lies between the two [@croissant2019, p. 168]. The random-effects estimator
quasi-demeans and is biased for the same reason [@pesaran2015, p. 678].

::: {.callout-tip appearance="simple"}
Where the bias comes from, term by term, and how GMM removes it:
[dynamic panel GMM](theory-dynamic-gmm.qmd).
:::

The exact large-N bias of the within estimator in the pure AR(1) with fixed
effects is [@pesaran2015, p. 679, eq. 27.11; @croissant2019, p. 165]

$$
\operatorname*{plim}_{N \to \infty}\left(\hat\gamma_{\text{within}}
  - \gamma\right)
  = -\frac{1 + \gamma}{T - 1}\,
    \frac{A}{1 - \dfrac{2\gamma A}{(1 - \gamma)(T - 1)}},
\qquad
A = 1 - \frac{1 - \gamma^{T}}{T(1 - \gamma)},
$$

where T is the number of periods in the regression after the lag is taken.
`nickell_bias(gamma, t)` in `R/simulate-panels.R` is this expression. The
leading factor $-(1 + \gamma)/(T - 1)$ is what the rest of the chapter means
by "the Nickell bias"; @pesaran2015 [p. 679] writes it more compactly as
$-(1 + \gamma)/T$ plus a term of order $T^{-2}$.

**The estimators.** No test here: two estimators on a known truth. Pooled OLS
(`plm::plm(model = "pooling")`) ignores the firm effect; the within estimator
(`model = "within"`) demeans it away. Both are inconsistent for fixed T.

**Run it.** `sim_dynamic_panel(n, t)` draws
$y_{it} = 0.5\, y_{i,t-1} + \alpha_i + \varepsilon_{it}$ for `n` firms with
$\alpha_i \sim N(0, 1)$ and $\varepsilon_{it} \sim N(0, 1)$, after a 50-period
burn-in, and keeps `t + 1` periods so that a regression on one lag uses
exactly `t`. First one draw of 100 firms at T = 10, both estimators in the
open:

```{r}
#| label: c06-1-one-draw
set.seed(1)
d <- sim_dynamic_panel(100, 10)
plm::plm(y ~ lag(y, 1), data = d, model = "pooling")
plm::plm(y ~ lag(y, 1), data = d, model = "within")
```

The truth is 0.5. OLS says 0.844, within says 0.367. One draw proves
nothing, so repeat 200 times at five values of T, and put the exact formula
beside the means:

```{r}
#| label: c06-1-grid
# One replication: the two calls above, kept for their slope on the lag.
fit_both <- function(t) {
  d <- sim_dynamic_panel(100, t)
  within <- plm::plm(y ~ lag(y, 1), data = d, model = "within")
  ols <- plm::plm(y ~ lag(y, 1), data = d, model = "pooling")
  c(within = unname(coef(within)), ols = unname(coef(ols)[2]))
}
set.seed(1)
t_grid <- c(5, 10, 20, 30, 50)
draws <- lapply(t_grid, function(t) replicate(200, fit_both(t)))
grid <- data.frame(
  T = t_grid,
  within = sapply(draws, function(m) mean(m["within", ])),
  within_sd = sapply(draws, function(m) sd(m["within", ])),
  ols = sapply(draws, function(m) mean(m["ols", ])),
  nickell = 0.5 + nickell_bias(0.5, t_grid),
  leading = 0.5 - 1.5 / (t_grid - 1)
)
round(grid, 3)
```

```{r}
#| label: c06-1-picture
#| fig-cap: "Left: mean estimate of γ over 200 draws against T, within
#|   (black) and pooled OLS (grey), with the exact Nickell curve dashed
#|   through the within points and the true 0.5 as a dotted line. Right: the
#|   200 within (black) and OLS (grey) estimates at T = 10 as histograms;
#|   the dotted line is the truth."
#| fig-alt: "Two panels. Left: a black line of points rises from 0.16 at T
#|   = 5 to 0.47 at T = 50 along a dashed curve, towards a dotted line at
#|   0.5. A grey line of points stays flat at 0.87 above it. Right: two
#|   separate histograms, a black one between 0.24 and 0.43 and a grey one
#|   between 0.82 and 0.92, with the dotted line at 0.5 in the empty gap
#|   between them."
op <- graphics::par(mfrow = c(1, 2))
plot(grid$T, grid$within, type = "b", pch = 16, ylim = c(0, 1),
     xlab = "periods in the regression, T", ylab = "mean estimate of gamma")
graphics::lines(grid$T, grid$ols, type = "b", pch = 16, col = "grey55")
graphics::lines(3:50, 0.5 + nickell_bias(0.5, 3:50), lty = 2)
graphics::abline(h = 0.5, lty = 3)
graphics::legend("bottomright", c("within", "pooled OLS", "exact bias curve"),
                 col = c("black", "grey55", "black"), pch = c(16, 16, NA),
                 lty = c(1, 1, 2), bty = "n")
at10 <- draws[[2]]
bins <- seq(0, 1, by = 0.01)
graphics::hist(at10["within", ], breaks = bins, col = "black",
               border = "white", xlim = c(0, 1),
               xlab = "estimate of gamma at T = 10", main = "")
graphics::hist(at10["ols", ], breaks = bins, col = "grey70",
               border = "white", add = TRUE)
graphics::abline(v = 0.5, lty = 3)
graphics::par(op)
```

The same two estimators on real data, where the truth is unknown. Pooled OLS
and the two-way (firm and year effects) within estimator of the
[effects tests](../05-core/index.qmd#sec-core-effects), on the Arellano–Bond
equation and on the one-lag equation:

```{r}
#| label: c06-1-emp
ab_f <- log(emp) ~ lag(log(emp), 1:2) + lag(log(wage), 0:1) + log(capital) +
  lag(log(output), 0:1)
one_f <- log(emp) ~ lag(log(emp), 1) + lag(log(wage), 0:1) +
  lag(log(capital), 0:1)
ab_ols <- plm::plm(ab_f, data = pe, model = "pooling")
ab_within <- plm::plm(ab_f, data = pe, model = "within", effect = "twoways")
one_ols <- plm::plm(one_f, data = pe, model = "pooling")
one_within <- plm::plm(one_f, data = pe, model = "within", effect = "twoways")
round(coef(summary(ab_ols))[2:3, 1:2], 3)
round(coef(summary(ab_within))[1:2, 1:2], 3)
round(coef(summary(one_ols))[2, 1:2], 3)
round(coef(summary(one_within))[1, 1:2], 3)
```

**Read it.**

- **Within, on the curve.** 0.164 at T = 5, 0.337 at T = 10, 0.469 at T =
  50, against the exact 0.169, 0.338 and 0.469. *So:* the formula is the
  bias, and the only cure is more years: at T = 30 the estimate is still 10%
  below the truth.
- **OLS, flat.** 0.872 or 0.873 at every T. *So:* years do not help OLS at
  all; the firm effect never leaves the error, and its share of the variance
  sets the bias.
- **The spread does not reach the truth.** At T = 10 the 200 within
  estimates run from 0.24 to 0.43 with standard deviation 0.031; not one
  reaches 0.5. *So:* this is bias, not noise, and averaging over draws or
  adding firms cannot remove it.
- **The real-data bracket.** On the Arellano–Bond equation γ₁ is 1.143 by
  OLS and 0.700 within, with γ₁ + γ₂ at 0.95 and 0.53; on the one-lag
  equation 0.962 and 0.626. *So:* the persistence of employment is somewhere
  inside each bracket, and nothing in this step says where.

::: {.callout-note appearance="simple"}
**Next.** Remove the firm effect by differencing instead of demeaning, and
find an instrument for the differenced lag
([difference and instrument](index.qmd#sec-dyn-ah)).
:::

::: {.callout-note collapse="true"}
## Recipe

```r
# d: a pdata.frame with outcome y; gamma: the true persistence.
plm::plm(y ~ lag(y, 1), data = d, model = "pooling")   # biased up
plm::plm(y ~ lag(y, 1), data = d, model = "within")    # biased down
nickell_bias(gamma, t)   # exact large-N bias of the within estimator
```
:::

::: {.callout-caution collapse="true"}
## Pitfalls

**Which T.** The T in the formula is the number of periods in the
regression, one fewer than the series has after the lag is taken.
`sim_dynamic_panel()` keeps `t + 1` periods for that reason; a first draft
of this step kept `t` and read 0.09 at "T = 5" where the formula said 0.17.

**The bias grows with γ.** @judson1999 found, in Monte Carlo at N = 20 or 100
and T up to 30, that the within bias "increases with ρ and decreases with
T", and "even for T = 30, this bias could be as much as 20% of the true
value" [@baltagi2021, p. 188]. The 10% error at γ = 0.5, T = 30 above is
the mild case.

**Bias-corrected within.** @kiviet1995 derived an approximation to this
bias and a within estimator that subtracts it [@baltagi2021, p. 188]. It
needs a balanced panel and treats the other regressors as exogenous
[@croissant2019, p. 166]. There is no R implementation in this book's
tools. Benchmark: Stata `xtlsdvc`, deferred.

**Random effects is no escape.** Quasi-demeaning introduces the same
correlation of order $1/T$, so RE is biased too [@pesaran2015, p. 678].
:::

::: {.callout-tip collapse="true"}
## Book check

@croissant2019 [p. 165] print the bias as −0.167 at T = 10 and γ = 0.5. That
is the leading factor, −1.5/9; the full expression they print on the same
page, `nickell_bias(0.5, 10)`, gives −0.162, and the 200-draw mean above is
0.337 − 0.5 = −0.163. The structure of the table of @nickell1981 is the
`nickell` column of the grid; the paper itself is not in `lit/`, so the
check is against the formula as @pesaran2015 [p. 679] prints it, not against
the table.

The two `plm::plm` calls reproduce Examples 7.2 and 7.3 of @croissant2019
[pp. 164–165] on `DemocracyIncome`: OLS with time effects 0.70637 (0.024293)
and two-way within 0.37863 (0.03344) on lagged democracy:

```{r}
#| label: c06-1-cm
data("DemocracyIncome", package = "pder")
cm_ols <- plm::plm(democracy ~ lag(democracy) + lag(income),
                   data = DemocracyIncome, index = c("country", "year"),
                   model = "within", effect = "time", subset = sample == 1)
cm_within <- update(cm_ols, effect = "twoways")
coef(summary(cm_ols))
coef(summary(cm_within))
```
:::
