## Difference and instrument {#sec-dyn-ah}

**Question.** With the firm effect differenced away, what can stand in for
last year's change in employment that does not carry this year's shock?

**Intuition.** Differencing removes $\alpha_i$ but makes the lag
$\Delta y_{i,t-1}$ share $\varepsilon_{i,t-1}$ with the error
$\Delta\varepsilon_{it}$. The level two periods back, $y_{i,t-2}$, is
correlated with $\Delta y_{i,t-1}$, since it is half of it, and not with
$\Delta\varepsilon_{it}$, as long as the $\varepsilon_{it}$ are not serially
correlated [@anderson1982; @baltagi2021, p. 188; @pesaran2015, p. 682]. Use
it as an instrument and the differenced equation can be estimated by
two-stage least squares: one instrument, one equation, consistent but not
efficient, because it uses one **moment condition**, a population average
the estimator sets to zero, where many hold [@baltagi2021, p. 188]. The
difference $\Delta y_{i,t-2}$ is a valid instrument too, but @arellano1989
found that version "has a singularity point and very large variances over a
significant range of parameter values", while the level "has no
singularities and much smaller variances" [@baltagi2021, p. 188];
@croissant2019 [p. 167] say the level "is often a much better instrument".
The instrument weakens as γ approaches one and fails at γ = 1 [@pesaran2015,
p. 682].

::: {.callout-tip appearance="simple"}
Why the within model cannot be instrumented this way, and the orthogonal
deviations alternative to differencing: [dynamic panel
GMM](theory-dynamic-gmm.qmd).
:::

The estimator rests on one moment condition per equation,

$$
E\left[ y_{i,t-2}\, \Delta\varepsilon_{it} \right] = 0,
\qquad t = 3, \dots, T,
$$

which holds because $y_{i,t-2}$ is built from shocks up to $t - 2$ and
$\Delta\varepsilon_{it}$ from the shocks at $t - 1$ and $t$ [@pesaran2015,
p. 682]. The first usable year is the third, since the lag costs one and the
difference another [@croissant2019, p. 168].

**The estimator.** Anderson–Hsiao [@anderson1982]: 2SLS on the differenced
equation, with $y_{i,t-2}$ (the level form) or $\Delta y_{i,t-2}$ (the
difference form) as the instrument for $\Delta y_{i,t-1}$ and the
differenced exogenous regressors as their own instruments. No null to
reject; the question is how wide the band is.

**Run it.** The one-lag equation of the chapter page, differenced inside the
formula, with the instruments after the `|`. `plm::plm` with a two-part
formula and `model = "pooling"` is the form of Example 7.4 of @croissant2019
[p. 168]; the Pitfalls box says why `model = "fd"` cannot be used. First the
level form:

```{r}
#| label: c06-2-ah-level
ah_level <- plm::plm(
  diff(log(emp)) ~ lag(diff(log(emp))) + diff(log(wage)) +
    lag(diff(log(wage))) + diff(log(capital)) + lag(diff(log(capital))) +
    year - 1 |
    lag(log(emp), 2) + diff(log(wage)) + lag(diff(log(wage))) +
    diff(log(capital)) + lag(diff(log(capital))) + year - 1,
  data = pe, model = "pooling")
coef(summary(ah_level))[1:5, ]
nobs(ah_level)
```

Then the difference form, the same call with `lag(diff(log(emp)), 2)` as the
instrument:

```{r}
#| label: c06-2-ah-diff
ah_diff <- plm::plm(
  diff(log(emp)) ~ lag(diff(log(emp))) + diff(log(wage)) +
    lag(diff(log(wage))) + diff(log(capital)) + lag(diff(log(capital))) +
    year - 1 |
    lag(diff(log(emp)), 2) + diff(log(wage)) + lag(diff(log(wage))) +
    diff(log(capital)) + lag(diff(log(capital))) + year - 1,
  data = pe, model = "pooling")
coef(summary(ah_diff))[1:5, ]
```

The Arellano–Bond equation has two lags of employment, so it needs two
instruments, $y_{i,t-2}$ and $y_{i,t-3}$, and is exactly identified (as
many instruments as endogenous regressors, none to spare):

```{r}
#| label: c06-2-ah-two
ah_two <- plm::plm(
  diff(log(emp)) ~ lag(diff(log(emp)), 1:2) + diff(log(wage)) +
    lag(diff(log(wage))) + diff(log(capital)) + diff(log(output)) +
    lag(diff(log(output))) + year - 1 |
    lag(log(emp), 2:3) + diff(log(wage)) + lag(diff(log(wage))) +
    diff(log(capital)) + diff(log(output)) + lag(diff(log(output))) +
    year - 1,
  data = pe, model = "pooling")
round(coef(summary(ah_two))[1:2, 1:2], 3)
```

And the pure AR(1), the level form with nothing but time dummies beside the
lag:

```{r}
#| label: c06-2-ar1
ah_ar1 <- plm::plm(
  diff(log(emp)) ~ lag(diff(log(emp))) + year - 1 |
    lag(log(emp), 2) + year - 1,
  data = pe, model = "pooling")
round(coef(summary(ah_ar1))[1, 1:2], 3)
```

```{r}
#| label: c06-2-picture
#| fig-cap: "The persistence of employment in the one-lag equation by four
#|   estimators, with 95% bands: pooled OLS, two-way within, Anderson–Hsiao
#|   with the level instrument, and with the difference instrument. The grey
#|   band is the OLS–within bracket of step 1."
#| fig-alt: "A dot-and-whisker chart with four rows. OLS at 0.96 and within
#|   at 0.63 have whiskers too short to see, and a grey vertical band
#|   spans the gap between them. Anderson–Hsiao with the level instrument
#|   sits at 1.23 with a whisker from minus 0.46 to 2.92, covering the whole
#|   band. With the difference instrument it sits at minus 0.11 with a
#|   whisker from minus 0.69 to 0.48, to the left of the band."
est <- c(`pooled OLS` = coef(summary(one_ols))[2, 1],
         `two-way within` = coef(summary(one_within))[1, 1],
         `AH, level y[t-2]` = coef(summary(ah_level))[1, 1],
         `AH, difference` = coef(summary(ah_diff))[1, 1])
se <- c(coef(summary(one_ols))[2, 2], coef(summary(one_within))[1, 2],
        coef(summary(ah_level))[1, 2], coef(summary(ah_diff))[1, 2])
op <- graphics::par(mar = c(4, 9, 1, 1))
plot(est, 4:1, xlim = c(-1, 3), yaxt = "n", pch = 16, ylab = "",
     xlab = "estimate of gamma, one-lag equation")
graphics::rect(est[2], 0, est[1], 5, col = "grey90", border = NA)
graphics::segments(est - 1.96 * se, 4:1, est + 1.96 * se, 4:1, lwd = 2)
graphics::points(est, 4:1, pch = 16)
graphics::axis(2, at = 4:1, labels = names(est), las = 1)
graphics::par(op)
```

**Read it.**

- **Level instrument.** 1.23 with standard error 0.86: a band from −0.46 to
  2.92. *So:* consistent and uninformative; the band holds the whole
  OLS–within bracket, and the point estimate above one means nothing.
- **Difference instrument.** −0.11 with standard error 0.30: a band from
  −0.69 to 0.48, below the bracket. *So:* the version Arellano warned of; a
  point below within and a band that crosses zero.
- **Two lags, two instruments.** 11.1 (58.5) and −2.3 (12.2). *So:* with
  nothing to spare, the estimator has no precision at all; this equation
  needs more instruments than years.
- **Misleads (the pure AR(1)).** 1.18 with standard error 1.23. *So:* a
  reader who prints the point reads explosive employment; the standard
  error says the number carries no information, and
  [step 3](index.qmd#sec-dyn-gmm) says why.

::: {.callout-note appearance="simple"}
**Next.** There are more valid instruments than one: every earlier level is
valid for every later difference
([difference and system GMM](index.qmd#sec-dyn-gmm)).
:::

::: {.callout-note collapse="true"}
## Recipe

```r
# d: a pdata.frame; y the outcome, x an exogenous regressor; level form.
plm::plm(diff(y) ~ lag(diff(y)) + diff(x) + year - 1 |
           lag(y, 2) + diff(x) + year - 1,
         data = d, model = "pooling")
```
:::

::: {.callout-caution collapse="true"}
## Pitfalls

**`model = "fd"` differences the instruments too.** With a two-part formula,
`plm::plm(model = "fd")` applies the first-difference transformation to both
parts, so `lag(log(emp), 2)` in the instrument part becomes
$\Delta y_{i,t-2}$, not the level. The only form that keeps the level is
`diff()` inside the formula with `model = "pooling"`, as above.

**Two rows per gap.** For every year a firm is missing, the differenced
equation loses two observations, that year and the next [@croissant2019,
p. 166]. The level-form fit above uses 751 firm-years of the 1031.

**T must stay small relative to N.** The differenced error is MA(1), a
first-order moving average: each differenced error shares one shock with
its neighbour. Ignoring that structure "leads to inconsistency of the IV
estimator if T/N → c > 0" [@pesaran2015, p. 682], a result of @alvarez2003
quoted there and not reproduced here.
:::

::: {.callout-tip collapse="true"}
## Book check

@croissant2019 [p. 168, Example 7.4] print 0.4687 (0.1182) on the lagged
difference of democracy and −0.1036 (0.3049) on the lagged difference of
income for the Anderson–Hsiao estimator on `DemocracyIncome`; the chunk
below is that call and reproduces both to the printed digits. No source on
disk prints an Anderson–Hsiao estimate on `EmplUK`, so the four fits above
have no published target.

```{r}
#| label: c06-2-cm
cm_ah <- plm::plm(
  diff(democracy) ~ lag(diff(democracy)) + lag(diff(income)) + year - 1 |
    lag(democracy, 2) + lag(income, 2) + year - 1,
  data = DemocracyIncome, index = c("country", "year"),
  model = "pooling", subset = sample == 1)
coef(summary(cm_ah))[1:2, ]
```
:::
