## Difference and system GMM {#sec-dyn-gmm}

**Question.** Using every lagged level a firm offers, how persistent is
employment, and when do those instruments stop carrying information?

**Intuition.** In the third year one level is a valid instrument, $y_{i1}$;
in the fourth, $y_{i1}$ and $y_{i2}$; and so on, "adding an extra valid
instrument with each forward period" [@baltagi2021, p. 190]. The instrument
matrix of a firm is block-diagonal with a block that grows by one column per
period [@holtzeakin1988; @baltagi2021, p. 190, eq. 8.6; @croissant2019,
p. 168, eq. 7.4]. With more moment conditions than parameters, GMM picks the
estimate that makes the weighted sum of squared sample moments smallest
[@hansen1982; @croissant2019, p. 169]. The **one-step** estimator weights
them by the known MA(1) covariance of a differenced error [@baltagi2021,
p. 190, eq. 8.8; @croissant2019, p. 169]; the **two-step** estimator
re-weights them by a covariance built from the one-step residuals
[@baltagi2021, p. 191, eq. 8.9; @croissant2019, p. 171]. The two are
asymptotically equivalent when the shocks are homoskedastic [@baltagi2021,
p. 191]. The trap is a persistent series: when γ is near one, $y_{i,t-2}$
barely predicts $\Delta y_{i,t-1}$, and the instruments are **weak**
[@blundell1998; @baltagi2021, p. 201; @croissant2019, p. 174]. @blundell1998
add the levels equation, instrumented by the lagged difference, which stays
informative when the lagged levels do not [@baltagi2021, p. 202]: that is
**system GMM**.

::: {.callout-tip appearance="simple"}
The instrument matrix, the two weighting matrices, the first-stage algebra
of the weak-instrument case and the redundancy argument behind the system
moments: [dynamic panel GMM](theory-dynamic-gmm.qmd).
:::

Two equations carry the step. The weak-instrument case is one first-stage
regression: with T = 3 the only instrument is $y_{i1}$, the first stage is
$\Delta y_{i2}$ on $y_{i1}$, and its slope has the limit [@blundell1998;
@baltagi2021, p. 201, eq. 8.34; @croissant2019, p. 174, eq. 7.19]

$$
\operatorname{plim}\,\hat\pi
  = (\gamma - 1)\,\frac{c}{c + \sigma_\alpha^2 / \sigma_\varepsilon^2},
\qquad c = \frac{1 - \gamma}{1 + \gamma},
$$

which goes to zero as $\gamma \to 1$, and also as the firm effect's variance
grows relative to the shock's. The system estimator adds to the difference
moments $E[y_{i,t-s}\,\Delta\varepsilon_{it}] = 0$, $s \ge 2$, the level
moments

$$
E\left[\Delta y_{i,t-1}\,(\alpha_i + \varepsilon_{it})\right] = 0,
\qquad t = 3, \dots, T,
$$

valid when each firm's distance from its steady state at the start of the
sample is uncorrelated with its effect [@blundell1998; @baltagi2021, p. 202;
@croissant2019, p. 176]. For T = 4 and a firm-effect variance equal to the
shock variance, the asymptotic variance of difference GMM is 1.75 times that
of system GMM at γ = 0, 3.26 times at γ = 0.5 and 55.4 times at γ = 0.9
[@blundell1998; @baltagi2021, p. 202].

**The estimators.** Difference GMM, one-step and two-step [@arellano1991];
system GMM [@arellano1995; @blundell1998]. In `plm::pgmm` the formula has
three parts, `y ~ regressors | GMM instruments | normal instruments`;
`lag(log(emp), 2:99)` means every lag from the second; `effect = "twoways"`
adds the time dummies as normal instruments; `transformation = "ld"` is the
system estimator [@croissant2019, pp. 171, 177]. Everything in the normal
part defaults to the regressors not used as GMM instruments.

**Run it.** The Arellano–Bond equation with every lag of employment from the
second as a GMM instrument. One-step first, with the robust (to
heteroskedasticity and within-firm correlation) covariance of the
[robust errors](../05-core/index.qmd#sec-core-robust-se) step:

```{r}
#| label: c06-3-ab-onestep
ab_g <- log(emp) ~ lag(log(emp), 1:2) + lag(log(wage), 0:1) + log(capital) +
  lag(log(output), 0:1) | lag(log(emp), 2:99)
ab_one <- plm::pgmm(ab_g, data = pe, effect = "twoways", model = "onestep")
summary(ab_one, robust = TRUE)
```

Two-step, printed twice: with the conventional two-step covariance, which is
the paper's table, and then with the corrected one, which is the one to
report. [Step 4](index.qmd#sec-dyn-diagnostics) is about the gap between
them:

```{r}
#| label: c06-3-ab-twostep
ab_two <- plm::pgmm(ab_g, data = pe, effect = "twoways", model = "twosteps")
summary(ab_two, robust = FALSE)
summary(ab_two, robust = TRUE)
```

The fit keeps its instrument matrix, one per firm, in `W`. Its column count
is the number of instruments, time dummies and exogenous regressors
included:

```{r}
#| label: c06-3-count
ncol(ab_two$W[[1]])
nobs(ab_two)
```

The weak-instrument case: the pure AR(1) of log employment, by OLS, within,
two-step difference GMM and two-step system GMM, with a Sargan test, which
asks whether the spare moment conditions hold
([diagnostics](index.qmd#sec-dyn-diagnostics)), on each GMM fit:

```{r}
#| label: c06-3-weak
ar1_ols <- plm::plm(log(emp) ~ lag(log(emp)), data = pe, model = "pooling")
ar1_within <- plm::plm(log(emp) ~ lag(log(emp)), data = pe,
                       model = "within", effect = "twoways")
ar1_dif <- plm::pgmm(log(emp) ~ lag(log(emp)) | lag(log(emp), 2:99),
                     data = pe, effect = "twoways", model = "twosteps")
ar1_sys <- plm::pgmm(log(emp) ~ lag(log(emp)) | lag(log(emp), 2:99),
                     data = pe, effect = "twoways", model = "twosteps",
                     transformation = "ld")
round(coef(ar1_ols)[2], 3)
round(coef(ar1_within), 3)
round(coef(summary(ar1_dif))[1, 1:2], 3)
plm::sargan(ar1_dif)
round(coef(summary(ar1_sys))[1, 1:2], 3)
plm::sargan(ar1_sys)
```

System GMM. First the one-lag equation in the form of the `?pgmm`
Blundell–Bond example, every lag from the second of all three variables as
GMM instruments, one-step and then two-step; then the Arellano–Bond equation,
one-step and two-step:

```{r}
#| label: c06-3-sys
bb_g <- log(emp) ~ lag(log(emp), 1) + lag(log(wage), 0:1) +
  lag(log(capital), 0:1) | lag(log(emp), 2:99) + lag(log(wage), 2:99) +
  lag(log(capital), 2:99)
bb_one <- plm::pgmm(bb_g, data = pe, effect = "twoways", model = "onestep",
                    transformation = "ld")
summary(bb_one, robust = TRUE)
bb_two <- plm::pgmm(bb_g, data = pe, effect = "twoways", model = "twosteps",
                    transformation = "ld")
round(coef(summary(bb_two, robust = TRUE))[1:5, 1:2], 4)
plm::sargan(bb_two)
ab_sys1 <- plm::pgmm(ab_g, data = pe, effect = "twoways", model = "onestep",
                     transformation = "ld")
round(coef(summary(ab_sys1, robust = TRUE))[1:2, 1:2], 3)
ab_sys <- plm::pgmm(ab_g, data = pe, effect = "twoways", model = "twosteps",
                    transformation = "ld")
round(coef(summary(ab_sys, robust = TRUE))[1:7, 1:2], 3)
plm::sargan(ab_sys)
ncol(ab_sys$W[[1]])
```

The bracket on the one-lag equation: difference GMM, one-step and two-step,
to sit beside the OLS and within fits of [step 1](index.qmd#sec-dyn-bias)
and the system fits above:

```{r}
#| label: c06-3-bracket
one_dif1 <- plm::pgmm(bb_g, data = pe, effect = "twoways", model = "onestep")
one_dif2 <- plm::pgmm(bb_g, data = pe, effect = "twoways", model = "twosteps")
round(rbind(
  `pooled OLS` = coef(summary(one_ols))[2, 1:2],
  `two-way within` = coef(summary(one_within))[1, 1:2],
  `DIF, one-step` = coef(summary(one_dif1, robust = TRUE))[1, 1:2],
  `DIF, two-step` = coef(summary(one_dif2, robust = TRUE))[1, 1:2],
  `SYS, one-step` = coef(summary(bb_one, robust = TRUE))[1, 1:2],
  `SYS, two-step` = coef(summary(bb_two, robust = TRUE))[1, 1:2]), 3)
plm::sargan(one_dif2)
ncol(one_dif2$W[[1]])
```

```{r}
#| label: c06-3-picture
#| fig-cap: "Left: the persistence of employment in the one-lag equation
#|   as a ladder, within at the bottom, OLS at the top, the four GMM fits
#|   between, each with its 95% band; the grey background is the
#|   Anderson–Hsiao band of step 2, which covers the whole axis. Right: the
#|   instrument matrix of one firm observed for nine years in the
#|   Arellano–Bond fit, black where an entry is non-zero: six differenced
#|   equations in rows, 38 instruments in columns, the staircase of lagged
#|   levels on the left and the regressors and time dummies on the right."
#| fig-alt: "Two panels. Left: a dot-and-whisker chart on a grey
#|   background with six rows. From bottom to top: pooled OLS at 0.96 with
#|   a tiny whisker, SYS two-step at 0.93 and SYS one-step at 0.94 with
#|   whiskers of about 0.05 each side, DIF two-step at 0.68 and DIF
#|   one-step at 0.71 with whiskers of about 0.17 each side, and two-way
#|   within at 0.63 with a whisker of 0.06. Right: a six-row block pattern.
#|   Row one has two black cells at the far left, row two three, up to row
#|   six with seven, forming a staircase; a block of black cells fills the
#|   rightmost columns of every row."
lad <- rbind(
  `pooled OLS` = coef(summary(one_ols))[2, 1:2],
  `SYS, one-step` = coef(summary(bb_one, robust = TRUE))[1, 1:2],
  `SYS, two-step` = coef(summary(bb_two, robust = TRUE))[1, 1:2],
  `DIF, one-step` = coef(summary(one_dif1, robust = TRUE))[1, 1:2],
  `DIF, two-step` = coef(summary(one_dif2, robust = TRUE))[1, 1:2],
  `two-way within` = coef(summary(one_within))[1, 1:2])
ah <- coef(summary(ah_level))[1, 1:2]
op <- graphics::par(mfrow = c(1, 2), mar = c(4, 7, 1, 1))
plot(lad[, 1], 6:1, xlim = c(0.4, 1.2), yaxt = "n", pch = 16, ylab = "",
     xlab = "gamma, one-lag equation")
graphics::rect(ah[1] - 1.96 * ah[2], 0, ah[1] + 1.96 * ah[2], 7,
               col = "grey92", border = NA)
graphics::segments(lad[, 1] - 1.96 * lad[, 2], 6:1,
                   lad[, 1] + 1.96 * lad[, 2], 6:1, lwd = 2)
graphics::points(lad[, 1], 6:1, pch = 16)
graphics::axis(2, at = 6:1, labels = rownames(lad), las = 1)
w <- ab_two$W[["127"]]
graphics::image(seq_len(ncol(w)), seq_len(nrow(w)),
                1 * t(w != 0)[, nrow(w):1], col = c("white", "black"),
                xlab = "instrument column", ylab = "", yaxt = "n")
graphics::axis(2, at = nrow(w):1, labels = rownames(w), las = 1)
graphics::par(op)
```

**Read it.**

- **Two-step difference GMM, the Arellano–Bond equation.** γ₁ = 0.474, γ₂ =
  −0.053; the wage −0.51 now and +0.22 lagged, capital 0.29, output 0.61 now
  and −0.45 lagged. One-step gives 0.535. *So:* the coefficient the help
  page names as the paper's Table 4(b); the long-run persistence
  γ₁ + γ₂ is 0.42.
- **The one-lag equation, inside the bracket.** Difference GMM 0.68 to 0.71,
  system GMM 0.93 to 0.94, between within at 0.63 and OLS at 0.96. *So:*
  both consistent estimators land where the theory says, yet 0.25 apart,
  and the Sargan tests accept both (p = 0.21 and 0.22). The instruments
  decide, not the data.
- **The Arellano–Bond equation, outside it.** Difference GMM 0.47 sits
  *below* within 0.70, and system GMM 1.16 above OLS 1.14. *So:* the
  symptom of weak lagged-level instruments on a persistent series, where
  the level restrictions "remain informative in cases where first
  differenced instruments become weak" [@baltagi2021, p. 202]; read 0.47 as
  a warning, not a result, and note that the system fit's Sargan rejects
  (p = 0.018) with 51 instruments.
- **Misleads (the pure AR(1)).** OLS 0.997, within 0.744, difference GMM
  0.31 (0.16) with Sargan p = 0.030, system GMM 1.09 (0.04) with Sargan
  p = 0.0002. *So:* a series this persistent breaks the difference
  instruments and the system restriction alike; this is a unit-root
  problem, not a dynamic-panel problem
  ([why it matters](../04-time/index.qmd#sec-ur-why)).

::: {.callout-note appearance="simple"}
**Next.** Which of these fits can be believed: the spare moment conditions,
the serial correlation of the residuals, the standard errors and the
instrument count ([diagnostics](index.qmd#sec-dyn-diagnostics)).
:::

::: {.callout-note collapse="true"}
## Recipe

```r
# d: a pdata.frame; plm attached, because pgmm() evaluates a plm() call.
library(plm)
f <- y ~ lag(y, 1:2) + x | lag(y, 2:99)          # x exogenous, own instrument
plm::pgmm(f, data = d, effect = "twoways", model = "twosteps")   # difference
plm::pgmm(f, data = d, effect = "twoways", model = "twosteps",
          transformation = "ld")                                # system
```
:::

::: {.callout-caution collapse="true"}
## Pitfalls

**Defaults that differ from `plm`.** `pgmm(effect = )` defaults to
`"twoways"` where `plm` defaults to `"individual"`; with
`effect = "individual"` the Arellano–Bond equation gives γ₁ = 0.449 (0.183),
not 0.474:

```{r}
#| label: c06-3-individual
ab_ind <- plm::pgmm(ab_g, data = pe, effect = "individual",
                    model = "twosteps")
round(coef(summary(ab_ind, robust = TRUE))[1, 1:2], 3)
```

`model` defaults to `"onestep"`; the paper's table is `"twosteps"`. And
`pgmm` fails with "could not find function" unless `plm` is attached,
because it builds and evaluates a `plm()` call.

**Two Sargan lines on one instrument set.** The one-step summary prints
Sargan 44.6 (p = 0.009) and the two-step summary 30.1 (p = 0.22) for the same
38 instruments. [Step 4](index.qmd#sec-dyn-diagnostics) reads the two-step
statistic, the value of the two-step objective function at the estimate
[@croissant2019, p. 180].

**System counts are not difference counts.** The system fit adds an
intercept, one more period and one level moment per period, so its
instrument count and Sargan degrees of freedom cannot be compared with the
difference fit's [@croissant2019, pp. 177, 180]: 51 against 38 on the
Arellano–Bond equation, 113 against 91 on the one-lag one.

**System GMM has its own weak instruments.** @bun2010 show the system
estimator keeps a weak-instrument problem that grows with the variance
ratio of the firm effect to the shock, and that its Wald tests over-reject
further as that ratio grows [@baltagi2021, p. 202]. @hayakawa2009 shows the
other side: when the initial condition makes the series mean-nonstationary
and the firm-effect variance is large, difference GMM does well and the
system restriction fails [@baltagi2021, pp. 202–203].

**Moments plm leaves out.** @ahn1995 derive T − 2 further nonlinear moment
conditions from the same assumptions [@baltagi2021, p. 189; @pesaran2015,
p. 685]. `plm` does not use them; **pdynmc** [@fritsch2021] does. Its
two-step on the `?pgmm` Table 4(a2) equation equals `pgmm`'s, and with
`use.mc.nonlin = TRUE` it adds the Ahn–Schmidt conditions; that fit took
2.7 minutes on this machine:

```{r}
#| label: c06-3-pdynmc
#| results: hide
#| message: false
# Results hidden: pdynmc prints progress lines; the next chunk reads the fits.
ab_a2 <- plm::pgmm(log(emp) ~ lag(log(emp), 1:2) + lag(log(wage), 0:1) +
  lag(log(capital), 0:2) + lag(log(output), 0:2) | lag(log(emp), 2:99),
  data = pe, effect = "twoways", model = "twosteps")
data("ABdata", package = "pdynmc")
abd <- ABdata
abd[, 4:7] <- log(abd[, 4:7])
pd_two <- pdynmc::pdynmc(
  dat = abd, varname.i = "firm", varname.t = "year",
  use.mc.diff = TRUE, use.mc.lev = FALSE, use.mc.nonlin = FALSE,
  include.y = TRUE, varname.y = "emp", lagTerms.y = 2,
  fur.con = TRUE, fur.con.diff = TRUE, fur.con.lev = FALSE,
  varname.reg.fur = c("wage", "capital", "output"),
  lagTerms.reg.fur = c(1, 2, 2),
  include.dum = TRUE, dum.diff = TRUE, dum.lev = FALSE, varname.dum = "year",
  w.mat = "iid.err", std.err = "corrected", estimation = "twostep",
  opt.meth = "none")
pd_nl <- pdynmc::pdynmc(
  dat = abd, varname.i = "firm", varname.t = "year",
  use.mc.diff = TRUE, use.mc.lev = FALSE, use.mc.nonlin = TRUE,
  include.y = TRUE, varname.y = "emp", lagTerms.y = 2,
  fur.con = TRUE, fur.con.diff = TRUE, fur.con.lev = FALSE,
  varname.reg.fur = c("wage", "capital", "output"),
  lagTerms.reg.fur = c(1, 2, 2),
  include.dum = TRUE, dum.diff = TRUE, dum.lev = FALSE, varname.dum = "year",
  w.mat = "iid.err", std.err = "corrected", estimation = "twostep",
  opt.meth = "BFGS")
```

```{r}
#| label: c06-3-pdynmc-read
round(cbind(pdynmc = coef(pd_two)[1:2], pgmm = coef(ab_a2)[1:2]), 4)
round(cbind(pdynmc = pd_two$stderr$step2[1:2],
            pgmm = sqrt(diag(plm::vcovHC(ab_a2)))[1:2]), 4)
summary(pd_nl)
```

Linear moments only: 0.6287 (0.1934) from both packages. With the
Ahn–Schmidt moments added: γ₁ = 0.665 (0.200), γ₂ = −0.084 (0.052), the wage
−0.50 (0.15), capital 0.26 (0.07), and a J test (pdynmc's name for the
Sargan–Hansen statistic of [step 4](index.qmd#sec-dyn-diagnostics)) of
40.35 on 29 degrees of freedom, p = 0.078. The extra moments move γ₁ up by
0.04 and do not narrow its band.

**Time-invariant regressors.** Differencing removes them with the firm
effect. @arellano1995 set the estimator in the Hausman–Taylor frame
[@hausman1981], a random-effects IV estimator that uses the exogenous
regressors as instruments
([fixed or random](../05-core/index.qmd#sec-core-fe-re)), where some level
instruments identify such coefficients [@baltagi2021, p. 194], and
@kripfganz2019 give a two-step estimator for them. Neither is in this
book's tools. Benchmark: Stata `xtseqreg`, deferred.
:::

::: {.callout-tip collapse="true"}
## Book check

`?pgmm` annotates `ab.b`, the two-step difference fit of the Arellano–Bond
equation with `robust = FALSE`, as "Arellano and Bond (1991), table 4
col. b" [@arellano1991], the table @baltagi2021 [p. 223] names as "Table 4
of Arellano and Bond (1991) (p. 290)"; `c06-3-ab-twostep` above is that
call and prints
0.474151 (0.085303), −0.052967 (0.027284), −0.513205 (0.049345), 0.224640
(0.080063), 0.292723 (0.039463), 0.609775 (0.108524), −0.446373 (0.124815),
Sargan 30.11 on 25 degrees of freedom. `?pgmm` annotates `bb.4`, the
one-step system fit of the one-lag equation with `robust = TRUE`, as
"Blundell and Bond (1998) table 4 (cf. DPD for OX p. 12 col. 4)"
[@blundell1998]; `c06-3-sys` above prints 0.935605 (0.026295). Neither
paper is in `lit/`, so what is matched here is the help page's annotation,
not a page of the paper.

Example 7.5 of @croissant2019 [pp. 171–172] prints 0.50499 (0.09049) for
the one-step and 0.554007 (0.10783) for the two-step difference fit, and
Example 7.7 [@croissant2019, p. 177] prints 0.6176 (0.05714) for the
two-step system fit, all on `DemocracyIncome`. All three reproduce to the
printed digits:

```{r}
#| label: c06-3-cm
cm_diff1 <- plm::pgmm(democracy ~ lag(democracy) + lag(income) |
                        lag(democracy, 2:99) | lag(income, 2),
                      data = DemocracyIncome, index = c("country", "year"),
                      model = "onestep", effect = "twoways",
                      subset = sample == 1)
cm_diff2 <- update(cm_diff1, model = "twosteps")
cm_sys2 <- plm::pgmm(democracy ~ lag(democracy) + lag(income) |
                       lag(democracy, 2:99) | lag(income, 2),
                     data = DemocracyIncome, index = c("country", "year"),
                     model = "twosteps", effect = "twoways",
                     transformation = "ld")
coef(summary(cm_diff1))
coef(summary(cm_diff2))
coef(summary(cm_sys2))
```
:::
