## Diagnostics {#sec-dyn-diagnostics}

**Question.** The fit prints a Sargan test, two autocorrelation tests and
two sets of standard errors: which numbers decide whether 0.47 is an
estimate or an artefact?

**Intuition.** With J instruments and K coefficients, the estimator needs K
moment conditions and has J − K to spare. The **Sargan test** asks whether
those spare moments are also near zero at the estimate; if the instruments
are valid they should be [@sargan1958; @hansen1982; @croissant2019, p. 180;
@baltagi2021, p. 191]. With too many of them it stops being able to tell:
@bowsher2002 found that at N = 100 and T = 15 the statistic's Monte Carlo
variance is 13.7 where the χ² says 180, so it "never rejects when T is too
large for a given N" [@baltagi2021, p. 192]. The **autocorrelation tests**
ask whether the original shocks are serially uncorrelated, which is what
makes $y_{i,t-2}$ a valid instrument: consistency "relies upon the fact that
$E[\Delta\nu_{it}\Delta\nu_{i,t-2}] = 0$" [@baltagi2021, p. 192]. A
differenced white-noise error is MA(1), so first-order correlation of the
differenced residuals is expected and second-order is not: m1 should reject
and m2 should not [@baltagi2021, p. 192; @croissant2019, p. 182]. The
**two-step standard errors** are built from a weight matrix estimated with
one-step residuals and understate the variance in small samples, so Wald
tests are oversized (they reject a true null more often than their nominal
level); @windmeijer2005 traced the bias to that estimation and derived the
correction [@baltagi2021, p. 193; @croissant2019, p. 179].
Finally, the **instrument count** is a choice. @ziliak1997 saw an
elasticity fall from 0.519 to 0.093 as the moment conditions rose from 9 to
212, with the standard error dropping from 0.36 to 0.07 [@baltagi2021,
p. 194]; @roodman2009note made the case against too many instruments and
`collapse` is the fix in software [@roodman2009; @baltagi2021, p. 194;
@croissant2019, p. 173].

::: {.callout-tip appearance="simple"}
The Sargan statistic as the GMM objective, the $a_l$ and $b_l$ of the
autocorrelation test, and the terms Windmeijer's expansion adds:
[dynamic panel GMM](theory-dynamic-gmm.qmd).
:::

Two statistics to read. The Sargan statistic is the two-step objective
function at the estimate, with $\hat{\Delta\varepsilon}$ the two-step
residuals and $W$ the instruments [@croissant2019, p. 180; @baltagi2021,
p. 191]:

$$
S = \left(\sum_i W_i'\hat{\Delta\varepsilon}_i\right)'
    \left(\sum_i W_i'\hat{\Delta\varepsilon}_i
          \hat{\Delta\varepsilon}_i' W_i\right)^{-1}
    \left(\sum_i W_i'\hat{\Delta\varepsilon}_i\right)
  \;\sim\; \chi^2_{J - K} .
$$

The autocorrelation test of order $l$ divides the sample covariance of the
differenced residuals with their own $l$-th lag by a standard error that
allows for the estimated coefficients [@arellano1991; @croissant2019,
pp. 181–182]:

$$
m_l = \frac{a_l}{\sqrt{b_l}},
\qquad
a_l = \frac{1}{\sqrt N}\,\hat{\Delta\varepsilon}'\hat{\Delta\varepsilon}_{-l},
\qquad m_l \sim N(0, 1) .
$$

**The tests.** Sargan–Hansen [@sargan1958; @hansen1982], `plm::sargan`.
Null: the overidentifying moment conditions hold. A rejection means at
least one instrument is correlated with the differenced error. Arellano–Bond
m1 and m2 [@arellano1991], `plm::mtest`. Null: no serial correlation of that
order in the differenced residuals. m2 rejecting means the levels error is
AR(1) and the instruments from $t - 2$ fail; m1 is expected to reject
[@croissant2019, p. 182]. Windmeijer's corrected covariance
[@windmeijer2005], `plm::vcovHC` and `summary(robust = TRUE)`.

**Run it.** All on `ab_two`, the two-step difference fit of the
Arellano–Bond equation from [step 3](index.qmd#sec-dyn-gmm). The Sargan
test:

```{r}
#| label: c06-4-sargan
plm::sargan(ab_two)
```

First-order autocorrelation, with the conventional and then the corrected
covariance:

```{r}
#| label: c06-4-m1
plm::mtest(ab_two, order = 1L)
plm::mtest(ab_two, order = 1L, vcov = plm::vcovHC)
```

Second order, the one that matters:

```{r}
#| label: c06-4-m2
plm::mtest(ab_two, order = 2L)
plm::mtest(ab_two, order = 2L, vcov = plm::vcovHC)
```

The same pair on the one-step fit with the robust covariance, which is the
form `?mtest` pairs with Table 2 of @windmeijer2005:

```{r}
#| label: c06-4-mtest-onestep
plm::mtest(ab_one, order = 1L, vcov = plm::vcovHC)
plm::mtest(ab_one, order = 2L, vcov = plm::vcovHC)
```

The two standard errors side by side, the form of Example 7.8 of
@croissant2019 [p. 179]:

```{r}
#| label: c06-4-windmeijer
round(rbind(twostep = sqrt(diag(vcov(ab_two))),
            windmeijer = sqrt(diag(plm::vcovHC(ab_two))))[, 1:7], 4)
```

The instrument count. Five two-step fits of the same equation, each `pgmm`
call written out: lags 2 to 3, 2 to 4, 2 to 6, every lag, and every lag
collapsed; then one table built from the five objects:

```{r}
#| label: c06-4-count
m_23 <- plm::pgmm(log(emp) ~ lag(log(emp), 1:2) + lag(log(wage), 0:1) +
  log(capital) + lag(log(output), 0:1) | lag(log(emp), 2:3),
  data = pe, effect = "twoways", model = "twosteps")
m_24 <- plm::pgmm(log(emp) ~ lag(log(emp), 1:2) + lag(log(wage), 0:1) +
  log(capital) + lag(log(output), 0:1) | lag(log(emp), 2:4),
  data = pe, effect = "twoways", model = "twosteps")
m_26 <- plm::pgmm(log(emp) ~ lag(log(emp), 1:2) + lag(log(wage), 0:1) +
  log(capital) + lag(log(output), 0:1) | lag(log(emp), 2:6),
  data = pe, effect = "twoways", model = "twosteps")
m_all <- plm::pgmm(log(emp) ~ lag(log(emp), 1:2) + lag(log(wage), 0:1) +
  log(capital) + lag(log(output), 0:1) | lag(log(emp), 2:99),
  data = pe, effect = "twoways", model = "twosteps")
m_coll <- plm::pgmm(log(emp) ~ lag(log(emp), 1:2) + lag(log(wage), 0:1) +
  log(capital) + lag(log(output), 0:1) | lag(log(emp), 2:99),
  data = pe, effect = "twoways", model = "twosteps", collapse = TRUE)
fits <- list(`2:3` = m_23, `2:4` = m_24, `2:6` = m_26, `2:99` = m_all,
             `2:99 collapsed` = m_coll)
# Plumbing: one row per fitted object, nothing estimated here.
count_tab <- t(sapply(fits, function(m) {
  s <- plm::sargan(m)
  c(instruments = ncol(m$W[[1]]),
    gamma1 = unname(coef(m)[1]),
    se_twostep = unname(sqrt(diag(vcov(m)))[1]),
    se_windmeijer = unname(sqrt(diag(plm::vcovHC(m)))[1]),
    sargan = unname(s$statistic), df = unname(s$parameter),
    p = unname(s$p.value),
    m2 = unname(plm::mtest(m, order = 2L, vcov = plm::vcovHC)$statistic))
}))
round(count_tab, 3)
```

```{r}
#| label: c06-4-picture
#| fig-cap: "Left: the two-step estimate of γ₁ with its Windmeijer 95% band
#|   against the instrument count, for the five instrument sets above;
#|   dashed, the two-way within estimate; dotted, pooled OLS. Right: the
#|   two standard errors of γ₁ against the instrument count, conventional
#|   two-step (filled) and Windmeijer-corrected (open)."
#| fig-alt: "Two panels. Left: five points with vertical whiskers. At 23
#|   and 28 instruments the point is near 0.02 with whiskers from about
#|   minus 0.5 to 0.5; at 35 it is 0.36 and at 38 it is 0.47 with whiskers
#|   of about 0.4 each side; the collapsed set at 18 instruments is at 0.85
#|   with a whisker from minus 0.25 to 1.96. A dashed line at 0.70 and a
#|   dotted line at 1.14 cross the panel. Right: filled points fall from
#|   0.16 at 23 instruments to 0.085 at 38, with 0.26 at the collapsed 18;
#|   open points stay between 0.19 and 0.28, with 0.56 at the collapsed 18."
x <- count_tab[, "instruments"]
g <- count_tab[, "gamma1"]
op <- graphics::par(mfrow = c(1, 2))
plot(x, g, pch = 16, ylim = c(-0.6, 2.1), xlim = c(15, 42),
     xlab = "instrument count", ylab = "gamma_1, two-step")
graphics::segments(x, g - 1.96 * count_tab[, "se_windmeijer"],
                   x, g + 1.96 * count_tab[, "se_windmeijer"], lwd = 2)
graphics::abline(h = coef(ab_within)[1], lty = 2)
graphics::abline(h = coef(ab_ols)[2], lty = 3)
graphics::text(x, g, rownames(count_tab), pos = 4, cex = 0.8)
plot(x, count_tab[, "se_twostep"], pch = 16, ylim = c(0, 0.6),
     xlim = c(15, 42), xlab = "instrument count",
     ylab = "standard error of gamma_1")
graphics::points(x, count_tab[, "se_windmeijer"], pch = 1)
graphics::legend("topright", c("two-step", "Windmeijer"), pch = c(16, 1),
                 bty = "n")
graphics::par(op)
```

**Read it.**

- **Sargan does not reject.** 30.1 on 25 degrees of freedom, p = 0.22. *So:*
  the 25 spare moment conditions are consistent with the 13 the estimate
  used; the instruments pass the test they can be given.
- **m1 rejects, m2 does not.** m1 = −2.43 (p = 0.015), m2 = −0.33
  (p = 0.74). *So:* the pattern a valid differenced model shows; the levels
  shocks look serially uncorrelated and $y_{i,t-2}$ stays a valid
  instrument. With the corrected covariance m1 is −1.54 (p = 0.12) and m2
  −0.28 (p = 0.78); on the one-step fit the robust pair is −2.49 and −0.36.
  The m2 verdict is the same in every version.
- **Windmeijer more than doubles the standard error.** The standard error of
  γ₁ goes from 0.085 to 0.185, and two coefficients lose their stars: γ₂
  (p from 0.05 to 0.31) and the lagged wage (0.005 to 0.11). *So:* the
  conventional two-step standard errors are not the ones to report.
- **The instrument count moves γ₁ across the whole bracket.** 0.02 with 23
  instruments, 0.36 with 35, 0.47 with 38, 0.85 with the 18 collapsed ones,
  and Sargan accepts the first four alike (p from 0.20 to 0.42); the only
  set it questions is the collapsed one (p = 0.04), which also has the
  widest band. As instruments are added the two-step standard error halves,
  0.159 to 0.085, while the corrected one falls by a third, 0.275 to 0.185.
  *So:* the Ziliak and Bowsher points at once; a Sargan p-value cannot pick
  the instrument set, so the set is a choice to report.
- **Misleads (a p-value near one).** With many instruments and few units a
  Sargan p-value near 1 is the Bowsher pathology, not a clean bill:
  @croissant2019 [p. 181] show p = 0.919 with every lag on 25 countries
  over seven periods, against 0.071 for their three-lag fit `diff25lim`,
  which their code runs on `DemocracyIncome` [@croissant2019, p. 173].
  *So:* report the count beside the p-value, and rerun with fewer lags
  before trusting either.

::: {.callout-note appearance="simple"}
**Next.** Report the two-step estimate with Windmeijer standard errors, the
Sargan statistic with its degrees of freedom, m1 and m2, and the instrument
count, and show how γ moves across instrument sets
([summary](index.qmd#sec-dyn-summary)).
:::

::: {.callout-note collapse="true"}
## Recipe

```r
# m: a two-step pgmm fit.
plm::sargan(m)
plm::mtest(m, order = 1L, vcov = plm::vcovHC)
plm::mtest(m, order = 2L, vcov = plm::vcovHC)
summary(m, robust = TRUE)            # Windmeijer standard errors
ncol(m$W[[1]])                       # instrument count
update(m, collapse = TRUE)           # the same with collapsed instruments
```
:::

::: {.callout-caution collapse="true"}
## Pitfalls

**`mtest` on a two-step fit uses one-step residuals.** The help page says
so, and it is what DPD, Arellano and Bond's own program, and Stata's
`xtabond` do. On the `?pgmm` Table 4(a2) equation this gives m2 = −0.416
where the help page says the paper prints −0.434; with
`vcov = plm::vcovHC` it is −0.352:

```{r}
#| label: c06-4-a2
plm::mtest(ab_a2, order = 2L)
plm::mtest(ab_a2, order = 2L, vcov = plm::vcovHC)
```

**`sargan(weights = "onestep")` is a different statistic.** On the same
two-step fit it gives 79.5 instead of 30.1 and is not the one to report:

```{r}
#| label: c06-4-sargan-onestep
plm::sargan(ab_two, weights = "onestep")
```

**The summary's AR lines follow its `robust` argument.** `summary(ab_two,
robust = FALSE)` prints m1 = −2.43 and `robust = TRUE` prints −1.54; the
Sargan line does not change.

**Wald tests on two-step fits over-reject.** @bond2001 suggest
criterion-based tests, the difference of the Sargan statistics of the
restricted and unrestricted fits, in place of Wald tests [@baltagi2021,
p. 193]. Not implemented here.

**Sargan's power can be bad.** On his own cigarette example at N = 46 and
T = 28, @baltagi2021 [p. 209] reads a non-rejection as possibly "due to the
bad power of this test" and collapses the instruments.
:::

::: {.callout-tip collapse="true"}
## Book check

`?mtest` annotates `ab.b.onestep` with `vcov = vcovHC` as "Windmeijer
(2005), table 2, onestep with corrected std. err" [@windmeijer2005];
`c06-4-mtest-onestep` above is that pair and prints m1 = −2.4934
(p = 0.013) and m2 = −0.3594 (p = 0.719). `?pgmm` annotates `ab.b` also as
"Windmeijer (2005), table 2, std. err" [@windmeijer2005] and
`summary(ab.b, robust = TRUE)` as "Windmeijer"; `c06-3-ab-twostep` with
`robust = TRUE` prints those standard errors, 0.185398 on γ₁ and the rest
of the row, and `c06-4-windmeijer` above repeats the row to four digits.
The paper is not in `lit/`, so the match is to the help-page annotation.
Ziliak's 0.519 → 0.093 [@ziliak1997; @baltagi2021, p. 194] and Bowsher's
variance of 13.7 against 180 [@bowsher2002; @baltagi2021, p. 192] are
quoted; the count table above reproduces the
structure of both, γ₁ from 0.02 to 0.85 across sets with Sargan p from 0.20
to 0.42, not their numbers.

On `DemocracyIncome`, @croissant2019 print standard errors 0.04795 and
0.04646 from `vcov` against 0.10783 and 0.06054 from `vcovHC` in Example
7.8 [@croissant2019, p. 179]; Sargan "chisq = 50, df = 44, p-value = 0.3"
for the difference fit and "56, 54, 0.4" for the system fit in Example 7.9
[@croissant2019, p. 180]; m2 "normal = 0.88, p-value = 0.4" in Example 7.10
[@croissant2019, p. 182]; and on `DemocracyIncome25` a Sargan p-value of
0.91890 with every lag [@croissant2019, p. 181]. Here: 49.881 (44)
p = 0.251, 55.678 (54) p = 0.411, 0.88094 p = 0.378 and 0.9189, all within
the rounding CM print.

```{r}
#| label: c06-4-cm
sqrt(diag(vcov(cm_diff2)))[1:2]
sqrt(diag(plm::vcovHC(cm_diff2)))[1:2]
plm::sargan(cm_diff2)
plm::sargan(cm_sys2)
plm::mtest(cm_diff2, order = 2)
data("DemocracyIncome25", package = "pder")
d25 <- plm::pgmm(democracy ~ lag(democracy) + lag(income) |
                   lag(democracy, 2:99) + lag(income, 2:99),
                 data = DemocracyIncome25, model = "twosteps")
plm::sargan(d25)$p.value
```
:::
