---
title: "Dynamic panels"
---

A firm's payroll this year is mostly last year's payroll. Hiring takes months
of search and training, firing takes notice periods and severance, so when
wages rise or demand falls, employment moves part of the way this year and the
rest over the next few. The question an economist asks is "how fast?", and
the answer is one number: the coefficient on last year's employment. At 0.5,
half the gap to the new level closes each year; at 0.9, the firm is still
adjusting a decade later. Every static estimator of the earlier chapters gets
that number wrong, and this chapter is about how wrong and what to do instead.

A **dynamic panel** is a panel regression with the lagged outcome among the
regressors. The lag carries the firm effect inside it, since last year's
employment was built by the same firm with the same effect, so the lag is
correlated with the error. The within transformation of the
[core chapter](../05-core/index.qmd#sec-core-fe-re), which subtracts each
firm's sample mean, removes the firm effect but puts the firm's whole error
history into every demeaned lag. The bias that results shrinks as the number
of years T grows and does not shrink as the number of firms N grows
[@nickell1981; @baltagi2021, p. 188].

**The equation.** With $\gamma$ the persistence, $x_{it}$ the other
regressors, $\alpha_i$ the firm effect and $\varepsilon_{it}$ a shock that is
not serially correlated,

$$
y_{it} = \gamma\, y_{i,t-1} + x_{it}'\beta + \alpha_i + \varepsilon_{it},
\qquad i = 1,\dots,N,\; t = 2,\dots,T .
$$

Its pure autoregressive form, without $x_{it}$, is the model @pesaran2015
[p. 678, eq. 27.5] and @baltagi2021 [p. 189, eq. 8.3] use to show the
bias and the estimators. Taking first differences removes $\alpha_i$:

$$
\Delta y_{it} = \gamma\, \Delta y_{i,t-1} + \Delta x_{it}'\beta
  + \Delta\varepsilon_{it} .
$$

But $\Delta y_{i,t-1} = y_{i,t-1} - y_{i,t-2}$ contains
$\varepsilon_{i,t-1}$, and so does
$\Delta\varepsilon_{it} = \varepsilon_{it} - \varepsilon_{i,t-1}$: the
differenced lag and the differenced error share one shock, so least squares
on the differenced equation is inconsistent as well [@pesaran2015, p. 682].
The whole chapter is about what to put in place of $\Delta y_{i,t-1}$.

**Terms from earlier chapters.** Fixed effects and the within transformation
are the subject of the [core chapter](../05-core/index.qmd#sec-core-fe-re).
A series with a **unit root** does not return to any level
([why it matters](../04-time/index.qmd#sec-ur-why)); when employment is that
persistent, every estimator here is in trouble, and
[step 3](index.qmd#sec-dyn-gmm) shows how. Serial correlation of
$\varepsilon_{it}$ is what the diagnostics of
[step 4](index.qmd#sec-dyn-diagnostics) test for, and the
[serial-correlation tests](../05-core/index.qmd#sec-core-serial) of the core
chapter are the static version of the same question.

**The running example.** `EmplUK` in **plm** [@croissant2008]: 140 UK firms
observed for 7 to 9 years between 1976 and 1984, 1031 firm-years, with
employment, the wage, capital and industry output, all in logs here. It is
the data of @arellano1991, "the benchmark data set used in Stata and EViews"
[@baltagi2021, p. 193], and @croissant2019 [p. 183] list @blundell1998,
@windmeijer2005 and @roodman2009 among the papers that reuse it. Three
equations on it do three jobs:

1. The **Arellano–Bond equation**: log employment on its first two lags, the
   log wage and its lag, log capital, log output and its lag, with time
   dummies. It carries the book checks and the diagnostics.
2. The **one-lag equation**: log employment on one lag, the log wage and log
   capital each with one lag, with time dummies. It carries the picture of
   OLS above, within below, and the consistent estimators between.
3. The **pure AR(1)** of log employment with time dummies: the case where
   every estimator misleads, because the series is close to a unit root.

**In this chapter.**

1. The bias: how far the within and OLS estimates sit from a known truth,
   and why only T shrinks the gap.
2. Difference and instrument: one instrument per equation, the
   Anderson–Hsiao estimator, consistent and imprecise.
3. All the instruments: difference GMM, the weak-instrument trap on a
   persistent series, and system GMM as the answer.
4. The diagnostics that decide which fit to believe: Sargan, the two
   autocorrelation tests, the corrected standard errors, the instrument
   count.
5. Summary: the recipe and one table of every estimate.

```{r}
#| label: c06-0-setup
# Shared simulators live in R/; see R/simulate-panels.R.
devtools::load_all(quiet = TRUE)
# pgmm() builds a plm() call and evaluates it in the caller's environment,
# so plm must be attached; the plm:: prefix stays for the reader.
library(plm)
data("EmplUK", package = "plm")
pe <- plm::pdata.frame(EmplUK, index = c("firm", "year"))
```

The bias, drawn. The curve is the large-N limit of the within estimate of
$\gamma = 0.5$ as a function of T, from the exact formula of
[step 1](index.qmd#sec-dyn-bias); the circles are one simulated panel of
100 firms at each T:

```{r}
#| label: c06-0-picture
#| fig-cap: "The within estimate of γ against the number of periods T, for
#|   γ = 0.5. Solid: the exact large-N bias curve (`nickell_bias()`).
#|   Circles: one simulated draw of 100 firms at T = 5, 10, 20, 30 and 50,
#|   within (black) and pooled OLS (grey). Dashed: the true value."
#| fig-alt: "A line chart with T from 3 to 50 on the horizontal axis and the
#|   estimate on the vertical axis. A solid curve rises steeply from just
#|   below zero at T = 3 to 0.47 at T = 50, approaching a dashed horizontal
#|   line at 0.5 from below. Black circles sit near the curve. Grey circles
#|   sit flat between 0.85 and 0.89, above the dashed line, at every T."
t_seq <- 3:50
set.seed(1)
one_draw <- sapply(c(5, 10, 20, 30, 50), function(t) {
  d <- sim_dynamic_panel(100, t)
  c(within = unname(coef(plm::plm(y ~ lag(y, 1), data = d,
                                  model = "within"))),
    ols = unname(coef(plm::plm(y ~ lag(y, 1), data = d,
                               model = "pooling"))[2]))
})
plot(t_seq, 0.5 + nickell_bias(0.5, t_seq), type = "l", lwd = 2,
     ylim = c(0, 1), xlab = "periods in the regression, T",
     ylab = "estimate of gamma")
graphics::abline(h = 0.5, lty = 2)
graphics::points(c(5, 10, 20, 30, 50), one_draw["within", ], pch = 21,
                 bg = "black", cex = 1.3)
graphics::points(c(5, 10, 20, 30, 50), one_draw["ols", ], pch = 21,
                 bg = "grey70", cex = 1.3)
graphics::legend("bottomright", c("within, one draw", "pooled OLS, one draw",
                            "exact within bias", "true gamma"),
                 pch = c(21, 21, NA, NA), pt.bg = c("black", "grey70", NA, NA),
                 lty = c(NA, NA, 1, 2), lwd = c(NA, NA, 2, 1), bty = "n")
```

{{< include 1-problem.qmd >}}

{{< include 2-anderson-hsiao.qmd >}}

{{< include 3-gmm.qmd >}}

{{< include 4-diagnostics.qmd >}}

{{< include 5-summary.qmd >}}
