## Why it matters {#sec-ur-why}

**Question.** If two series both wander, will a regression find a relationship
that does not exist?

**Intuition.** A random walk has no level it returns to. Any two of them drift,
and over a short window two drifts look like a trend in common. A regression
reads that coincidence as a slope and, worse, as a *precise* slope. This is the
spurious regression of @granger1974. Stationary series cannot do this: each one
keeps returning to its mean, so a chance co-movement does not last.

```{r}
#| label: s1-1-picture
#| fig-cap: "Five random walks (left) and five stationary AR(1) series (right)."
#| fig-alt: "Two panels of five lines each over 60 periods. On the left the
#|   random walks drift far apart and none returns to zero. On the right the
#|   stationary series stay in a narrow band around zero."
set.seed(2)
op <- graphics::par(mfrow = c(1, 2))
graphics::matplot(sim_ar_panel(5, 60, k = 0), type = "l", lty = 1,
                  main = "Random walks", xlab = "period", ylab = "")
graphics::matplot(sim_ar_panel(5, 60, k = 5), type = "l", lty = 1,
                  main = "Stationary, root 0.5", xlab = "period", ylab = "")
graphics::par(op)
```

**The experiment.** No test yet, only the damage. Draw 49 independent random
walks for `y` and 49 more for `x`, 30 periods each, so the true slope is zero.
Regress `y` on `x` with unit fixed effects and record whether the slope is
"significant" at 5%. Repeat 200 times. A sound procedure says "significant" 5%
of the time.

```{r}
#| label: s1-1-spurious
spurious <- function(model) {
  d <- as_pdata(sim_rw_panel(49, 30), sim_rw_panel(49, 30))
  fit <- plm::plm(y ~ x, data = d, model = model)
  summary(fit)$coefficients["x", "Pr(>|t|)"] < 0.05
}
set.seed(1)
c(levels = mean(replicate(200, spurious("within"))),
  differences = mean(replicate(200, spurious("fd"))))
```

**Read it.**

- **In levels, it misleads.** The fixed-effects regression finds a
  relationship that is not there in 57% of panels. Having 49 units does not
  help.
- **In first differences, it does not.** Regressing changes on changes gives
  3.5%, close to the nominal 5%. Differencing removes the wandering, and with
  it the spurious fit.

**What the panel does buy.** @phillips1999 show that the pooled slope itself
is not the problem. With many units it settles on a meaningful long-run
average, here zero. The problem is the default standard error, which assumes
the errors forget their past. Errors from two random walks never do. Standard
errors clustered by unit allow any pattern over time within a unit, and they
bring the rejection rate back near 5%:

```{r}
#| label: s1-1-clustered
spurious_clustered <- function() {
  d <- as_pdata(sim_rw_panel(49, 30), sim_rw_panel(49, 30))
  fit <- plm::plm(y ~ x, data = d, model = "within")
  v <- plm::vcovHC(fit, cluster = "group")
  summary(fit, vcov = v)$coefficients["x", "Pr(>|t|)"] < 0.05
}
set.seed(1)
mean(replicate(200, spurious_clustered()))
```

That fix only works if you know the series wander. So test first.

::: {.callout-note appearance="simple"}
**Next.** Find out whether your series wander:
[what a rejection means](unit-roots.qmd#sec-ur-alternative).
:::
