## Tests that assume independent units {#sec-ur-first-gen}

**Question.** Do log house prices carry a unit root in every state, if the
states are independent of each other?

**Intuition.** Run a **Dickey–Fuller regression** on each state: this year's
change in log price on last year's level, plus a few past changes. If the
level does not predict the change, the series has no level to return to, and
the *t* statistic on the level is small. Then combine the 49 answers into
one. The tests differ only in what they combine:

| Test | Combines | Alternative | `test =` |
|---|---|---|---|
| Levin–Lin–Chu [@levin2002] | one slope from all states stacked | all, alike | `"levinlin"` |
| Im–Pesaran–Shin [@im2003] | the average of 49 *t* statistics | a share | `"ips"` |
| Maddala–Wu [@maddala1999] | the 49 p-values | a share | `"madwu"` |
| Hadri [@hadri2000] | 49 stationarity statistics; **null reversed** | — | `"hadri"` |

Hadri is the odd one out. Its null is "every series is stationary". When the
first three fail to reject a unit root and Hadri rejects stationarity, two
opposite questions give one answer.

**The test.** Null: every series has a unit root (Hadri: every series is
stationary). Key assumption: states are independent once a common year effect
is removed. That assumption is what the next section drops.

**Run it.** Log prices in levels, with a trend and two lags in each state's
regression, as in @croissant2019 [§8.4.2]:

```{r}
#| label: c04-3-levels
plm::purtest(log(php$price), test = "levinlin", lags = 2, exo = "trend")
plm::purtest(log(php$price), test = "ips", lags = 2, exo = "trend")
plm::purtest(log(php$price), test = "madwu", lags = 2, exo = "trend")
plm::purtest(log(php$price), test = "hadri", lags = 2, exo = "trend")
```

**Read it: does not reject (levels).** LLC, IPS and Maddala–Wu give
p = 0.10, 0.78 and 0.43: no evidence against a unit root. Hadri rejects
stationarity at p ≈ 0. All four agree: price *levels* wander.

Now the growth rates, the first difference of log prices. A growth rate should
not wander, so the tests should reject:

```{r}
#| label: c04-3-growth
dlp <- diff(log(php$price))
plm::purtest(dlp, test = "levinlin", lags = 1, exo = "intercept")
plm::purtest(dlp, test = "ips", lags = 1, exo = "intercept")
plm::purtest(dlp, test = "madwu", lags = 1, exo = "intercept")
plm::purtest(dlp, test = "hadri", lags = 1, exo = "intercept")
```

The 49 state statistics that IPS averages, in levels and in growth rates:

```{r}
#| label: c04-3-picture
#| fig-cap: "The state-by-state Dickey–Fuller t statistics that IPS
#|   averages: log price levels with a trend and two lags (left), growth
#|   rates with an intercept and one lag (right). Thick bar: the mean of the
#|   49, IPS's raw input. Dashed: the asymptotic 5% critical value of a
#|   single Dickey–Fuller test with that deterministic part."
#| fig-alt: "Two vertical clouds of 49 points. The levels cloud is centred
#|   near minus two, with most points above the dashed line at minus 3.4.
#|   The growth cloud is centred near minus three, with most points below
#|   the dashed line at minus 2.9."
lev <- plm::purtest(log(php$price), test = "ips", lags = 2, exo = "trend")
gro <- plm::purtest(dlp, test = "ips", lags = 1, exo = "intercept")
t_stat <- function(pu) sapply(pu$idres, \(r) unname(r$trho))
ts <- data.frame(levels = t_stat(lev), growth = t_stat(gro))
graphics::stripchart(ts, vertical = TRUE, method = "jitter", pch = 16,
                     col = "grey50", ylab = "Dickey–Fuller t, by state")
graphics::segments(c(0.8, 1.8), colMeans(ts), c(1.2, 2.2), colMeans(ts),
                   lwd = 4)
graphics::segments(c(0.7, 1.7), c(-3.41, -2.86), c(1.3, 2.3),
                   c(-3.41, -2.86), lty = 2)
```

**Read it: rejects (growth rates).** LLC, IPS and Maddala–Wu all reject at
p < 2.2 × 10⁻¹⁶. In the picture the whole cloud drops: the mean *t* goes
from −1.98 in levels to −2.98 in growth rates, and 30 of the 49 states would
reject on their own. Growth rates do not wander.

**Read it: misleads (Hadri on growth rates).** Hadri *also* rejects, z = 12.3:
it says growth rates are not stationary either. The two families now
contradict each other, and the reason is the assumption they share. The states
are far from independent: the CD test gives z = 53 on levels and z = 71 on
growth rates.

```{r}
#| label: c04-3-cd
plm::pcdtest(dlp, test = "cd")
```

When 49 states move together, they are not 49 independent pieces of evidence.
Every test in this section counts them as if they were. Treat its numbers as
the ones to beat, not the ones to report.

::: {.callout-note appearance="simple"}
**Next.** CD rejects here, so go to
[tests that allow a common factor](index.qmd#sec-ur-second-gen). On a
panel where CD does not reject, report LLC or IPS with its alternative, and go
to [which series are stationary](index.qmd#sec-ur-fraction).
:::

::: {.callout-note collapse="true"}
## Recipe

```r
# y: a pseries, e.g. plm::pdata.frame(df, index = c("id", "year"))$y
plm::pcdtest(y, test = "cd")  # first: are units independent?
plm::purtest(y, test = "ips", exo = "intercept", lags = "AIC", pmax = 4)
plm::purtest(y, test = "hadri", exo = "intercept", lags = "AIC", pmax = 4)
```
:::

::: {.callout-caution collapse="true"}
## Pitfalls

**A formula silently ignores `exo`.** `purtest` takes a matrix, a `pseries`
or a formula. With a formula, the right-hand side sets the deterministic part
and `exo` is dropped without a warning:

```{r}
#| label: c04-3-exo-trap
data("Grunfeld", package = "plm")
pG <- plm::pdata.frame(Grunfeld, index = c("firm", "year"))
plm::purtest(inv ~ 1, data = Grunfeld, index = "firm", test = "madwu",
             pmax = 4, exo = "trend")
plm::purtest(inv ~ trend, data = Grunfeld, index = "firm", test = "madwu",
             pmax = 4)
plm::purtest(pG$inv, test = "madwu", pmax = 4, exo = "trend")
```

The first call asks for a trend and gets intercepts only: 14.719 against
78.977. Pick one interface and keep to it.

**LLC and Hadri need a balanced panel.** IPS and Maddala–Wu do not. On the
unbalanced `EmplUK` [@arellano1991], two of four refuse:

```{r}
#| label: c04-3-unbalanced
#| error: true
#| results: markup
data("EmplUK", package = "plm")
pE <- plm::pdata.frame(EmplUK, index = c("firm", "year"))
# `error: true` prints each refusal and carries on; `results: markup` keeps
# each answer under the call that produced it.
plm::purtest(log(pE$emp), test = "levinlin", exo = "intercept", lags = 0)
plm::purtest(log(pE$emp), test = "ips", exo = "intercept", lags = 0)
plm::purtest(log(pE$emp), test = "madwu", exo = "intercept", lags = 0)
plm::purtest(log(pE$emp), test = "hadri", exo = "intercept", lags = 0)
```

**Install urca.** Maddala–Wu combines per-state p-values. **plm** reads them
from @mackinnon1996 through **urca**; without it, it falls back to
@mackinnon1994. On house prices that moves p from 0.4292 to 0.4313:

```{r}
#| label: c04-3-padf
plm::purtest(log(php$price), test = "madwu", lags = 2, exo = "trend",
             p.approx = "MacKinnon1994")
```

**IPS `tbar`.** `ips.stat = "tbar"` is exact for fixed T, but only without
lags, and it returns tabulated critical values instead of a p-value.

**Missing tests.** Breitung [@breitung2000] and Harris–Tzavalis
[@harris1999], which have the best size record [@pesaran2015, p. 838], are not
in **plm**. The benchmark is Stata's `xtunitroot`.
:::

::: {.callout-tip collapse="true"}
## Book check

@croissant2019 [§8.4.2] runs LLC, Maddala–Wu and IPS on log house prices with
`lags = 2, exo = "trend"`; the levels run above is that run. The second check
is the `?purtest` example: Maddala–Wu on Grunfeld investment must give the same
statistic through all three interfaces.

```{r}
#| label: c04-3-book-check
wide <- data.frame(split(Grunfeld$inv, Grunfeld$firm))
plm::purtest(wide, test = "madwu", pmax = 4, exo = "intercept")
plm::purtest(inv ~ 1, data = Grunfeld, index = "firm", test = "madwu",
             pmax = 4)
plm::purtest(pG$inv, test = "madwu", pmax = 4, exo = "intercept")
```

14.719 from all three.
:::
