## Tests that allow a common factor {#sec-ur-second-gen}

**Question.** Do house prices still look like random walks once we allow all
states to be hit by one common shock, each state to its own degree?

**Intuition.** A national credit boom lifts every state's prices, some more
than others. That shared shock makes 49 states look like fewer independent
pieces of evidence. The fix is to give each state's Dickey–Fuller regression a
proxy for the shock: the cross-state average of prices, last year and this
year's change. Each state gets its own coefficient on the average, so each
state can react to the shock by its own amount. Then average the 49 *t*
statistics, as IPS does. That average is **CIPS** [@pesaran2007].

**The test.** Null: every series has a unit root. Alternative: a share is
stationary. Assumption: one common factor, and no other link between states.
CIPS has no exact p-value; **plm** returns the nearest tabulated level, so
`0.1` means *above* 0.10 and `0.01` means *below* 0.01.

**Run it.** Levels get a drift; growth rates get neither drift nor trend:

```{r}
#| label: c04-4-levels
plm::cipstest(log(php$price), type = "drift")
```

```{r}
#| label: c04-4-growth
plm::cipstest(diff(log(php$price)), type = "none")
```

What the averages do to each state's evidence. `cipstest` keeps the 49
augmented regressions in `$tmods`; the *t* statistic on the lagged level,
`le`, is each state's CADF statistic, and their mean is CIPS:

```{r}
#| label: c04-4-picture
#| fig-cap: "Each state's unit-root t statistic on log price levels without
#|   the cross-state averages (horizontal) and with them (vertical), both
#|   with an intercept and two lags. Dashed: the 45-degree line. Cross: the
#|   two means; the vertical one is the CIPS statistic."
#| fig-alt: "A scatter of 49 points around a dashed diagonal line, most of
#|   them below it, so the vertical statistic is more negative than the
#|   horizontal one. A cross marks the means at about minus 1.6 across and
#|   minus 2.0 up."
cips <- plm::cipstest(log(php$price), type = "drift")
adf <- plm::purtest(log(php$price), test = "ips", lags = 2, exo = "intercept")
t_adf <- sapply(adf$idres, \(r) unname(r$trho))
t_cadf <- sapply(cips$tmods, \(m) summary(m)$coefficients["le", "t value"])
graphics::plot(t_adf, t_cadf, pch = 16, col = "grey40",
               xlab = "Dickey–Fuller t, no averages",
               ylab = "CADF t, with cross-state averages")
graphics::abline(0, 1, lty = 2)
graphics::points(mean(t_adf), mean(t_cadf), pch = 3, cex = 2.5, lwd = 2)
c(mean_adf = mean(t_adf), mean_cadf = mean(t_cadf), cips = cips$statistic)
```

**Read it.**

- **Does not reject (levels).** CIPS = −2.03, p above 0.10. After the common
  shock is allowed for, price levels still wander. This now carries weight,
  because the test no longer assumes the states are independent. The picture
  shows the averages at work: they move almost every state's statistic, most
  of them down, and the mean goes from −1.62 to −2.03, with no state near
  the −3 or so a single test would need.
- **Rejects (growth rates).** CIPS = −1.82, p below 0.01. Growth rates do not
  wander. There is no contradiction left: the Hadri result of the
  [previous section](index.qmd#sec-ur-first-gen) was the common shock
  speaking.

**Rejects, and it means something: the residuals of a regression.** Prices
wander, and so do incomes. If a regression of one on the other leaves
residuals that do *not* wander, the two share one trend: they are
*cointegrated*, and the levels regression is meaningful rather than spurious
([why it matters](index.qmd#sec-ur-why)). Take the two CCE regressions of the
[CCE section](../02-cross-dependence/index.qmd#sec-cce) and test their
residuals:

```{r}
#| label: c04-4-residuals
# pcce() builds a call to plm(), so plm must be attached, not just namespaced.
library(plm)
ccemg <- plm::pcce(log(price) ~ log(income), data = HousePricesUS,
                   model = "mg")
ccep <- plm::pcce(log(price) ~ log(income), data = HousePricesUS,
                  model = "p")
plm::cipstest(residuals(ccemg), type = "none")
plm::cipstest(residuals(ccep), type = "none")
```

Both reject. Log price and log income move together in the long run, and the
CCE estimates of the income elasticity, 1.14 and 1.20, describe that long-run
link.

::: {.callout-note appearance="simple"}
**Next.** Levels wander, residuals do not: estimate in levels with CCE
([CCE](../02-cross-dependence/index.qmd#sec-cce)). To find out which states
drive a rejection: [which series are stationary](index.qmd#sec-ur-fraction).
Before you trust any of it: [when the tests lie](index.qmd#sec-ur-traps).
:::

::: {.callout-note collapse="true"}
## Recipe

```r
# y: a pseries. Levels of a trending variable: "drift" or "trend".
# Differences and regression residuals: "none".
plm::cipstest(y, type = "drift")
plm::cipstest(diff(y), type = "none")
plm::cipstest(residuals(fit), type = "none")  # fit: a pcce() model
```
:::

::: {.callout-caution collapse="true"}
## Pitfalls

- **`pseries` only.** `cipstest` takes no formula and no matrix. Wrap a matrix
  with `as_pseries()` (in `R/panel-reshape.R`).
- **`type` is your choice, not the data's.** Levels of a price get a drift;
  differences and residuals get neither.
- **Never quote the p-value as exact.** The warnings that say "p-value
  greater/smaller than printed" are hidden in this book's chunk options, so the
  qualification has to travel in your text.
- **One factor only.** With several common shocks the averages are not enough.
  CIPS with more factors and CSB [@pesaran2013], Moon–Perron [@moon2004],
  PANIC [@bai2004] and Phillips–Sul [@phillips2003] have no R implementation
  here.
:::

::: {.callout-tip collapse="true"}
## Book check

@croissant2019 [§8.4.3] runs exactly these four tests. Targets: CIPS about −2
with p ≈ 0.1 in levels, about −1.8 with p ≈ 0.01 on differences, and
rejections for both CCE residuals. Results above: −2.0342, −1.8199, and two
rejections. Four of four.
:::
