## When the tests lie {#sec-ur-traps}

Every result so far assumes the test rejects a true null 5% of the time. Two
things break that, and neither shows up in the output. Both checks here are
simulations, with no published number to match: they are evidence about this
code on these designs, and nothing more.

### Panel shape and spatial shocks {#sec-ur-spatial-size}

**Question.** At my N, T and spatial structure, how often does each test
reject a unit root that is really there?

**Intuition.** Two forces push a test to reject too often. First, **shape**:
LLC's theory needs T to grow faster than N [@levin2002], and many panels have
far more units than years. Second, **spatial shocks**: when a shock in one
state spills into its neighbours, neighbours stop being independent evidence
[@baltagi2007]. Factor-based fixes like CIPS target a shock common to all, not
a local spillover, and @pesaran2015 [p. 839] warns they can be invalid there.

**Run it.** 49 random walks on a 7 × 7 grid, where each cell neighbours the
cells beside it (`rook_w(7)`). Shocks spill over with strength λ: 0 is none,
0.8 is strong. Every series truly has a unit root, so a correct test rejects
5% of the time. First one draw at λ = 0.8, every test in the open:

```{r}
#| label: c04-6-one-draw
W <- rook_w(7)
set.seed(1)
m <- sim_spatial_rw_panel(W, lambda = 0.8)
plm::purtest(m, test = "levinlin", exo = "none", lags = 1)
plm::purtest(m, test = "levinlin", exo = "intercept", lags = 1)
plm::purtest(m, test = "ips", exo = "intercept", lags = 1)
plm::purtest(m, test = "madwu", exo = "intercept", lags = 1)
plm::cipstest(as_pseries(m), type = "drift", lags = 1)
```

Four say "do not reject"; Maddala–Wu rejects at p = 0.011. One draw proves
nothing, so repeat 200 times at three values of λ:

```{r}
#| label: c04-6-size
# One replication: the five calls above, kept for their p-values.
p_values <- function(m) {
  llc_none <- plm::purtest(m, test = "levinlin", exo = "none", lags = 1)
  llc_int <- plm::purtest(m, test = "levinlin", exo = "intercept", lags = 1)
  ips <- plm::purtest(m, test = "ips", exo = "intercept", lags = 1)
  mw <- plm::purtest(m, test = "madwu", exo = "intercept", lags = 1)
  cips <- plm::cipstest(as_pseries(m), type = "drift", lags = 1)
  c(LLC_none = unname(llc_none$statistic$p.value),
    LLC_intercept = unname(llc_int$statistic$p.value),
    IPS = unname(ips$statistic$p.value),
    MW = unname(mw$statistic$p.value),
    CIPS = unname(cips$p.value))
}
set.seed(1)
lambdas <- c(0, 0.4, 0.8)
size <- sapply(lambdas, function(l) {
  rowMeans(replicate(200, p_values(sim_spatial_rw_panel(W, l))) <= 0.05)
})
colnames(size) <- paste0("lambda_", lambdas)
size
```

**Read it.**

- **Holds its size: CIPS.** 0.04 at λ = 0, 0.07 at λ = 0.8. Close to 5%
  throughout.
- **Over-rejects already at λ = 0: the four first-generation tests.** With no
  spatial link at all, they reject a true null 2.5 to 12 times too often.
- **Spatial shocks add to it, but less.** LLC without intercept 0.13 → 0.31,
  IPS 0.18 → 0.28. With 200 draws, the noise in these rates is about ±0.015
  near 0.05 and ±0.035 near 0.5.

So at λ = 0 the damage comes from shape, not space. Fix λ at 0 and vary only
N and T, for LLC:

```{r}
#| label: c04-6-shape
llc_size <- function(n, t, reps = 200) {
  p <- replicate(reps, unname(plm::purtest(sim_rw_panel(n, t),
                                           test = "levinlin",
                                           exo = "intercept",
                                           lags = 1)$statistic$p.value))
  mean(p <= 0.05)
}
set.seed(1)
shape <- data.frame(N = c(10, 25, 49, 100), T = c(250, 100, 30, 30))
shape$N_over_T <- round(shape$N / shape$T, 2)
shape$size <- mapply(llc_size, shape$N, shape$T)
shape
```

At N/T = 0.04, LLC's design regime, size is 0.09. At N/T = 3.33 it is 0.855:
five false alarms in six. A panel of 170 regions over 20 years has N/T ≈ 8.5.

::: {.callout-note appearance="simple"}
**Next.** When N/T is well above 1, report CIPS and read every
first-generation p-value as an upper bound on the evidence. Rerun this table
with your own N, T and W before you trust any of them.
:::

### Shared trends across units {#sec-ur-cross-coint}

**Question.** Could two units share a long-run trend, and would any
dependence test notice?

**Intuition.** Units can move together in two ways. Their *shocks* can be
correlated: each still wanders on its own path. Or they can share one
*trend*: then some mix of the two does not wander at all. The second is
*cross-unit cointegration*, and @pesaran2015 [p. 836] warns that it can bias
panel unit-root tests severely. CIPS allows one common trend for all units
and rules out any other [@pesaran2015, p. 837].

**Run it.** Two simulated pairs of random walks. `sim_factor_pair()` loads two
series on one shared trend: cointegrated. `sim_correlated_pair()` gives each
series its own trend with shocks correlated at 0.8: not cointegrated.
Johansen's rank test [@pfaff2008] checks for a shared trend; the correlation
is what CD and α see:

```{r}
#| label: c04-6-johansen
johansen <- function(dgp, reps = 200) {
  out <- replicate(reps, {
    x <- dgp()
    z <- urca::ca.jo(x, type = "trace", ecdet = "none", K = 2)
    c(unname(z@teststat[2]) > z@cval[2, "5pct"], abs(cor(x)[1, 2]))
  })
  c(reject_r0 = mean(out[1, ]), mean_abs_cor = mean(out[2, ]))
}
set.seed(1)
round(rbind(shared_trend = johansen(sim_factor_pair),
            correlated_shocks = johansen(sim_correlated_pair)), 3)
```

**Read it.**

- **Rejects (shared trend).** Johansen finds the shared trend in every draw.
- **Does not reject (correlated shocks).** 5% of draws: its nominal size.
- **Misleads (the correlation).** 0.90 against 0.77. No correlation-based
  check, CD or α included, can tell these two pairs apart. The quantity that
  breaks the unit-root tests is invisible to the diagnostics that get reported.

::: {.callout-note appearance="simple"}
**Next.** No panel test for cross-unit cointegration exists in R, so carry it
as a stated limitation: "CIPS results assume no shared trends between units
beyond one common factor."
:::

::: {.callout-caution collapse="true"}
## Pitfalls

- **Pairwise Johansen is not a fix.** At N = 49 it is 1176 tests, at N = 170
  it is 14 365, and none of this book's sources gives a rule for combining
  them.
- **Subsampling needs N small.** The test of @choi2007 allows shared trends
  but needs N small relative to T [@pesaran2015, p. 838].
:::

::: {.callout-tip collapse="true"}
## Book check

None, for either subsection. `.docs/panel-spat-PLAN.md` records no canned
target: @baltagi2007's size tables need their exact design, and no source
prints a number for a cross-unit cointegration test that has no
implementation. `urca::ca.jo` reproduces Johansen's money-demand examples on
its own help page; that is the credit it is used on here.
:::
