## Cointegration {#sec-ur-coint}

**Question.** Prices wander and incomes wander. Does a state's price level
stay tied to its income level, so that the gap between them does not wander,
and does the panel as a whole say so?

**Intuition.** Incomes in a state rise 10%, prices rise 10%, and then both
drift on together for a decade: the ratio holds, whatever either level does.
Two I(1) series are **cointegrated** when some combination of them is I(0)
[@engle1987; @croissant2019, p. 204]. That combination is the long-run
relation, and its residual is the gap that closes. There are three ways to
look for it, from one state to the panel. (1) Regress one series on the other
and test the residual for a unit root, the Engle–Granger route [@engle1987;
@croissant2019, p. 204]. (2) Ask how many stationary combinations the pair
has, its **rank**, with Johansen's trace test [@johansen1988; @johansen1991]:
rank 0 is no shared trend, rank 1 is one long-run relation, rank 2 means both
series are stationary on their own. (3) Pool the per-state evidence: Pedroni's
tests average per-state residual statistics [@pedroni1999; @pedroni2004;
@baltagi2021, pp. 360–361], as IPS averages per-state Dickey–Fuller
statistics and CIPS averages the same statistics with the cross-state means
added ([tests that allow a common factor](index.qmd#sec-ur-second-gen)). The
residual-based panel tests are built for rank 1 in every unit, with no
cointegration among the regressors [@pesaran2015, p. 839], so the rank
question comes first.

::: {.callout-tip appearance="simple"}
Why the rank matters, what Pedroni's statistics average, and what the panel
adds: [cointegration in panels](theory-cointegration.qmd).
:::

**The tests.** Johansen's trace test, one state at a time [@johansen1991;
@pfaff2008]. Null: rank 0, no shared trend, against the full rank 2; if that
rejects, null rank 1 against rank 2 [@johansen1991; @pesaran2015, p. 541].
The `ca.jo` call of [shared trends across
units](index.qmd#sec-ur-cross-coint) on simulated pairs is the same test;
here it runs on real states. Pedroni's seven statistics [@pedroni1999].
Null: no cointegration in any state [@pedroni1999; @baltagi2021, p. 360].
Alternative: the gap closes with one common speed (the four *panel*
statistics) or with a speed of its own in each state (the three *group*
statistics) [@pedroni1999; @baltagi2021, p. 372]. Panel *v* rejects in the
right tail, the other six in the left [@pedroni1999].

Four tests from the same literature have no R implementation and are stated
limitations. Kao's pooled Dickey–Fuller and ADF tests fix one slope for all
units [@kao1999; @baltagi2021, pp. 357–358]. The McCoskey–Kao LM test puts
cointegration in the null and "is not equipped to deal with cross-sectional
dependence" [@mccoskey1998; @baltagi2021, p. 359]. The LR-bar of Larsson,
Lyhagen and Löthgren averages the trace statistics below, as IPS averages
Dickey–Fuller statistics, but "requires a large time-series dimension" and
its size "will be severely distorted" otherwise [@larsson2001; @baltagi2021,
p. 361]. Westerlund's error-correction tests [@westerlund2007] are in no R
package either.

**Run it.** One state in the open first. `ecdet = "none"` leaves the
constant outside the long-run relation, so the levels may drift, as the
[second-generation step](index.qmd#sec-ur-second-gen) allowed them to. The
two rows are the null rank at most 1 and the null rank 0, with their critical
values:

```{r}
#| label: c04-7-one-state
ca <- subset(HousePricesUS, names == "California")
ca <- ca[order(ca$year), ]
z_ca <- urca::ca.jo(cbind(price = log(ca$price), income = log(ca$income)),
                    type = "trace", ecdet = "none", K = 2, spec = "longrun")
cbind(trace = z_ca@teststat, z_ca@cval)
tx <- subset(HousePricesUS, names == "Texas")
tx <- tx[order(tx$year), ]
z_tx <- urca::ca.jo(cbind(price = log(tx$price), income = log(tx$income)),
                    type = "trace", ecdet = "none", K = 2, spec = "longrun")
cbind(trace = z_tx@teststat, z_tx@cval)
```

California rejects rank 0 (24.43 against 17.95) and not rank 1 (1.52 against
8.18): one shared trend. Texas does not reject rank 0 (3.38). All 49 states
go through `johansen_by_unit()` (in `R/cointegration.R`), which makes the
same call once per state and reads the two rows at 5%; the last line is the
mean of the rank-0 statistic over the states:

```{r}
#| label: c04-7-by-state
jo <- johansen_by_unit(php, c("price", "income"), log = TRUE, ecdet = "none",
                       K = 2, label = "names")
table(rank = jo$rank)
jo[jo$rank >= 1, c("label", "trace_r0", "trace_r1", "rank", "slope")]
mean(jo$trace_r0)
```

The Engle–Granger route, state by state: regress log price on log income in
each state, then a Dickey–Fuller regression with two lags on the residual.
California in the open first, with no constant and no trend in the test
regression, since the residual has mean zero by construction:

```{r}
#| label: c04-7-one-state-eg
u_ca <- residuals(lm(log(price) ~ log(income), data = ca))
urca::ur.df(u_ca, type = "none", lags = 2)@teststat
```

The 49 state regressions are the mean-group fit of the [CCE
step](../02-cross-dependence/index.qmd#sec-cce), and `adf_by_unit()` makes
the same `ur.df` call once per state and returns the 49 *t* statistics. Run
it once on those residuals and once on the residuals of the CCEMG fit of the
[second-generation step](index.qmd#sec-ur-second-gen), where the cross-state
averages have taken the national factor out:

```{r}
#| label: c04-7-eg
mg <- plm::pmg(log(price) ~ log(income), data = HousePricesUS, model = "mg")
t_ols <- adf_by_unit(residuals(mg), lags = 2, type = "none")
t_cce <- adf_by_unit(residuals(ccemg), lags = 2, type = "none")
c(mean_ols = mean(t_ols), mean_cce = mean(t_cce))
c(below_ols = sum(t_ols < -3.37), below_cce = sum(t_cce < -3.37))
```

California's −4.42 is below the line −3.37, the 5% value Engle and Granger
tabulated for the Dickey–Fuller test on the residual of a two-variable
regression [@engle1987]; with two lags it is a reference line, not an exact
critical value. Pedroni's group statistics are built from these per-state
residual statistics, and CIPS on residuals averages them after augmentation.

Pedroni's test is in **pco** 1.0.1, archived on CRAN. Before its numbers on
house prices can be read (they are in the Pitfalls), the implementation has
to hold its size. 200 panels of 49 independent pairs over 29 years, random
walks with no shared trend, then 200 where `y = 1 + x + 0.5 e` in every
state; a 5% one-sided rule on the standardised values, each statistic
centred and scaled by Pedroni's tabulated moments so that it is N(0, 1)
under the null, right tail for panel *v*, left tail for the rest:

```{r}
#| label: c04-7-pco-size
# One replication: the seven standardised statistics, kept as 5% verdicts.
pedroni_rejects <- function(sim) {
  s <- pco::pedroni99(sim$Y, sim$X, type.stat = 2)$STATISTIC[, "standardized"]
  c(s[1] > qnorm(0.95), s[-1] < qnorm(0.05))
}
set.seed(1)
size <- rowMeans(replicate(200, pedroni_rejects(
  sim_coint_pair_panel(49, 29, coint = FALSE))))
power <- rowMeans(replicate(200, pedroni_rejects(
  sim_coint_pair_panel(49, 29, coint = TRUE))))
cbind(size, power)
```

```{r}
#| label: c04-7-picture
#| fig-height: 7
#| fig-cap: "Left: the Johansen trace statistic for rank 0 in each of the
#|   49 states, sorted, with the 10% line (dotted, 15.66) and the 5% line
#|   (dashed, 17.95); states above the 5% line in black. Right: each state's
#|   Dickey–Fuller t on its own OLS residual (horizontal) against the same t
#|   on its CCEMG residual (vertical). Dashed: the 45-degree line. Cross: the
#|   two means."
#| fig-alt: "Left: 49 dots climbing from about 3 to 29 along a vertical list
#|   of state abbreviations, with two vertical lines near 16 and 18; the top
#|   eight dots, to the right of the dashed line, are black and the rest
#|   grey. Right: a scatter of 49 points, most of them below a dashed
#|   diagonal, with a cross at about minus 2.2 across and minus 2.8 up."
op <- graphics::par(mfrow = c(1, 2), mar = c(4.5, 4, 1, 1))
o <- order(jo$trace_r0)
plate <- as.character(HousePricesUS$plate)[match(jo$label, HousePricesUS$names)]
above <- jo$trace_r0[o] > 17.95
graphics::plot(jo$trace_r0[o], seq_along(o), pch = 16,
               col = ifelse(above, "black", "grey55"), yaxt = "n",
               xlab = "Johansen trace statistic, rank 0", ylab = "",
               xlim = c(0, 32))
graphics::text(-0.6, seq_along(o), plate[o], cex = 0.55, adj = 1, xpd = NA)
graphics::abline(v = c(15.66, 17.95), lty = c(3, 2))
graphics::plot(t_ols, t_cce, pch = 16, col = "grey40",
               xlab = "Dickey–Fuller t, state OLS residuals",
               ylab = "Dickey–Fuller t, CCEMG residuals")
graphics::abline(0, 1, lty = 2)
graphics::points(mean(t_ols), mean(t_cce), pch = 3, cex = 2.5, lwd = 2)
graphics::par(op)
```

**Read it.**

- **Rejects (seven states, rank 1).** California, Maryland, Massachusetts,
  New Jersey, Pennsylvania, Rhode Island and Virginia, with trace statistics
  from 19.16 to 25.49 against 17.95. *So:* in these states the price–income
  gap closes. Their long-run slopes from the first eigenvector run from 0.14
  to 2.17, too wide at T = 29 to report as elasticities.
- **Does not reject (41 states).** Trace below 17.95; mean over the 49
  states 11.24. *So:* one state over 29 years cannot show a shared trend, and
  with 49 tests at 5% about 2.5 of the eight rejections are expected by
  chance. The precondition of rank 1 in every unit [@pesaran2015, p. 839]
  cannot be confirmed state by state.
- **Misleads (New Hampshire, rank 2).** 29.35, then 11.16 against 8.18,
  which says both series are stationary, against every unit-root result of
  this chapter. *So:* one rank-2 verdict in 49 is a false alarm, not a
  finding.
- **Cannot be read (Pedroni in pco).** Under no cointegration five of the
  seven statistics reject in 0 of 200 draws, panel PP in 4%, panel ADF in
  100%; under cointegration six reject in every draw and panel *v* in none.
  *So:* the standardised values separate the two hypotheses in direction and
  are not N(0, 1) at this shape; the house-price values (Pitfalls) are
  reported as unverified, not as a verdict.
- **With the averages (Engle–Granger on CCEMG residuals).** The mean *t*
  falls from −2.23 to −2.79, and ten states pass −3.37 against five. *So:*
  the shared trend appears once the common shock is partialled out, which is
  the CIPS result of the [second-generation
  step](index.qmd#sec-ur-second-gen) seen state by state.

::: {.callout-note appearance="simple"}
**Next.** Rank 1 in enough states, and residuals that do not wander once the
factor is removed: estimate the long-run slope
([estimating the long run](index.qmd#sec-ur-long-run)). Rank 0 everywhere:
difference, and model the short run. Rank 2 in a unit: go back to the
unit-root tests for that unit.
:::

::: {.callout-note collapse="true"}
## Recipe

```r
# y, x: one unit's two series, logged; d: a pdata.frame; Y, X: T x N
# matrices from panel_wide(d, log(y)); fit: a pmg() or pcce() model
z <- urca::ca.jo(cbind(y = y, x = x), type = "trace", ecdet = "none",
                 K = 2, spec = "longrun")
cbind(trace = z@teststat, z@cval)           # rows: rank <= 1, rank 0
johansen_by_unit(d, c("y", "x"), log = TRUE)  # every unit, rank at 5%
adf_by_unit(residuals(fit), lags = 2)         # Engle–Granger t per unit
pco::pedroni99(Y, X, type.stat = 2)$STATISTIC # five of seven usable,
#   unverified; rows nipanel, rhopanel, tpanelnonpar, tpanelpar, rhogroup,
#   tgroupnonpar, tgrouppar are panel v, rho, PP, ADF, group rho, PP, ADF
```
:::

::: {.callout-caution collapse="true"}
## Pitfalls

**Pedroni on house prices, unverified.** `type.stat = 2` puts an intercept in
each state's regression; the Newey–West lag, the number of autocovariances
that enter the long-run variance estimate of each residual, is the default
`round(4 * (29/100)^(2/9))` = 3 and the ADF lag 2:

```{r}
#| label: c04-7-pedroni
Y <- panel_wide(php, log(price))
X <- panel_wide(php, log(income))
ped <- pco::pedroni99(Y, X, type.stat = 2)
ped$STATISTIC
```

Rows `nipanel` to `tgrouppar` are panel *v*, panel rho, panel PP, panel ADF,
group rho, group PP and group ADF. Read against Pedroni's one-sided rule,
the five usable statistics (panel rho, panel PP, group rho, group PP, group
ADF) sit 4.1 to 11.9 standard deviations into the non-rejection tail: a test
that treats the 49 states as independent, with the national factor left in
every residual, finds no shared trend. Given the size simulation above, that
is evidence about this code, not about the panel. The two rows left out are
the next trap.

**Two `pco` statistics change with the units of the data.** Pedroni's
statistics are scale-free. Multiply both series by 100:

```{r}
#| label: c04-7-scale-trap
ped100 <- pco::pedroni99(100 * Y, 100 * X, type.stat = 2)
cbind(original = ped$STATISTIC[, "empirical"],
      times_100 = ped100$STATISTIC[, "empirical"])
```

Panel *v* goes from 0.185 to 1848.9 and panel ADF from −5334.2 to −0.53;
the other five do not move. Never report those two.

**The standardised values are not N(0, 1) at this shape.** The simulation
in the step: size 0 in 200 draws for five statistics and 1.00 for panel ADF;
power 1.00 for six. The moments `pco` applies are hard-coded in the body of
`pedroni99` as two matrices, `statsm` and `statsv`, one column per
deterministic case; the `intercept` column is the one `type.stat = 2` uses:

```{r}
#| label: c04-7-pco-moments
src <- deparse(pco::pedroni99)
pco_env <- new.env()
eval(parse(text = src[grep("statsm <- cbind", src):grep("colnames(statsv)",
                                                        src, fixed = TRUE)]),
     pco_env)
cbind(mean = pco_env$statsm[, "intercept"],
      variance = pco_env$statsv[, "intercept"])
```

For panel PP with an intercept (`tpanel`), mean −2.177 and variance 0.964,
against the one pair the anchor text prints for that model, $Z_{\hat t} +
1.73\sqrt{N} \to N(0, 0.93)$ [@baltagi2021, p. 361]. Pedroni's Table 2
[@pedroni1999] is not in `lit/`, so the moments are unchecked.

**`pedroni99m` fails with one regressor.** It drops the regressor sheets of
the array to a matrix when there is one of them:

```{r}
#| label: c04-7-pco-trap
#| error: true
pco::pedroni99m(panel_array(php, log(price), log(income)), type.stat = 2)
```

Use `pedroni99(Y, X)` for the bivariate case; `pedroni99m` needs three or
more series.

**Pedroni assumes independent states.** Under a common factor the Kao and
Pedroni statistics "are no longer asymptotically normal, and converge at the
rate T rather than √N T" [@gengenbach2006; @baltagi2021, p. 362]. The cure
Gengenbach, Palm and Urbain propose, defactor first and test the residuals,
is the [next step](index.qmd#sec-ur-long-run).

**Cross-unit cointegration can substantially oversize single-equation and
system panel cointegration tests** [@banerjee2004; @baltagi2021, p. 362]; the
trap is shown in [shared trends across units](index.qmd#sec-ur-cross-coint).

**The deterministic term decides the count.** The same loop with a constant
in the long-run relation, with a trend in it, and with one more lag:

```{r}
#| label: c04-7-ecdet
jo_const <- johansen_by_unit(php, c("price", "income"), log = TRUE,
                             ecdet = "const", K = 2)
table(rank = jo_const$rank)
jo_trend <- johansen_by_unit(php, c("price", "income"), log = TRUE,
                             ecdet = "trend", K = 2)
table(rank = jo_trend$rank)
jo_k3 <- johansen_by_unit(php, c("price", "income"), log = TRUE,
                          ecdet = "none", K = 3)
table(rank = jo_k3$rank)
```

With `ecdet = "const"` the ranks are 29 / 16 / 4, with `"trend"` 36 / 13 / 0,
and with `K = 3` under `"none"` 46 / 3 / 0. The role of the constant and the
trend in the long-run relation is in @johansen1994.

**T = 29 is below the range where these tests have power.** Wagner and
Hlouskova find that Pedroni's two ADF-type tests perform best and that every
other single-equation test is "severely undersized", with power "dismal for
T ≤ 25" [@wagner2010; @baltagi2021, p. 363].

**Reading `ca.jo`.** `summary()` on its result prints `Length Class Mode`
unless **urca** is attached; read `@teststat`, `@cval` and `@V` instead.
`@teststat` is ordered rank ≤ 1 first, rank 0 last. The matrix needs column
names, or `ca.jo` stops with `length of 'dimnames' [2] not equal to array
extent`. `pedroni99` resets `ka` below 2 to 2 with a warning.

**pco is archived on CRAN.** `renv::install("pco")` fails; the project
installs `pco@1.0.1` from the archive and `renv.lock` records it.

Benchmark: Stata `xtcointtest kao`, deferred. Benchmark: Stata
`xtcointtest pedroni`, deferred. Benchmark: Stata `xtcointtest westerlund`,
deferred.
:::

::: {.callout-tip collapse="true"}
## Book check

**Johansen.** The Examples section of `?ca.jo` runs the Danish money-demand
data of Johansen and Juselius [@johansen1990] (`ecdet = "const"`, `K = 2`,
quarterly dummies) and prints `summary()` of the fit; the help page itself
claims no numbers. The target numbers are the paper's: eigenvalues 0.4332,
0.1776, 0.1128, 0.0434 and maximal-eigenvalue statistics 30.09, 10.36, 6.34,
2.35. The paper is not in `lit/`, so its table is neither named nor checked
here:

```{r}
#| label: c04-7-check-denmark
data("denmark", package = "urca")
sjd <- denmark[, c("LRM", "LRY", "IBO", "IDE")]
z_dk <- urca::ca.jo(sjd, ecdet = "const", type = "eigen", K = 2,
                    spec = "longrun", season = 4)
round(z_dk@lambda[1:4], 4)
round(rev(z_dk@teststat), 2)
```

Four of four eigenvalues and four of four statistics match the paper's
values. **Pedroni.**
`?pedroni99` runs the test on OECD investment and saving shares, `gdi` and
`gds` (25 countries, 1973–2013), the panel of the Feldstein–Horioka
regression [@feldstein1980], and prints no output; the target is "runs":

```{r}
#| label: c04-7-check-pedroni
data("gdi", package = "pco")
data("gds", package = "pco")
pco::pedroni99(gdi, gds)$STATISTIC
```

Seven statistics return (panel *v* 34.69, panel rho −28.70, panel PP −8.97,
panel ADF −7.64, group rho −43.11, group PP −10.58, group ADF −9.96); nothing
to match them against. @baltagi2021 [Tables 12.3 and 12.4, pp. 371–372]
prints Kao (ADF −2.3058, p = 0.0106) and Pedroni on the R&D panel of
@kao1999rd, which no R package ships: not reproduced. For house prices no
source prints a cointegration test: None, by `.docs/panel-spat-PLAN.md`.
:::
