## Panel shape {#sec-panel-shape}

**Question.** House prices in 49 states over 29 years: how many units, how
many periods, is every state there every year, and which variables change
within a state at all?

**Intuition.** A panel is a grid, units down the side and periods across.
Three facts about the grid decide which tools apply, before any test runs.

- **Dimensions.** N units and T periods. *Micro* panels have many units and
  few periods, thousands of households over five years; *macro* panels have
  a moderate number of units over decades, twenty countries over forty years
  [@baltagi2021, p. 1]. A method's guarantees hold as one of these shapes
  grows, and a test built for a long T can fail badly when T is short
  ([when the tests lie](../04-time/index.qmd#sec-ur-traps)). With N large
  and T short, 170 regions over 20 years say, panel unit-root and
  cointegration analysis "can proceed only under restrictive assumptions"
  [@pesaran2015, p. 817].
- **Balance.** A panel is **balanced** when every unit is observed in every
  period. In an unbalanced panel, comparing each unit with its own average
  over time, the fixed-effects idea of the
  [road map](../00-road-map/index.qmd), works as before; anything that
  weights units by their own T has to be reworked [@baltagi2021,
  pp. 230–232], and several unit-root tests refuse the panel outright
  ([tests that assume independent
  units](../04-time/index.qmd#sec-ur-first-gen)).
- **Variation.** A variable that never changes within a unit, a state's
  region say, cannot explain changes within that unit: comparing the state
  with itself absorbs it. A variable that never changes across units, the
  year, is absorbed the same way by comparing each year with itself.

**The tools.** `plm::pdim()` reports N, T and balance. `plm::punbalancedness()`
measures how far a panel is from balanced, with two indices of @ahrens1981
that equal 1 for a balanced panel and fall toward 0. `plm::pvar()` lists which
variables vary within units and which across.

**Run it.** The running example first, then two panels from **pder**
[@croissant2019] chosen for their shape: `TobinQ`, large and balanced, and
`Tileries`, small and unbalanced.

```{r}
#| label: c01-1-pdim
data("HousePricesUS", package = "pder")
data("TobinQ", package = "pder")
data("Tileries", package = "pder")
plm::pdim(HousePricesUS)
plm::pdim(TobinQ)
plm::pdim(Tileries)
```

```{r}
#| label: c01-1-punbalancedness
plm::punbalancedness(Tileries)
```

```{r}
#| label: c01-1-pvar
plm::pvar(HousePricesUS)
```

A balanced panel is a full rectangle. The unbalanced one has holes, and
where they sit matters:

```{r}
#| label: c01-1-picture
#| fig-cap: "Which tile works are observed in which weeks in `Tileries`. A
#|   filled square is one observation."
#| fig-alt: "A dot grid with weeks 1 to 22 across and 25 tile works down.
#|   Most rows are filled from the first week to the last. A few rows start
#|   late or stop early, and a few have a missing square in the middle."
works <- factor(Tileries$id)
graphics::plot(Tileries$week, as.integer(works), pch = 15, cex = 0.8,
               xlab = "week", ylab = "tile works", yaxt = "n")
graphics::axis(2, at = seq_along(levels(works)), labels = levels(works),
               las = 1, cex.axis = 0.6)
```

**Read it.**

- **Balanced (`HousePricesUS`).** 49 states × 29 years, 1 421 observations,
  no gaps. N/T = 1.7, a macro shape. *So:* every tool in this book runs on
  it.
- **Balanced and large (`TobinQ`).** 188 firms × 35 years, 6 580
  observations. *So:* the same shape class, with N large enough for results
  that need it.
- **Unbalanced (`Tileries`).** 25 tile works over 12 to 22 weeks, 483
  observations. Both indices are 0.97–0.98, close to balanced, and the
  picture shows why: most works are seen in 19 weeks or more, a few start
  late, and a few skip a week. *So:* fixed-effects estimates are fine;
  tests that need a balanced panel are not.
- **Variation (`HousePricesUS`).** `state`, `region` and the other labels
  never change within a state: a state effect absorbs them. `year` never
  changes across states: a year effect absorbs it. Only `price` and `income`
  vary both ways. *So:* only they can carry a within-state slope.

::: {.callout-note appearance="simple"}
**Next.** With the shape known, ask whether the states move together:
[cross-sectional dependence](../02-cross-dependence/index.qmd).
:::

::: {.callout-note collapse="true"}
## Recipe

```r
# d: a data frame whose first two columns are the unit and the period,
# or a plm::pdata.frame(d, index = c("id", "year")).
plm::pdim(d)              # N, T, balance
plm::punbalancedness(d)   # 1, 1 when balanced
plm::pvar(d)              # which variables vary within and across units
```
:::

::: {.callout-caution collapse="true"}
## Pitfalls

**The index is guessed.** On a plain data frame, **plm** takes the first two
columns as unit and period. When they are elsewhere, say so once with
`plm::pdata.frame(d, index = c("id", "year"))` and pass that object to every
call.

**Unbalanced can also mean gappy.** A unit that starts late is one thing; a
unit that skips a period in the middle is another, because a lag or a first
difference across the gap spans two periods instead of one.
`plm::is.pconsecutive()` finds the gaps, and `Tileries` has six works with
one:

```{r}
#| label: c01-1-gaps
table(plm::is.pconsecutive(Tileries))
```
:::

::: {.callout-tip collapse="true"}
## Book check

@croissant2019 prints `n = 188, T = 35, N = 6580` for `TobinQ` [p. 32] and
`n = 25, T = 12-22, N = 483` for `Tileries` [p. 62]; the `pdim` calls above
print the same. The book prints no `punbalancedness` output.
:::
