## Testing for cross-sectional dependence {#sec-csd-tests}

::: {.callout-tip appearance="simple"}
Why CD's signed summation is both its virtue and its fatal flaw, and what its
implicit null actually is:
[cross-sectional dependence](theory-cross-sectional-dependence.qmd).
:::

**Theory.** Pesaran's CD [@pesaran2004; @pesaran2015, eq. 29.76, p. 786] is the
average pairwise residual correlation, scaled:

$$
CD = \sqrt{\frac{2T}{N(N-1)}} \sum_{i<j} \hat\rho_{ij}
  \;\xrightarrow{d}\; N(0,1).
$$

It sums raw $\hat\rho$ rather than $\hat\rho^{\,2}$, which makes it exactly
centred at zero for any *T* — and blind whenever the correlations cancel in
sign. Three later statistics attack that blindness from different directions:

| Test | Source | What it changes |
|---|---|---|
| CD | @pesaran2004 | The benchmark |
| CDw | @juodis2022 | Rademacher weights, so estimated period-specific parameters cannot make CD diverge |
| PEA | @fan2015 | A power-enhancement term that fires on sparse alternatives |
| CD\* | @pesaran2021cd | A PCA bias correction, valid whether latent factors are weak or strong |

**Data.** `HousePricesUS` [@holly2010] in **pder** [@croissant2019] for the
alternative; an independent simulated panel of the same shape for the null.
All four statistics come from `dcce::pcd_test` [@wasseja2026].

```{r}
#| label: c02-1-setup
# pcd_test()$rho_ij is exactly cor() across the N series, so the two averages
# CD is built on need no test of their own.
rho_bars <- function(m) {
  r <- stats::cor(t(m))
  diag(r) <- NA
  c(rho_bar = mean(r, na.rm = TRUE),
    abs_rho_bar = mean(abs(r), na.rm = TRUE))
}
```

**Rejecting the null.** Raw log house prices. Every test but CDw fires, and
$\bar\rho$ stays far from zero — 0.29, against an
$\overline{\lvert\rho\rvert}$ of 0.51 — which is the signature of pervasive,
same-signed co-movement rather than local correlations cancelling out. Each
statistic is asked for on its own line:

```{r}
#| label: c02-1-reject
logp <- as_matrix(log(hp$price))
dcce::pcd_test(logp, test = "pesaran")
# CDw draws Rademacher weights; it is the only one of the four needing a seed.
set.seed(1)
dcce::pcd_test(logp, test = "cdw")
dcce::pcd_test(logp, test = "pea")
dcce::pcd_test(logp, test = "cdstar")
rho_bars(logp)
```

**Accepting the null.** The same panel shape with no common component at all.
All four statistics sit inside ±1.96, which is the nominal N(0,1) they claim:

```{r}
#| label: c02-1-accept
set.seed(1)
iid <- matrix(rnorm(49 * 29), nrow = 49)
dcce::pcd_test(iid, test = "pesaran")
set.seed(1)
dcce::pcd_test(iid, test = "cdw")
dcce::pcd_test(iid, test = "pea")
dcce::pcd_test(iid, test = "cdstar")
rho_bars(iid)
```

**The non-rejection that means nothing.** Two-way fixed effects force
$\sum_i \hat e_{it} = 0$, which drives $\bar\rho$ to zero *mechanically*. CD
duly reports nothing — p ≈ 0.98, a cleaner "acceptance" than the genuine null
above. But $\overline{\lvert\rho\rvert}$ barely moves from the raw series, and
PEA and CD\* both still fire:

```{r}
#| label: c02-1-fe-artefact
fe <- plm::plm(log(price) ~ log(income), data = HousePricesUS,
               model = "within", effect = "twoways")
e <- as_matrix(residuals(fe))
dcce::pcd_test(e, test = "pesaran")
set.seed(1)
dcce::pcd_test(e, test = "cdw")
dcce::pcd_test(e, test = "pea")
dcce::pcd_test(e, test = "cdstar")
rho_bars(e)
```

Two readings of one panel, and only the average absolute correlation is honest
in both: the within transformation removed almost none of the dependence
($\overline{\lvert\rho\rvert}$ 0.51 → 0.49) while removing all of CD's ability
to see it. This is the @croissant2019 [§10.1.2.2] artefact, and the general
statement of it is @juodis2022: estimated period-specific parameters make CD
diverge even when the dependence *has* been handled.

CDw is the fix for that divergence, but it buys size at a real cost in power —
it fails to reject on the raw series too, where the dependence is not in doubt.
Read it as a guard against false positives, not as evidence of absence; CD\* is
the statistic that keeps power on the raw series and on the residuals alike.

**Book check.** `dcce::pcd_test(test = "pesaran")` should be the same statistic
`plm::pcdtest(test = "cd")` computes, and **plm** is the implementation
@croissant2019 documents:

```{r}
#| label: c02-1-book-check
# Attached because pcdtest() on a formula calls plm() by its bare name.
library(plm)
plm::pcdtest(log(price) ~ 1, data = HousePricesUS, test = "cd")
```

z = 53.26 on both, so the new package's CD is the one the book describes, and
its other three statistics are read on that credit.
