## The CD test {#sec-csd-tests}

**Question.** Do house prices in the 49 states move together more than chance
allows?

**Intuition.** Take any two states and correlate their log prices over the 29
years. Do it for all 1 176 pairs. If the states were independent, those
correlations would scatter around zero, some positive and some negative, and
their average would sit close to zero. If one common force moves every state,
the average sits well above zero. Pesaran's **CD** statistic is that average,
scaled so that it is a standard normal when the states are independent
[@pesaran2004].

::: {.callout-tip appearance="simple"}
Why the signed average is both the test's virtue and its blind spot, and what
its null really is:
[cross-sectional dependence](theory-cross-sectional-dependence.qmd).
:::

With $\hat\rho_{ij}$ the correlation between units *i* and *j* over the T
periods [@pesaran2015, eq. 29.76, p. 786]:

$$
CD = \sqrt{\frac{2T}{N(N-1)}} \sum_{i<j} \hat\rho_{ij}
  \;\xrightarrow{d}\; N(0,1).
$$

It sums the correlations with their signs, not their squares. That keeps it
centred at zero for any T, however large N is. It also makes the test blind
whenever positive and negative correlations cancel.

**The test.** CD [@pesaran2004]. Null: the units are independent, or more
precisely, any dependence among them is too weak to survive averaging. A
rejection means a pervasive, same-signed co-movement.

**Run it.** Log house prices as an N × T matrix, built with `as_matrix()`
from the setup chunk. `dcce::pcd_test` [@wasseja2026] computes CD and the
three refinements met later in this section:

```{r}
#| label: c02-1-cd
logp <- as_matrix(log(hp$price))
dcce::pcd_test(logp, test = "pesaran")
```

Two comparison panels. First an independent one of the same shape. Then the
residuals of a two-way fixed-effects regression of log price on log income, a
state effect and a year effect, which removes everything fixed about a state
and everything common to a year ([road map](../00-road-map/index.qmd)):

```{r}
#| label: c02-1-compare
set.seed(1)
iid <- matrix(rnorm(49 * 29), nrow = 49)
fe <- plm::plm(log(price) ~ log(income), data = HousePricesUS,
               model = "within", effect = "twoways")
e <- as_matrix(residuals(fe))
dcce::pcd_test(iid, test = "pesaran")
dcce::pcd_test(e, test = "pesaran")
```

The 1 176 correlations that each statistic averages:

```{r}
#| label: c02-1-picture
#| fig-cap: "The pairwise correlations behind each CD statistic: log prices
#|   (left), an independent panel of the same shape (middle), and the
#|   residuals of the two-way fixed-effects regression (right). Dashed line
#|   at zero."
#| fig-alt: "Three histograms of correlations from minus one to one. Left:
#|   a wide pile leaning to the right, most of it above zero. Middle: a
#|   narrow symmetric pile centred on zero. Right: a pile as wide as the
#|   left one but symmetric around zero, with as much mass below zero as
#|   above."
pair_cor <- function(m) {
  r <- stats::cor(t(m))
  r[lower.tri(r)]
}
op <- graphics::par(mfrow = c(1, 3))
graphics::hist(pair_cor(logp), breaks = 30, xlim = c(-1, 1),
               main = "Log prices", xlab = "pairwise correlation")
graphics::abline(v = 0, lty = 2)
graphics::hist(pair_cor(iid), breaks = 30, xlim = c(-1, 1),
               main = "Independent panel", xlab = "pairwise correlation")
graphics::abline(v = 0, lty = 2)
graphics::hist(pair_cor(e), breaks = 30, xlim = c(-1, 1),
               main = "Two-way FE residuals", xlab = "pairwise correlation")
graphics::abline(v = 0, lty = 2)
graphics::par(op)
```

Two averages summarise each histogram: the signed mean $\bar\rho$, which CD
is built on, and the mean of absolute values $\overline{\lvert\rho\rvert}$,
which ignores sign:

```{r}
#| label: c02-1-rho-bars
rho_bars <- function(m) {
  r <- pair_cor(m)
  c(rho_bar = mean(r), abs_rho_bar = mean(abs(r)))
}
rbind(prices = rho_bars(logp), independent = rho_bars(iid),
      fe_residuals = rho_bars(e))
```

**Read it.**

- **Rejects (log prices).** z = 53.26. The pile of correlations sits to the
  right of zero, $\bar\rho$ = 0.29. *So:* the states share a common force,
  and whatever is estimated on this panel has to allow for it.
- **Does not reject (independent panel).** z inside ±1.96, correlations
  scattered symmetrically around zero. *So:* this is what "no dependence"
  looks like, and CD reads it right.
- **Misleads (two-way FE residuals).** z = 0.02, p = 0.98: a cleaner
  acceptance than the genuine null above. But the histogram is as wide as
  the raw one, and $\overline{\lvert\rho\rvert}$ barely moves, 0.51 to 0.49.
  *So:* the year effect forces the residuals to sum to zero across states
  in every year, which makes their correlations cancel in sign. CD sees
  nothing because nothing *signed* is left, not because the dependence is
  gone [@croissant2019, sec. 10.1.2.2].

**Three refinements.** Each attacks CD's blindness from a different side.
None is in **plm**; all three are in `dcce::pcd_test`.

| Test | Source | What it changes |
|---|---|---|
| CDw | @juodis2022 | Random ±1 weights on the units, so a model's estimated year effects or averages cannot push CD off zero |
| PEA | @fan2015 | Adds a term that fires when a few pairs are strongly correlated, however the rest cancel |
| CD\* | @pesaran2021cd | Removes the bias from estimated common factors, so it stays valid on residuals |

The same three panels, one call per test. CDw draws random weights, so it is
the only one that needs a seed:

```{r}
#| label: c02-1-refine-prices
set.seed(1)
dcce::pcd_test(logp, test = "cdw")
dcce::pcd_test(logp, test = "pea")
dcce::pcd_test(logp, test = "cdstar")
```

```{r}
#| label: c02-1-refine-iid
set.seed(1)
dcce::pcd_test(iid, test = "cdw")
dcce::pcd_test(iid, test = "pea")
dcce::pcd_test(iid, test = "cdstar")
```

```{r}
#| label: c02-1-refine-fe
set.seed(1)
dcce::pcd_test(e, test = "cdw")
dcce::pcd_test(e, test = "pea")
dcce::pcd_test(e, test = "cdstar")
```

**Read them.**

- **On the independent panel**, all three sit inside ±1.96, as CD did. *So:*
  none raises false alarms here.
- **On the FE residuals**, PEA and CD\* both reject while CD could not. *So:*
  the dependence CD lost to sign cancellation is still there, and these two
  see it.
- **CDw does not reject even on raw prices**, where the dependence is not in
  doubt. *So:* read CDw as a guard against false positives, never as
  evidence of absence. CD\* is the one that keeps power on raw series and
  on residuals alike.

::: {.callout-note appearance="simple"}
**Next.** CD rejects: ask how strong the dependence is,
[the exponent α](index.qmd#sec-csd-exponent). CD does not reject: do not
stop. Read $\overline{\lvert\rho\rvert}$, and if the units sit on a map, ask
whether the dependence follows it ([spatial
dependence](../03-spatial/index.qmd)).
:::

::: {.callout-note collapse="true"}
## Recipe

```r
# m: an N x T matrix, one row per unit; y: a pseries or a one-sided formula.
dcce::pcd_test(m, test = "pesaran")   # CD
plm::pcdtest(y, test = "cd")          # the same statistic in plm
dcce::pcd_test(m, test = "cdstar")    # on residuals from a model with
                                      # time effects or CCE averages
```
:::

::: {.callout-caution collapse="true"}
## Pitfalls

**Residuals from any model with estimated period effects.** Two-way FE and
CCE both estimate something per period, and @juodis2022 show that plain CD
then diverges as T grows even when the dependence has been handled. On such
residuals use CD\* or CDw, never CD alone.

**`plm::pcdtest` on a formula needs `library(plm)`.** It builds a call to
`plm()` by its bare name, which the `::` prefix does not make visible.

**CDw depends on its seed.** Its weights are random draws; set the seed and
report it.
:::

::: {.callout-tip collapse="true"}
## Book check

`dcce::pcd_test(test = "pesaran")` should equal `plm::pcdtest(test = "cd")`,
the implementation @croissant2019 [sec. 4.4] document:

```{r}
#| label: c02-1-book-check
# Attached because pcdtest() on a formula calls plm() by its bare name.
library(plm)
plm::pcdtest(log(price) ~ 1, data = HousePricesUS, test = "cd")
```

z = 53.26 on both, so the new package's CD is the one the book describes,
and its other three statistics are read on that credit.
:::
