## The exponent α {#sec-csd-exponent}

**Question.** The states move together. Is that a national force that reaches
every state, or a local one that reaches a few neighbours?

**Intuition.** Average log prices across states. If the co-movement is
national, averaging over more states does not calm the average: every state
carries the same shock, so adding states adds the same thing again. If the
co-movement is local, each state shares its shock with a few others only, and
averaging over many states washes it out, the way averaging independent draws
does. The **exponent α** measures how fast averaging washes the co-movement
out. At α = 1 it never does: **strong** dependence. At α = ½ it does as fast
as for independent draws: **weak** dependence [@bailey2016].

Why an estimate cares: a year effect removes a shock that hits every state
*equally*. A national shock that hits each state *to its own degree*, with
state-specific **loadings**, survives it, and if that shock also moves income
it biases the slope ([road map](../00-road-map/index.qmd)). α says whether
such a shock is there to worry about.

::: {.callout-tip appearance="simple"}
What α counts, how to read it as an effective sample size, and why the CD
test cannot stand in for it:
[cross-sectional dependence](theory-cross-sectional-dependence.qmd).
:::

For the cross-section average $\bar z_t$ of N units [@bailey2016;
@pesaran2015, eq. 29.6, p. 754]:

$$
\operatorname{Var}(\bar z_t) = \kappa_0 N^{2(\alpha-1)} + \kappa_1 N^{-1}
  + O(N^{\alpha-2}), \qquad \tfrac12 < \alpha \le 1.
$$

The second term is what independence gives, $\sigma^2/N$. The first is the
co-movement. On log scales the variance of the average falls with N at slope
$2(\alpha - 1)$: flat at α = 1, as steep as independence, −1, at α = ½.

**The estimator.** Two, in `dcce::csd_exp` [@wasseja2026]. The **variable**
method [@bailey2016] inverts the equation above. The **residual** method
[@bailey2019] counts the pairwise correlations that survive a significance
threshold, and is the one to use on most series (see Pitfalls). Neither is a
test: α is a number to read, with a bootstrap standard error.

**Run it.** Log prices, as the N × T matrix `logp` of the
[previous section](index.qmd#sec-csd-tests):

```{r}
#| label: c02-2-alpha-strong
set.seed(1)
dcce::csd_exp(logp, use_residuals = TRUE)
```

Now remove the national force and ask again. **CCEMG**, the estimator of the
[next section](index.qmd#sec-cce), adds cross-state averages to each state's
regression, which soaks up a common shock whatever each state's loading. Its
residuals are the panel with that shock taken out:

```{r}
#| label: c02-2-alpha-weak
# pcce() builds a call to plm(), so plm must be attached, not just namespaced.
library(plm)
ccemg <- plm::pcce(log(price) ~ log(income), data = HousePricesUS,
                   model = "mg")
e_cce <- as_matrix(residuals(ccemg))
set.seed(1)
dcce::csd_exp(e_cce, use_residuals = TRUE)
```

The variance equation can be watched directly. Average random subsets of
2 to 49 states and see how fast the variance of the average falls:

```{r}
#| label: c02-2-picture
#| fig-cap: "Variance of the cross-state average against the number of
#|   states averaged, both on log scales; each point averages 200 random
#|   subsets. The slope is 2(α − 1): flat means α = 1, slope −1 means
#|   α = ½."
#| fig-alt: "Three lines on log-log axes from N = 2 to N = 49. Log prices
#|   is a nearly flat line in the middle. The independent panel at the top
#|   and the CCEMG residuals at the bottom are two parallel lines falling
#|   steeply, with slope close to minus one."
var_avg <- function(m, n, reps = 200) {
  mean(replicate(reps, var(colMeans(m[sample(nrow(m), n), , drop = FALSE]))))
}
ns <- c(2, 4, 8, 16, 32, 49)
set.seed(1)
v <- cbind(
  prices = sapply(ns, \(n) var_avg(logp, n)),
  cce_residuals = sapply(ns, \(n) var_avg(e_cce, n)),
  independent = sapply(ns, \(n) var_avg(iid, n))
)
cols <- c("black", "firebrick", "grey60")
graphics::matplot(ns, v, log = "xy", type = "b", pch = 16, lty = 1,
                  col = cols, xlab = "states averaged, N",
                  ylab = "variance of the average")
graphics::legend("bottomleft", colnames(v), col = cols, pch = 16, lty = 1,
                 bty = "n")
```

The slope of each line, turned into α:

```{r}
#| label: c02-2-slopes
slope <- function(v) coef(lm(log(v) ~ log(ns)))[[2]]
round(1 + apply(v, 2, slope) / 2, 2)
```

**Read it.**

- **Strong (log prices).** α = 0.83, far above ½, and the price line is
  nearly flat: averaging 49 states calms the average little more than
  averaging 8.
  *So:* a national force with state-specific loadings. Year effects will not
  remove it; [CCE](index.qmd#sec-cce) will.
- **Weak (CCEMG residuals).** α = 0.51, the floor, and the line falls as
  steeply as the independent panel's. *So:* once the national force is out,
  what is left is local. Only now is a spatial model a legitimate question
  ([spatial dependence](../03-spatial/index.qmd)).
- **The picture's own estimate.** The slopes give 0.85, 0.47 and 0.51, next
  to 0.83 and 0.51 from the residual method. *So:* α is not a black box; it
  is the slope of this picture with a finite-N correction, and that
  correction is where the variable method can fail (Pitfalls).

::: {.callout-note appearance="simple"}
**Next.** α near 1: [common correlated effects](index.qmd#sec-cce). α near
½: fixed effects are consistent; ask whether the dependence follows a map
([spatial dependence](../03-spatial/index.qmd)). Either way, do not route on
α alone: the spatial chapter's randomised-W test is the corroboration.
:::

::: {.callout-note collapse="true"}
## Recipe

```r
# m: an N x T matrix, one row per unit, no missing cells.
set.seed(1)
dcce::csd_exp(m, use_residuals = TRUE)    # residual method, bootstrap SE
dcce::csd_exp(m)                          # variable method
```
:::

::: {.callout-caution collapse="true"}
## Pitfalls

**The variable method collapses when the average is calm.** It is only
informative when the common component survives averaging. `price` is an
index re-based to 100 in 1980 for every state, so the cross-state average of
log price moves in a band of 4.51–4.76 and its variance is small. The
correction for $\kappa_1 N^{-1}$ then dominates and the estimate collapses
onto ½, the opposite of what CD and the residual method report. On log real
GDP in `pwt8`, whose average carries the common trend, it works:

```{r}
#| label: c02-2-alpha-variable
data("pwt8", package = "dcce")
lgdp <- do.call(rbind, by(pwt8, pwt8$country, \(d) d$log_rgdpo))
lgdp <- lgdp[, colSums(is.na(lgdp)) == 0, drop = FALSE]
c(hpus = var(colMeans(logp)), pwt8 = var(colMeans(lgdp)))
c(hpus = dcce::csd_exp(logp)$alpha, pwt8 = dcce::csd_exp(lgdp)$alpha)
```

Thirteen times more variance in the average is the whole difference. Prefer
the residual method unless the series is a level whose average carries the
common trend.

**The two methods report standard errors on different footings.** For the
variable method `csd_exp` returns `se = 1/(2 ln N √N)`, a function of N
alone: two panels with the same N get the same SE whatever they contain. The
residual method bootstraps, so its SE and its asymmetric interval respond to
the data, and to the seed. Read the first as a point estimate only.
:::

::: {.callout-tip collapse="true"}
## Book check

The **dcce** vignette [@wasseja2026] prints α = 0.8348, SE = 0.0114 and a
95% CI of [0.8124, 0.8573] for log real GDP on `pwt8`, 93 countries over 48
years, with the variable method:

```{r}
#| label: c02-2-book-check
a <- dcce::csd_exp(lgdp)
data.frame(
  vignette = c(0.8348, 0.0114, 0.8124, 0.8573),
  ours = c(a$alpha, a$se, a$ci),
  row.names = c("alpha", "se", "ci_lo", "ci_hi")
) |> round(4)
1 / (2 * log(a$N) * sqrt(a$N))
```

The estimate reproduces to four decimals, and the last line confirms that
the printed SE is exactly $1/(2\ln N\sqrt{N})$. @bailey2016 report α in the
0.77–0.92 range for macroeconomic series, which brackets both the 0.8348
here and the 0.83 on house prices above.
:::
