### Exponent of cross-sectional dependence α {#sec-csd-exponent}

**Intuition.** α counts how many units share the same shock: at α = 1 a fixed
share of the panel moves together, at α = ½ only a sparse local handful does.
It does not tell you whether you need fixed effects — it tells you whether fixed
effects are *enough*, because a common factor with unit-specific loadings is
exactly what the within transformation cannot remove.

::: {.callout-tip appearance="simple"}
Full treatment, including the effective-sample-size reading and why the CD test
cannot substitute for α: [theory note 01](theory/01-cross-sectional-dependence.qmd).
:::

**Theory.** α measures how fast averaging across units washes out their common
movement. For the cross-section average z̄ₜ [@bailey2016; @pesaran2015,
eq. 29.6, p. 754]:

$$
\operatorname{Var}(\bar z_t) = \kappa_0 N^{2(\alpha-1)} + \kappa_1 N^{-1}
  + O(N^{\alpha-2}), \qquad \tfrac12 < \alpha \le 1.
$$

- **α < 1: weak dependence.** Averaging eventually removes it. Almost all
  spatial models are of this kind [@pesaran2015, pp. 756, 802].
- **α = 1: strong dependence.** Averaging never removes it. This is the case
  of common factor models [@pesaran2015, p. 754].

Two estimators exist, and they fail in different places. The **variable**
method [@bailey2016] inverts the variance equation above directly. The
**residual** method [@bailey2019] instead sums the pairwise correlations that
survive a significance threshold, so it does not depend on the common component
surviving the averaging step.

**Data.** `HousePricesUS` [@holly2010] in **pder** [@croissant2019], 49 states
× 29 years; `pwt8` in **dcce** [@wasseja2026], 93 countries × 48 years.

Both estimators want an N × T matrix:

```{r}
#| label: s0-2-matrix
data("HousePricesUS", package = "pder")
hp <- plm::pdata.frame(HousePricesUS)
idx <- plm::index(hp)
# plm stores a panel stacked long; tapply folds it back to N x T.
as_matrix <- function(v) {
  m <- tapply(as.numeric(v), list(idx[[1]], idx[[2]]), identity)
  m[, colSums(is.na(m)) == 0, drop = FALSE]
}
lprice <- as_matrix(log(hp$price))
dim(lprice)
```

**Strong case.** Raw log house prices. α sits far above ½, which routes to
[CCE](panel-spat.qmd#sec-cce):

```{r}
#| label: s0-2-alpha-strong
dcce::csd_exp(lprice, use_residuals = TRUE)
```

**Weak case.** Now run the @bailey2016 sequence: defactor, then re-estimate α
on the residuals. CCEMG removes the common factors, and α falls to the ½ floor
— the remaining dependence is local, and only *now* is a spatial model
legitimate:

```{r}
#| label: s0-2-alpha-weak
# pcce() builds a call to plm(), so plm must be attached, not just namespaced.
library(plm)
ccemg <- plm::pcce(log(price) ~ log(income), data = HousePricesUS,
                   model = "mg")
dcce::csd_exp(as_matrix(residuals(ccemg)), use_residuals = TRUE)
```

So the same panel reads α ≈ 0.83 before defactoring and α ≈ 0.51 after. That
gap is the whole routing decision, and it is why fitting a spatial model to the
raw series would have attributed the common factor to geography.

**Where the variable method breaks.** The variance equation is only informative
when the common component survives averaging. `price` is an index re-based to
100 in 1980 for every state, so the cross-section average of log price moves in
a band of 4.51–4.76 and its variance is small. The correction term for
$\kappa_1 N^{-1}$ then dominates and the estimate collapses onto ½ — the
opposite of what CD and the residual method both report:

```{r}
#| label: s0-2-alpha-variable
data("pwt8", package = "dcce")
lgdp <- do.call(rbind, by(pwt8, pwt8$country, \(d) d$log_rgdpo))
lgdp <- lgdp[, colSums(is.na(lgdp)) == 0, drop = FALSE]
c(hpus = var(colMeans(lprice)), pwt8 = var(colMeans(lgdp)))
c(hpus = dcce::csd_exp(lprice)$alpha, pwt8 = dcce::csd_exp(lgdp)$alpha)
```

Thirteen times more variance in the average is the whole difference. Prefer the
residual method unless the series is a level whose average carries the common
trend.

::: {.callout-warning}
**The two methods report standard errors on different footings.** For the
variable method `csd_exp` returns `se = 1/(2 ln N √N)`, a function of N alone:
two panels with the same N get the same SE whatever they contain, and the 95%
CI is ±1.96 times it. The residual method bootstraps instead, so its SE and its
asymmetric percentile interval do respond to the data — 0.0190 on the raw
series above against 0.1085 on the CCEMG residuals. Read the first as a point
estimate only; the second carries genuine sampling information.
:::

**Book check.** The **dcce** vignette [@wasseja2026] prints α = 0.8348,
SE = 0.0114 and a 95% CI of [0.8124, 0.8573] for log real GDP on `pwt8`:

```{r}
#| label: s0-2-book-check
a <- dcce::csd_exp(lgdp)
data.frame(
  vignette = c(0.8348, 0.0114, 0.8124, 0.8573),
  ours = c(a$alpha, a$se, a$ci),
  row.names = c("alpha", "se", "ci_lo", "ci_hi")
) |> round(4)
1 / (2 * log(a$N) * sqrt(a$N))
```

The estimate reproduces to four decimals, and the last line confirms that the
variable method's printed SE is exactly $1/(2\ln N\sqrt{N})$. @bailey2016 report
α in the 0.77–0.92 range for macroeconomic series, which brackets both the
0.8348 here and the 0.83 on house prices above.
