## Common correlated effects {#sec-cce}

**Question.** Income rises 10%. How much do house prices rise, once the
national forces that move both are allowed for?

**Intuition.** The national force is unobserved. Its footprint is not: every
state carries it, so the cross-state average of prices, and of incomes,
carries it too, with the state-specific noise averaged away. Put those
averages into each state's regression as extra regressors, and the force is
controlled for without ever being measured. That is **common correlated
effects**, CCE [@pesaran2006]. Each state gets its own coefficient on the
averages, so each can feel the force to its own degree, which is exactly
what a year effect cannot allow.

::: {.callout-tip appearance="simple"}
Why the averages span the space of the unobserved factors, and what CCE
cannot fix: [common factors and CCE](theory-common-factors-cce.qmd).
:::

State *i* in year *t* is fitted as

$$
y_{it} = \alpha_i + \beta_i x_{it} + c_i \bar y_t + d_i \bar x_t + e_{it},
$$

with $\bar y_t$ and $\bar x_t$ the cross-state averages. **CCEMG** averages
the 49 slopes $\hat\beta_i$ [@pesaran2015, eq. 29.37, p. 767]. **CCEP** pools
the states into one slope β, keeping the state-specific $c_i$ and $d_i$
[@pesaran2015, eq. 29.41, p. 768].

**The estimators.** Three, so that the averages' effect is visible. **MG**,
mean group: the 49 state regressions without the averages, then the mean
slope. **CCEMG**: the same with the averages. **CCEP**: the pooled version.
No test here; the evidence is how the estimates move.

**Run it.** `plm::pmg` gives MG and, with `model = "cmg"`, CCEMG;
`plm::pcce` gives CCEMG again and is the only way to get CCEP
[@croissant2019, sec. 8.3.2.1]:

```{r}
#| label: c02-3-pmg
# pmg() and pcce() build a call to plm(), so plm must be attached, not just
# namespaced.
library(plm)
mg <- plm::pmg(log(price) ~ log(income), data = HousePricesUS, model = "mg")
cmg <- plm::pmg(log(price) ~ log(income), data = HousePricesUS, model = "cmg")
coef(mg)
coef(cmg)
```

```{r}
#| label: c02-3-pcce
ccemg <- plm::pcce(log(price) ~ log(income), data = HousePricesUS, model = "mg")
ccep <- plm::pcce(log(price) ~ log(income), data = HousePricesUS, model = "p")
summary(ccemg)
summary(ccep)
```

The 49 state slopes that MG and CCEMG average:

```{r}
#| label: c02-3-picture
#| fig-cap: "Income elasticity of house prices, one point per state, from
#|   the state-by-state regressions without (MG) and with (CCEMG) the
#|   cross-state averages. Thick bars: the mean of each cloud, which is the
#|   reported estimate. Dotted line at zero."
#| fig-alt: "Two vertical clouds of 49 points. The MG cloud is centred
#|   near 0.3 with about a quarter of its points below zero. The CCEMG
#|   cloud sits higher, centred near 1.1, and spreads more widely."
slopes <- data.frame(MG = mg$indcoef["log(income)", ],
                     CCEMG = ccemg$indcoef[1, ])
graphics::stripchart(slopes, vertical = TRUE, method = "jitter", pch = 16,
                     col = "grey50", ylab = "income elasticity, by state")
graphics::segments(c(0.8, 1.8), colMeans(slopes), c(1.2, 2.2),
                   colMeans(slopes), lwd = 4)
graphics::abline(h = 0, lty = 3)
```

**Read it.**

- **MG, without the averages: 0.30.** The slopes scatter around 0.30, a
  quarter of them negative. *So:* with the national force left in each
  regression, the within-state relation between income and prices is
  diluted and noisy.
- **CCEMG, with the averages: 1.14.** The whole cloud shifts up, and the
  mean is 1.14 with a standard error of 0.20. *So:* once the national force
  is controlled, a 10% rise in income goes with an 11% rise in prices. The
  gap between 0.30 and 1.14 is the leak the [road
  map](../00-road-map/index.qmd) described.
- **CCEP: 1.20.** Pooling the states gives almost the same slope. *So:* the
  spread in the CCEMG cloud is noise around one slope rather than 49
  different slopes, which @croissant2019 [p. 199] read as support for
  pooling.

::: {.callout-note appearance="simple"}
**Next.** Two checks on the residuals. Do they still move together, now
weakly ([the exponent α](index.qmd#sec-csd-exponent))? Do they wander
([tests that allow a common
factor](../04-time/index.qmd#sec-ur-second-gen))? If they do not wander,
the slope above is a long-run relation, not a spurious one.
:::

::: {.callout-note collapse="true"}
## Recipe

```r
library(plm)   # pmg() and pcce() call plm() by its bare name
plm::pmg(y ~ x, data = d, model = "mg")     # MG, no averages
plm::pcce(y ~ x, data = d, model = "mg")    # CCEMG
plm::pcce(y ~ x, data = d, model = "p")     # CCEP
```
:::

::: {.callout-caution collapse="true"}
## Pitfalls

**One average per variable, so k + 1 factors at most.** With k regressors
there are k + 1 averages, and they can proxy at most k + 1 common factors.
More factors than that, or a factor that every state loads on with the same
sign and size so that it averages to nothing, and CCE is incomplete; see
[common factors and CCE](theory-common-factors-cce.qmd).

**`coef()` of CCEMG is a plain mean.** The reported slope is the mean of
`indcoef`, so one state with a wild slope moves it. Look at the cloud, not
only its centre.
:::

::: {.callout-tip collapse="true"}
## Book check

Slope on log(income), with book values from @croissant2019 [pp. 197–199]:

```{r}
#| label: c02-3-book-check
mods <- list(MG = mg, CCEMG = ccemg, CCEP = ccep)
b <- sapply(mods, \(m) coef(m)[["log(income)"]])
se <- sapply(mods, \(m) sqrt(vcov(m)["log(income)", "log(income)"]))
data.frame(
  book_coef = c(0.302, 1.135, 1.199), coef = b,
  book_se = c(0.093, 0.195, 0.207), se = se
) |> round(3)
```

The book also prints RSS 5.66 / R² 0.74 for CCEMG and RSS 6.89 / R² 0.696
for CCEP; compare with the summaries above.
:::
