### Strong-dependence branch: CCE {#sec-cce}

::: {.callout-tip appearance="simple"}
Why cross-section averages stand in for the unobserved factors:
[theory note 04](theory/04-common-factors-cce.qmd).
:::

**Theory.** Under strong dependence, unobserved common factors bias standard
panel estimators. Common Correlated Effects (CCE) proxies the factors with
cross-section averages of y and X and adds them to each unit's regression
[@pesaran2006; @pesaran2015, p. 766]. **CCEMG** averages the unit-by-unit
slopes (eq. 29.37); **CCEP** pools the units into one slope (eq. 29.41)
[@pesaran2015, pp. 767–768].

**Data.** `HousePricesUS` [@holly2010], shipped in **pder** [@croissant2019]:
49 US states × 29 years.

Mean Group (MG) ignores the factors. `model = "cmg"` adds the averages `y.bar`
and `log(income).bar`, and the income elasticity jumps from 0.30 to 1.14:

```{r}
#| label: s0-3-pmg
# pmg() and pcce() build a call to plm(), so plm must be attached, not just
# namespaced.
library(plm)
data("HousePricesUS", package = "pder")
mg <- plm::pmg(log(price) ~ log(income), data = HousePricesUS, model = "mg")
cmg <- plm::pmg(log(price) ~ log(income), data = HousePricesUS, model = "cmg")
coef(mg)
coef(cmg)
```

`pcce` gives the same CCEMG slope and is the only way to get CCEP
[@croissant2019, sec. 8.3.2.1]:

```{r}
#| label: s0-3-pcce
ccemg <- plm::pcce(log(price) ~ log(income), data = HousePricesUS, model = "mg")
ccep <- plm::pcce(log(price) ~ log(income), data = HousePricesUS, model = "p")
summary(ccemg)
summary(ccep)
```

**Book check.** Slope on log(income), with book values from @croissant2019
[pp. 197–199]:

```{r}
#| label: s0-3-book-check
mods <- list(MG = mg, CCEMG = ccemg, CCEP = ccep)
b <- sapply(mods, \(m) coef(m)[["log(income)"]])
se <- sapply(mods, \(m) sqrt(vcov(m)["log(income)", "log(income)"]))
data.frame(
  book_coef = c(0.302, 1.135, 1.199), coef = b,
  book_se = c(0.093, 0.195, 0.207), se = se
) |> round(3)
```

The book also prints RSS 5.66 / R² 0.74 (CCEMG) and RSS 6.89 / R² 0.696
(CCEP); compare with the summaries above. CCEMG and CCEP are close, which
@croissant2019 [p. 199] read as support for pooling.
