## What a rejection means {#sec-ur-alternative}

**Question.** When a panel unit-root test rejects, what exactly has it found?

**Intuition.** One series of 30 years can barely tell a root of 0.9 from a
root of 1. A panel adds units instead of years: 49 short series carry more
evidence than one. But "the panel rejects a unit root" can mean three different
things, and the test you pick decides which one you found
[@pesaran2015, pp. 818–820]:

| | What is stationary under the alternative | Tests built for it |
|---|---|---|
| **All, alike** | every series, with the same root | LLC |
| **A share** | some series, a fixed fraction of the panel | IPS, Fisher, CIPS |
| **A handful** | a fixed number of series, however large N gets | none |

Every test averages over units. Averaging finds a share easily, because more
units bring more evidence. It loses a handful, because their signal is diluted
by the growing number of units that carry none.

**The test.** All tests share one null: *every* series has a unit root. They
differ in the alternative, and the printed output does not say which one. You
have to state it yourself.

**Run it.** Simulated panels, because only a simulation knows how many series
are truly stationary. `sim_ar_panel(n, t, k)` draws `n` independent series of
length `t`, of which the first `k` are stationary (root 0.5) and the rest are
random walks. For each panel size N, we count how often IPS rejects at 5% over
200 panels of T = 30:

```{r}
#| label: s1-2-power
reject_rate <- function(n, k, reps = 200, t = 30) {
  p <- replicate(reps, plm::purtest(sim_ar_panel(n, t, k), test = "ips",
                                    exo = "intercept",
                                    lags = 0)$statistic$p.value)
  mean(p < 0.05)
}
set.seed(1)
ns <- c(10, 30, 90)
data.frame(
  N = ns,
  none = sapply(ns, function(n) reject_rate(n, k = 0)),
  two = sapply(ns, function(n) reject_rate(n, k = 2)),
  share_40 = sapply(ns, function(n) reject_rate(n, k = 0.4 * n))
)
```

**Read it.**

- **Rejects (a share is stationary).** Column `share_40`: 40% of series are
  stationary. Power goes 0.92 → 1.00 → 1.00 as N grows. More units, more
  evidence.
- **Does not reject (nothing is stationary).** Column `none`: every series is
  a random walk, so this is the false-alarm rate. It should be 0.05. It is
  0.085–0.15, because T = 30 is short. Keep this number in mind: the test
  over-rejects even when nothing is wrong.
- **Misleads (a handful is stationary).** Column `two`: two series are
  stationary whatever N is. Power *falls* from 0.40 to 0.25 as N grows, while
  the false-alarm rate rises to 0.15. At N = 90 the test can no longer tell two
  stationary series from none.

So a rejection says "the panel as a whole does not look like N random walks".
It does not say which series are stationary, nor that all of them are. At short
T, @pesaran2015 [p. 820, Remark 8] limit the reading to this "average sense".

::: {.callout-note appearance="simple"}
**Next.** Pick a test. If your units are independent:
[tests that assume independent units](unit-roots.qmd#sec-ur-first-gen). If
they move together: [tests that allow a common
factor](unit-roots.qmd#sec-ur-second-gen).
:::

::: {.callout-caution collapse="true"}
## Pitfalls

**The output does not name the alternative.** LLC rejects towards "all,
alike" and IPS towards "a share", but both print the same word:

```{r}
#| label: s1-2-label
set.seed(1)
m <- sim_ar_panel(n = 20, t = 30, k = 8)
llc <- plm::purtest(m, test = "levinlin", exo = "intercept", lags = 0)
ips <- plm::purtest(m, test = "ips", exo = "intercept", lags = 0)
c(llc = llc$statistic$alternative, ips = ips$statistic$alternative)
```

`"stationarity"` for both, which is true of neither. Write the alternative
into the text by hand.
:::

::: {.callout-tip collapse="true"}
## Book check

None. The three alternatives are a classification
[@pesaran2015, pp. 818–820, eqs. 31.4–31.6], not a statistic, and
`.docs/panel-spat-PLAN.md` records no numerical target for this section.
:::
