## Which series are stationary? {#sec-ur-fraction}

**Question.** The panel rejects a unit root. Which units are behind it?

**Intuition.** A panel rejection says "some share is stationary" and stops.
To name the units, test each one on its own, then correct for having run N
tests: at 5%, 49 tests produce two or three false alarms by chance alone.
@hanck2013 does this with two classical tools. **Simes' test** [@simes1986]
gives one global p-value that stays valid when units are correlated.
**Hommel's procedure** [@hommel1988] then decides unit by unit, keeping the
chance of *any* false alarm at 5%.

**The test.** Input: the per-unit p-values that a Maddala–Wu or IPS test has
already computed. Output: a global verdict and a list of named units.

**Run it.** Price levels, through the Maddala–Wu test of the
[first-generation section](unit-roots.qmd#sec-ur-first-gen):

```{r}
#| label: s1-5-levels
mw <- plm::purtest(log(php$price), test = "madwu", lags = 2, exo = "trend")
h_lev <- plm::phansitest(mw)
c(states = length(h_lev$p), named = h_lev$rejected.no)
```

`phansitest` names units but does not print Simes' global p-value correctly
(see Pitfalls), so compute it from its formula: sort the p-values and take
the smallest N·p₍ⱼ₎/j.

```{r}
#| label: s1-5-simes
simes <- function(p) min(length(p) * sort(p) / seq_along(p))
simes(h_lev$p)
```

**Read it: does not reject (price levels).** None of the 49 states is named,
and Simes' global p-value is 0.12. No state's price level is shown to be
stationary.

**Read it: rejects (real exchange rates).** The 19 countries of @hanck2013
[Table 11], with the per-country p-values printed in `?phansitest`:

```{r}
#| label: s1-5-hanck
pvals <- c(0.0001, 0.0001, 0.0001, 0.0001, 0.0001, 0.0001, 0.0050, 0.0050,
           0.0050, 0.0050, 0.0175, 0.0175, 0.0200, 0.0250, 0.0400, 0.0500,
           0.0575, 0.2375, 0.2475)
names(pvals) <- c("Argentina", "Sweden", "Norway", "Mexico", "Italy",
                  "Finland", "France", "Germany", "Belgium", "U.K.", "Brazil",
                  "Australia", "Netherlands", "Portugal", "Canada", "Spain",
                  "Denmark", "Switzerland", "Japan")
h_rer <- plm::phansitest(pvals)
simes(pvals)
h_rer$name[h_rer$rejected]
```

Simes' p-value is 0.0003, and ten countries are named. Both answers a short-T
panel can give: the panel is not all random walks, *and* these ten are the
stationary ones.

**Read it: misleads (growth rates).** Growth rates rejected the panel null at
p < 2.2 × 10⁻¹⁶. How many states are named?

```{r}
#| label: s1-5-growth
mw_g <- plm::purtest(diff(log(php$price)), test = "madwu", lags = 1,
                     exo = "intercept")
h_g <- plm::phansitest(mw_g)
c(states = length(h_g$p), named = h_g$rejected.no)
```

Only four. The other 45 are not shown to wander; their individual tests have
28 years each and too little power to clear the multiple-testing bar. A
missing name means "not proven stationary", never "has a unit root".

::: {.callout-note appearance="simple"}
**Next.** Report the global verdict *and* the named units. Then check whether
the test itself can be trusted at your panel's shape:
[when the tests lie](unit-roots.qmd#sec-ur-traps).
:::

::: {.callout-note collapse="true"}
## Recipe

```r
mw <- plm::purtest(y, test = "madwu", exo = "intercept", lags = "AIC",
                   pmax = 4)
h <- plm::phansitest(mw)
h$name[h$rejected]                          # the named units
min(length(h$p) * sort(h$p) / seq_along(h$p))  # Simes' global p-value
```
:::

::: {.callout-caution collapse="true"}
## Pitfalls

**The printed verdict is wrong.** In **plm** 2.6.7 the print method says "H0
rejected (globally)" whatever the result. Here it is on price levels, where
no state is named and Simes gives 0.12:

```{r}
#| label: s1-5-print-trap
h_lev
```

Read `$rejected.no` and compute Simes' p-value yourself.

**Hadri is refused.** `phansitest` needs per-unit p-values; Hadri's statistic
is not built from them:

```{r}
#| label: s1-5-hadri
hadri <- plm::purtest(log(php$price), test = "hadri", lags = 2, exo = "trend")
tryCatch(plm::phansitest(hadri), error = conditionMessage)
```
:::

::: {.callout-tip collapse="true"}
## Book check

@hanck2013 [Table 11] rejects the individual unit-root null for the ten
countries with the smallest p-values and no others. The run above names the
same ten: Argentina to the U.K. Spain, at an individual p of 0.05, is not
among them: testing 19 hypotheses at once is what costs it.
:::
