## Poolability {#sec-core-pool}

**Question.** Can one investment equation serve all ten firms, and one
price–income elasticity all 49 states, or does each need its own?

**Intuition.** Fit one regression to everyone, then one to each unit, and
compare the residual sums of squares. If the single line fits almost as
well as the ten, pooling costs nothing and buys precision; if it fits much
worse, the units do not share the line. Chow's F does exactly this
[@chow1960; @baltagi2021, p. 77]. The restricted model can be pooled OLS,
which forces all coefficients to be equal, or the within regression, which
gives every unit its own intercept and forces only the slopes to be equal.
Under random effects the errors of one unit are correlated across years,
the Chow F is no longer an F, and it rejects a true null far too often
[@baltagi1981; @baltagi2021, p. 79]. The **Roy–Zellner test** runs the same
F on the data after the GLS transformation, where the errors are spherical
again [@roy1957; @zellner1962; @baltagi2021, pp. 78–79].

::: {.callout-tip appearance="simple"}
What "one regression" assumes about the units, and why the transformation
restores the F: [fixed and random effects](theory-fixed-random.qmd).
:::

With $e$ the residuals of the restricted model and $e_i$ those of unit
$i$'s own regression, $K'$ counting the intercept [@baltagi2021, eq. 4.8],

$$
F = \frac{\left(e'e - \sum_i e_i'e_i\right) / (N - 1)K'}
         {\sum_i e_i'e_i / N(T - K')}
  \;\sim\; F\big((N - 1)K',\, N(T - K')\big).
$$

The Roy–Zellner statistic is the same ratio after $y$ and $X$ have been
premultiplied by $\Omega^{-1/2}$, the inverse square root of the
error-component covariance [@baltagi2021, eq. 4.14]; with one-way effects
that is quasi-demeaning by $\theta$, each variable less $\theta$ times its
unit mean.

**The tests.** `pooltest` [@croissant2019, sec. 8.2.3]: `model =
"pooling"` restricts all coefficients, `model = "within"` the slopes only,
and `effect = "time"` asks the same question across years. `roy_zellner()`
from `R/` is the GLS version. Null: the coefficients are the same for every
unit. A rejection means the units do not share the line.

**Run it, Grunfeld.** Investment on market value and capital stock, ten
firms, twenty years:

```{r}
#| label: c05-1-gr
plm::pooltest(fg, data = gr, model = "pooling")
plm::pooltest(fg, data = gr, model = "within")
plm::pooltest(fg, data = gr, model = "pooling", effect = "time")
roy_zellner(fg, gr)
```

**Run it, house prices.** Log price on log income, 49 states, 29 years.
The unrestricted fit, one regression per state, is kept: the next step
reads its slopes.

```{r}
#| label: c05-1-hp
plm::pooltest(fh, data = hp, model = "pooling")
plm::pooltest(fh, data = hp, model = "within")
hp_np <- plm::pvcm(fh, data = hp, model = "within")
```

The dispersion the F measures, drawn: each unit's own line over the pooled
one.

```{r}
#| label: c05-1-picture
#| fig-cap: "Unit-by-unit regression lines (thin) over the pooled line
#|   (thick): investment on market value for the ten firms (left), log
#|   price on log income for the 49 states (right)."
#| fig-alt: "Two scatterplots with many thin fitted lines and one thick
#|   one. On the left, ten thin lines fan out from near the origin at
#|   slopes from almost flat to steep, around a thick line of middling
#|   slope. On the right, 49 thin lines cross one another at every angle,
#|   some sloping down, over a thick line that slopes gently up."
unit_lines <- function(y, x, g, main, xlab, ylab) {
  plot(x, y, pch = 16, cex = 0.5, col = "grey75", main = main,
       xlab = xlab, ylab = ylab)
  for (k in levels(g)) {
    i <- g == k
    graphics::abline(lm(y[i] ~ x[i]), col = "grey40")
  }
  graphics::abline(lm(y ~ x), lwd = 3)
}
op <- graphics::par(mfrow = c(1, 2))
unit_lines(as.numeric(gr$inv), as.numeric(gr$value), plm::index(gr)[[1]],
           "Grunfeld", "market value", "investment")
unit_lines(as.numeric(hp$lp), as.numeric(hp$li), plm::index(hp)[[1]],
           "House prices", "log income", "log price")
graphics::par(op)
```

**Read it.**

- **Rejects (across units, both panels).** Grunfeld 27.7 on F(27, 170) for
  all coefficients and 5.78 on F(18, 170) for the slopes alone; house
  prices 25.8 on F(96, 1323) and 16.1 on F(48, 1323). *So:* neither panel
  pools, even when every unit keeps its own intercept.
- **Does not reject (Grunfeld across years).** 1.12 on F(57, 140), p =
  0.29. *So:* the firms' lines are stable over time; the heterogeneity is
  across firms, not across years.
- **Misleads (the plain Chow F under random effects).** Roy–Zellner gives
  4.35 on F(27, 170), still a rejection, but a sixth of the plain 27.7.
  *So:* with a firm effect this large (72 % of the error variance, step 4)
  most of the plain F was the firm effect, not the slopes; Baltagi's Monte
  Carlo shows the plain F rejecting a true null often when the variance
  components are large [@baltagi1981; @baltagi2021, p. 79].

::: {.callout-note appearance="simple"}
**Next.** Rejected: how far apart are the slopes, and is the average still
worth estimating ([slope heterogeneity](index.qmd#sec-core-slopes))? Not
rejected: pooled OLS, then the [effects tests](index.qmd#sec-core-effects),
which decide whether even the intercepts need to differ.
:::

::: {.callout-note collapse="true"}
## Recipe

```r
# f: a formula; d: a pdata.frame. plm must be attached for pooltest.
plm::pooltest(f, data = d, model = "pooling")   # all coefficients
plm::pooltest(f, data = d, model = "within")    # slopes only
plm::pooltest(f, data = d, model = "pooling", effect = "time")
roy_zellner(f, d)                               # under random effects
```
:::

::: {.callout-caution collapse="true"}
## Pitfalls

**`pooltest` needs plm attached.** Its formula method calls `plm()`
unqualified, and `plm::pooltest()` alone fails with "could not find
function". The setup chunk attaches the package for that reason.

**Two nulls, one function.** `model = "pooling"` tests all coefficients,
`model = "within"` the slopes only. Baltagi's "poolability across firms"
is the first, "poolability of slopes only" the second [@baltagi2021, p.
80]. Report which one was run.

**Rejected does not mean no pooled slope.** The pooled slope with unit
effects is still the average effect; what the rejection says is that the
dispersion around it has to be reported beside it. That is the mean-group
estimator of the [cross-dependence
chapter](../02-cross-dependence/index.qmd#sec-cce) and the next step.

**Roy–Zellner is not in plm.** `roy_zellner()` is one-way and refuses an
unbalanced panel. It uses the Swamy–Arora θ of `plm::ercomp()`; Baltagi's
Monte Carlo used the Amemiya and Wallace–Hussain variance estimates, and
both gave the test the right size [@baltagi2021, p. 79].
:::

::: {.callout-tip collapse="true"}
## Book check

@baltagi2021 [sec. 4.1.3, Example 1, p. 80] prints, for Grunfeld, 27.75 on
F(27, 170) for all coefficients, 5.78 on F(18, 170) for the slopes, 1.12 on
F(57, 140) across time, and 4.35 on F(27, 170) for Roy–Zellner across
firms. The `c05-1-gr` chunk above gives 27.749, 5.7805, 1.1204 and 4.3466:
all four reproduced.

@croissant2019 [sec. 8.2.3, Example 8.4, p. 194] print 26 on df (96, 1300)
against pooling and 16 on df (48, 1300) against within for house prices,
with two significant digits. The `c05-1-hp` chunk gives 25.778 and 16.074
on df (96, 1323) and (48, 1323): the statistics reproduce; the printed
denominator df is rounded to 1300, and the exact value is
$N(T - K - 1) = 49 \times 27 = 1323$.
:::
