## Slope heterogeneity {#sec-core-slopes}

**Question.** When the slopes differ, how much do they differ, and is the
spread more than sampling noise would give?

**Intuition.** Estimate each unit's own regression and look at the spread
of the slopes. Part of the spread is sampling noise, because each slope
comes from only $T$ observations; the rest is real. Swamy's
**random-coefficients model** treats each unit's coefficient vector as a
draw around a common mean and estimates the variance of the draws, net of
the noise [@swamy1970; @croissant2019, sec. 8.2.2]. A large estimated
variance says the units genuinely differ; a mean with a standard deviation
of its own size says the average is one number among many. The **mean
group** estimator of the [cross-dependence
chapter](../02-cross-dependence/index.qmd#sec-cce) is the plain average of
the unit slopes, and its dispersion is their empirical covariance; it is the
Swamy estimator with equal weights, and the two converge as $T$ grows
[@croissant2019, p. 190].

::: {.callout-tip appearance="simple"}
Random coefficients as an error-component model, and why their covariance
can come back negative: [the error of a panel
regression](theory-panel-error.qmd).
:::

With $\hat\beta_n$ unit $n$'s OLS slopes, $\bar\beta$ their average and
$\hat\sigma_n^2$ unit $n$'s residual variance, Swamy's estimate of the
dispersion of the draws is [@croissant2019, p. 188]

$$
\hat\Delta = \frac{1}{N - 1}\sum_{n}
  (\hat\beta_n - \bar\beta)(\hat\beta_n - \bar\beta)'
  \;-\; \frac{1}{N}\sum_{n} \hat\sigma_n^2 (Z_n'Z_n)^{-1},
$$

the raw spread of the slopes less what sampling noise contributes. Swamy's
own test of equal slopes is the weighted sum of squared distances of the
$\hat\beta_n$ from a pooled estimate, $\chi^2$ with $(N - 1)K$ degrees of
freedom as $T \to \infty$ with $N$ fixed [@swamy1970; @baltagi2021, p.
102]. For the panels this book worries about, many units and few periods,
@pesaran2008 standardise that statistic: their $\Delta$ is
$\sqrt{N}\,(S/N - K)/\sqrt{2K}$, normal as $N$ and $T$ grow together under
a rate condition of the form $\sqrt{N}/T \to 0$ or $\sqrt{N}/T^2 \to 0$,
depending on which residual variance it uses; the adjusted version keeps
its size even when $T$ is very small relative to $N$ [@pesaran2008;
@baltagi2021, p. 102]. @blomquist2013 give a version robust
to serial correlation.

**The tests.** `pvcm(model = "within")` [@croissant2019, p. 187] and its
`summary()`, the spread of the unit slopes; `pvcm(model = "random")`,
Swamy's mean and `$Delta`, and the homogeneity test its `summary()` prints;
`pmg(model = "mg")`, the mean-group comparison. Null, for the homogeneity
test: the slopes are the same for every unit [@swamy1970; @baltagi2021, p.
102]. A rejection means the spread is more than noise.

::: {.callout-important appearance="simple"}
**Limitation.** Pesaran–Yamagata's $\Delta$ and its HAC (heteroskedasticity-
and autocorrelation-consistent) version have no implementation in the
packages this book uses; the Swamy test printed below
is the $N$ fixed, $T \to \infty$ statistic it was built to replace.
Benchmark: Stata `xthst` [@bersvendsen2021], deferred.
:::

**Run it, Grunfeld.** The spread of the ten firms' slopes, then Swamy's
mean with the standard deviation of the draws beside it:

```{r}
#| label: c05-2-gr-np
gr_np <- plm::pvcm(fg, data = gr, model = "within")
summary(gr_np)
```

```{r}
#| label: c05-2-gr-swamy
gr_sw <- plm::pvcm(fg, data = gr, model = "random")
summary(gr_sw)
cbind(mean = coef(gr_sw), sd = sqrt(diag(gr_sw$Delta)))
```

**Run it, house prices.** The same on the 49 state slopes from
[poolability](index.qmd#sec-core-pool), with the mean-group average beside
Swamy's:

```{r}
#| label: c05-2-hp-np
summary(hp_np)
```

```{r}
#| label: c05-2-hp-swamy
hp_sw <- plm::pvcm(fh, data = hp, model = "random")
summary(hp_sw)
hp_mg <- plm::pmg(fh, data = hp, model = "mg")
cbind(Swamy = coef(hp_sw), MG = coef(hp_mg))
```

The spread, drawn: a histogram of the unit slopes for each panel, with the
two averages marked.

```{r}
#| label: c05-2-picture
#| fig-cap: "The unit slopes: the ten firms' coefficients on market value
#|   (left) and the 49 states' income elasticities (right). The solid
#|   line is Swamy's mean, the dashed line the mean-group average."
#| fig-alt: "Two histograms. On the left, ten bars spread between 0 and
#|   0.18 with two vertical lines close together near 0.09. On the right,
#|   49 bars spread from about minus 1.1 to 2, with most mass between
#|   minus 0.5 and 1 and two vertical lines close together near 0.3."
gr_mg <- plm::pmg(fg, data = gr, model = "mg")
op <- graphics::par(mfrow = c(1, 2))
graphics::hist(coef(gr_np)$value, breaks = 10, col = "grey80",
               main = "Grunfeld: slope on value", xlab = "slope")
graphics::abline(v = coef(gr_sw)["value"], lwd = 2)
graphics::abline(v = coef(gr_mg)["value"], lwd = 2, lty = 2)
graphics::hist(coef(hp_np)$li, breaks = 15, col = "grey80",
               main = "House prices: income elasticity", xlab = "slope")
graphics::abline(v = coef(hp_sw)["li"], lwd = 2)
graphics::abline(v = coef(hp_mg)["li"], lwd = 2, lty = 2)
graphics::par(op)
```

**Read it.**

- **Rejects (house prices).** Swamy's homogeneity test 2440 on 96 degrees
  of freedom. The state elasticities run from −1.14 to 2.04; Swamy's mean
  is 0.287 and the mean-group average 0.302, and the standard deviation of
  the draws, 0.63, is twice the mean. *So:* the income elasticity is not
  one number, and the mean-group choice of the cross-dependence chapter
  was right: report the average with its spread.
- **Rejects (Grunfeld).** 901 on 27 degrees of freedom. The slopes on
  market value run from 0.005 to 0.175 around a mean of 0.085, with a
  standard deviation of the draws of 0.056. *So:* the firms differ in
  slope as much as in scale; the pooled slope of the next steps is an
  average over firms that do not share it.
- **Does not reject.** No case on these panels. A panel whose unit lines
  agree would print a small homogeneity statistic and a `$Delta` diagonal
  near zero; neither panel here comes close.

::: {.callout-note appearance="simple"}
**Next.** Whatever the spread, the [effects
tests](index.qmd#sec-core-effects) come next: a pooled slope with unit
effects is the average effect, and the question is what kind of effect
sits under it.
:::

::: {.callout-note collapse="true"}
## Recipe

```r
# f: a formula; d: a pdata.frame.
np <- plm::pvcm(f, data = d, model = "within")   # one fit per unit
summary(np)                                      # spread of the slopes
sw <- plm::pvcm(f, data = d, model = "random")   # Swamy
cbind(coef(sw), sqrt(diag(sw$Delta)))            # mean, sd of the draws
plm::pmg(f, data = d, model = "mg")              # mean group
```
:::

::: {.callout-caution collapse="true"}
## Pitfalls

**Swamy's covariance can come back non-positive-definite.** The second
term of $\hat\Delta$ is subtracted, and with few periods it can exceed the
first. The mean-group dispersion omits that term, is biased but always
non-negative definite, and Swamy himself proposed it as the fallback
[@croissant2019, p. 190].

**Swamy's test is a large-$T$ test.** Its $\chi^2$ holds as $T \to
\infty$ with $N$ fixed [@baltagi2021, p. 102]; on a panel with many units
and few periods it over-rejects, and the extent of the over-rejection
shrinks only as $T$ grows relative to $N$ [@pesaran2008; @baltagi2021, p.
102]. $\Delta$ is the test built for $N \gg T$, and its absence here is a
gap of the tools, not of the method. Benchmark: Stata `xthst`, deferred.

**The Hausman-type test has no power.** A test of the fixed-effects slope
against the mean-group average keeps its size but has no power against
random draws around a common mean, whatever the sample size
[@pesaran2008; @baltagi2021, p. 102]. Do not read "FE equals MG" as
homogeneity.
:::

::: {.callout-tip collapse="true"}
## Book check

@croissant2019 [sec. 8.2.2, Example 8.2, p. 190] print Swamy 0.2867 and
mean group 0.3018 for the income elasticity: the `c05-2-hp-swamy` chunk
gives 0.2867 and 0.3018, reproduced. Their `summary(housep.np)`
[@croissant2019, sec. 8.2.3, Example 8.4, pp. 193–194] prints mean 0.302,
min −1.141, max 2.037 for the state elasticities: the `c05-2-hp-np` chunk
gives 0.3018, −1.1409 and 2.0369, reproduced. Their Figure 8.1
[@croissant2019, p. 193] is the right-hand histogram above.
No target for Pesaran–Yamagata: no implementation in R.
:::
