## Fixed or random {#sec-core-fe-re}

**Question.** Does a firm's unobserved character move with its capital
stock? If it does, random effects is wrong; if it does not, fixed effects
throws away precision.

**Intuition.** Fixed effects uses only the variation inside each unit over
time, so whatever the unit effect is correlated with, it cannot bias the
slope. Random effects also uses the variation between units, the
differences in their means, and that is where a correlation between the
effect and the regressors would bite. Under the null of no correlation both
estimators are consistent and random effects is efficient; under the
alternative only fixed effects is consistent [@baltagi2021, pp. 89–90]. The
**Hausman test** compares the two estimates: far apart means the between
variation carries the correlation [@hausman1978; @baltagi2021, eq. 4.41].
The same test is an F on extra regressors in an augmented regression, which
is how it is made robust [@arellano1993; @baltagi2021, p. 92]. Mundlak's
model says why: if the unit effect is a linear function of the unit's mean
regressors, GLS on the model augmented with those means is the within
estimator, and the test is whether the means matter [@mundlak1978;
@croissant2019, p. 90]. When a regressor never moves over time, education
in a wage equation, fixed effects cannot estimate its coefficient at all;
**Hausman–Taylor** keeps it, instrumenting the endogenous regressors with
the exogenous ones [@hausman1981; @baltagi2021, sec. 7.4].

::: {.callout-tip appearance="simple"}
GLS as a weighted average of within and between, the three equivalent
Hausman contrasts, Chamberlain's Π matrix and Hausman–Taylor's instrument
set: [fixed and random effects](theory-fixed-random.qmd).
:::

With $\hat q = \hat\beta_{GLS} - \hat\beta_{Within}$, the covariance of the
contrast is the difference of the covariances, because the efficient
estimator is uncorrelated with its distance from any other consistent one
[@baltagi2021, eq. 4.40]:

$$
m = \hat q'\,[\operatorname{var}(\hat\beta_{Within})
  - \operatorname{var}(\hat\beta_{GLS})]^{-1}\,\hat q \;\sim\; \chi^2_K .
$$

The augmented regression puts the within-transformed regressors next to
the GLS-transformed ones [@baltagi2021, eq. 4.42],

$$
y^* = X^*\beta + \tilde X\gamma + w,
$$

and tests $\gamma = 0$; the coefficient on $\tilde X$ is
$\hat\beta_{Within} - \hat\beta_{Between}$, and the three contrasts
(GLS–within, GLS–between, within–between) give the same statistic
[@hausman1981; @baltagi2021, eqs. 4.46–4.48 and p. 91]. A robust covariance
in that regression is the robust Hausman test [@arellano1993; @baltagi2021,
p. 92].

**The tests.** `phtest` [@croissant2019, sec. 4.2.2, Examples 5.9–5.10]:
`method = "chisq"`, the contrast; `method = "aux"`, the augmented
regression; `method = "aux", vcov = plm::vcovHC`, the same with clustered
errors. Null: the unit effect is uncorrelated with the regressors, so both
estimators are consistent [@baltagi2021, p. 90]. A rejection means random
effects is inconsistent. Chamberlain's test of the fixed-effects
restrictions and the
Angrist–Newey version are `piest` and `aneweytest` [@chamberlain1982;
@angrist1991; @croissant2019, pp. 94–95]; Hausman–Taylor is `plm(model =
"random", random.method = "ht", inst.method = "baltagi")`, with
`inst.method = "am"` for the Amemiya–MaCurdy instruments [@amemiya1986].

**Run it, Grunfeld.** The one-way contrast three ways, then the two-way
contrast under two variance-component estimators:

```{r}
#| label: c05-4-gr
gr_re <- plm::plm(fg, data = gr, model = "random")
plm::ercomp(gr_re)
plm::phtest(gr_w, gr_re)
plm::phtest(fg, data = gr, method = "aux")
plm::phtest(fg, data = gr, method = "aux", vcov = plm::vcovHC)
```

```{r}
#| label: c05-4-gr-two
gr_re2 <- plm::plm(fg, data = gr, model = "random", effect = "twoways")
plm::ercomp(gr_re2)
plm::phtest(gr_w2, gr_re2)
gr_re2a <- plm::plm(fg, data = gr, model = "random", effect = "twoways",
                    random.method = "amemiya")
plm::ercomp(gr_re2a)
plm::phtest(gr_w2, gr_re2a)
```

**Run it, house prices.** The same on the precomputed logs:

```{r}
#| label: c05-4-hp
hp_re <- plm::plm(fh, data = hp, model = "random")
plm::ercomp(hp_re)
plm::phtest(hp_w, hp_re)
plm::phtest(fh, data = hp, method = "aux")
plm::phtest(fh, data = hp, method = "aux", vcov = plm::vcovHC)
```

```{r}
#| label: c05-4-hp-two
hp_re2 <- plm::plm(fh, data = hp, model = "random", effect = "twoways")
plm::ercomp(hp_re2)
plm::phtest(hp_w2, hp_re2)
hp_re2a <- plm::plm(fh, data = hp, model = "random", effect = "twoways",
                    random.method = "amemiya")
plm::phtest(hp_w2, hp_re2a)
```

**Run it, wages.** The Cornwell–Rupert equation on `Wages`, 595 workers
over 1976–82 [@cornwell1988; @baltagi2021, pp. 173–174]: log wage on weeks
worked, region, city, marital status, experience and its square,
occupation, industry, union, sex, race and education. The second part of
the formula lists the exogenous regressors, time-varying and not; the
third the endogenous time-varying ones. Education is time-invariant and
endogenous, the case the estimator exists for.

```{r}
#| label: c05-4-wages
fw <- lwage ~ wks + south + smsa + married + exp + I(exp^2) + bluecol +
  ind + union + sex + black + ed |
  bluecol + south + smsa + ind + sex + black |
  wks + married + union + exp + I(exp^2)
wg_ht <- plm::plm(fw, data = wg, model = "random", random.method = "ht",
                  inst.method = "baltagi")
summary(wg_ht)
wg_am <- plm::plm(fw, data = wg, model = "random", random.method = "ht",
                  inst.method = "am")
coef(summary(wg_am))[c("(Intercept)", "ed"), ]
```

The two Hausman tests of the wage equation: GLS against within, which
says whether the worker effect is correlated with anything; and
Hausman–Taylor against within, which says whether the instruments chosen
as exogenous are [@baltagi2021, p. 175]. Then the Angrist–Newey test of the
fixed-effects restrictions themselves, on the time-varying regressors. Its
restrictions multiply with $T$, and the Chamberlain test, which reaches
the same decision, understates its own variance as $T$ grows [@baltagi2021,
p. 93], so the seven-year wage panel suits it better than Grunfeld's twenty
years:

```{r}
#| label: c05-4-wages-tests
fw1 <- lwage ~ wks + south + smsa + married + exp + I(exp^2) + bluecol +
  ind + union + sex + black + ed
wg_w <- plm::plm(fw1, data = wg, model = "within")
wg_re <- plm::plm(fw1, data = wg, model = "random")
plm::phtest(wg_re, wg_w)
plm::phtest(wg_ht, wg_w)
plm::aneweytest(update(fw1, . ~ . - sex - black - ed), data = wg)
```

The contrast, drawn: the within, between and random-effects estimates of
each slope with their 95 % bands. The Hausman test asks whether the within
and between dots could be the same number.

```{r}
#| label: c05-4-picture
#| fig-cap: "Within, between and random-effects estimates with 95 % bands:
#|   Grunfeld's slopes on market value and capital (left and centre), the
#|   house-price income elasticity (right)."
#| fig-alt: "Three dot-and-whisker panels, each with three dots labelled
#|   within, between and RE. For market value the three dots sit near 0.11
#|   to 0.13 with overlapping bands. For capital the within and RE dots sit
#|   at 0.31 with short bands and the between dot at 0.03 with a band
#|   wide enough to reach them. For the income elasticity the within and
#|   RE dots sit at 0.35 and the between dot at 0.50 with a wide band."
gr_b <- plm::plm(fg, data = gr, model = "between")
hp_b <- plm::plm(fh, data = hp, model = "between")
dot_whisker <- function(est, se, main) {
  k <- length(est)
  plot(est, k:1, xlim = range(est - 2 * se, est + 2 * se),
       ylim = c(0.5, k + 0.5), pch = 16, yaxt = "n", ylab = "",
       xlab = "estimate", main = main)
  graphics::segments(est - 1.96 * se, k:1, est + 1.96 * se, k:1)
  graphics::axis(2, at = k:1, labels = names(est), las = 1)
}
pick <- function(fits, term) {
  est <- sapply(fits, function(f) coef(f)[term])
  se <- sapply(fits, function(f) sqrt(diag(vcov(f)))[term])
  names(est) <- names(se) <- names(fits)
  list(est = est, se = se)
}
gr_fits <- list(within = gr_w, between = gr_b, RE = gr_re)
hp_fits <- list(within = hp_w, between = hp_b, RE = hp_re)
op <- graphics::par(mfrow = c(1, 3), mar = c(4, 5, 2, 1))
v <- pick(gr_fits, "value"); dot_whisker(v$est, v$se, "value")
v <- pick(gr_fits, "capital"); dot_whisker(v$est, v$se, "capital")
v <- pick(hp_fits, "li"); dot_whisker(v$est, v$se, "income elasticity")
graphics::par(op)
```

**Read it.**

- **Does not reject (one-way, both panels; HT on wages).** Grunfeld 2.33
  on 2 df, p = 0.31; house prices 1.39 on 1 df, p = 0.24; Hausman–Taylor
  against within on wages 5.26. *So:* random effects is admissible on both
  one-way models, and the instruments chosen for the wage equation are
  legitimate.
- **Rejects (two-way, both panels; GLS against within on wages).**
  Grunfeld 13.5 (Swamy–Arora) or 8.96 (Amemiya); house prices 58.6 or
  25.6; wages 5075 on 9 df. *So:* once the year effect is in, the firm (or
  state) effect is correlated with the regressors; on wages the worker
  effect is correlated with education, and Hausman–Taylor recovers a
  return to schooling of 0.138 that within cannot estimate at all.
- **Rejects (Angrist–Newey on wages).** 484.26 on 369 df, p = 5e-05.
  *So:* the restrictions the within model imposes do not hold on the wage
  panel either; a Hausman rejection names random effects as wrong, it does
  not name fixed effects as right [@baltagi2021, p. 93].
- **Misleads (the clustered `aux` test on Grunfeld).** 8.30, p = 0.016,
  a rejection where the three classical forms agree on 2.1 to 2.3. *So:*
  the Arellano covariance here is an average over ten firms, and
  clustering by unit rests on $N$-asymptotics, a large, short panel
  [@croissant2019, p. 111]; Baltagi's own robust F on the same regression
  is 1.58, p = 0.21 [@baltagi2021, p. 95]. Read a robust test on ten
  clusters as a warning, not a verdict.

::: {.callout-note appearance="simple"}
**Next.** Not rejected: random effects, and the serial-correlation tests
that condition on it ([serial correlation](index.qmd#sec-core-serial)).
Rejected: fixed effects, and then the first-difference comparison in the
same step, which asks whether fixed effects is even the right way to remove
the unit effect.
:::

::: {.callout-note collapse="true"}
## Recipe

```r
# f: a formula; d: a pdata.frame; w, re: within and random fits of f.
plm::phtest(w, re)                                   # the contrast
plm::phtest(f, data = d, method = "aux")             # augmented regression
plm::phtest(f, data = d, method = "aux", vcov = plm::vcovHC)   # robust
# Hausman-Taylor: f3 is y ~ x | exogenous | endogenous time-varying.
plm::plm(f3, data = d, model = "random", random.method = "ht",
         inst.method = "baltagi")
```
:::

::: {.callout-caution collapse="true"}
## Pitfalls

**`vcov` is honoured only under `method = "aux"`.** Under the default
`"chisq"` it is silently ignored and the classical 2.33 comes back.

**`"aux"` breaks on `log()` in the formula.** It builds the auxiliary
formula from variable names, and `log(income).tilde` does not parse. That
is why the setup chunk precomputes `lp` and `li`:

```{r}
#| label: c05-4-pitfall-log
#| error: true
plm::phtest(log(price) ~ log(income), data = hp, method = "aux")
```

**The two-way test depends on the variance estimator.** Swamy–Arora and
Wallace–Hussain return a negative year variance on Grunfeld that plm
truncates to zero; Amemiya gives 244. The statistic moves from 13.5 to
8.96, and Baltagi's 8.842, computed by EViews with the Wansbeek–Kapteyn
components, is not available in plm [@baltagi2021, p. 99; @wansbeek1989].
And plm returns one two-way statistic, the contrast of two-way GLS with
two-way within, where @kang1985 names five testable hypotheses, each
contrasting a different pair of estimators [@baltagi2021, p. 99]; the other
four have no implementation here.

**Rejection does not validate fixed effects.** It says the random-effects
estimator is inconsistent; whether the fixed-effects restrictions hold is a
separate question, the one Chamberlain's and Angrist–Newey's tests ask
[@baltagi2021, p. 93]. Both impose a set of restrictions that grows with
$T$, and the Chamberlain test understates its variance as $T$ grows
[@baltagi2021, p. 93]: on Grunfeld `piest` stops with "system is
computationally singular" and `aneweytest` returns a negative statistic; on
`Wages` `piest` stops too:

```{r}
#| label: c05-4-pitfall-piest
#| error: true
plm::piest(update(fw1, . ~ . - sex - black - ed), data = wg)
```

**The pretest distorts the second stage.** Choosing fixed or random
effects by a Hausman test and then testing a slope gives a two-stage test
whose size is wrong; @guggenberger2010 recommends fixed-effects inference
without the pretest [@baltagi2021, p. 94].

**Hausman–Taylor's degrees of freedom.** `phtest(ht, within)` reports 9,
the number of common coefficients; Baltagi uses 3, the number of
over-identifying conditions, the count of exogenous time-varying regressors
less the count of endogenous time-invariant ones [@baltagi2021, p. 175].
Same statistic, different p-value. `plm::pht()` is the deprecated route to
the same fit.
:::

::: {.callout-tip collapse="true"}
## Book check

@baltagi2021 [Table 4.4, p. 96] prints the Stata Hausman test for Grunfeld,
2.33 with p = 0.3119: the `c05-4-gr` chunk gives 2.3304, p = 0.3119,
reproduced. On p. 95 the $m_3$ form gives 2.131: the `aux` call gives
2.1314, reproduced. The robust F(2, 195) of 1.58 is not reproduced: plm's
`aux` with `vcovHC` gives $\chi^2_2$ = 8.30 with the Arellano (clustered)
covariance and 3.10 with plm's `white2`:

```{r}
#| label: c05-4-book-robust
plm::phtest(fg, data = gr, method = "aux",
            vcov = function(x) plm::vcovHC(x, method = "white2"))
```

@baltagi2021 [p. 99 and Table 4.6, p. 100] prints the two-way Hausman test
8.842 (p = 0.012). Not reproduced: plm gives 13.46 (`swar`), 8.963
(`amemiya`), 14.15 (`walhus`) and 5.42 (`nerlove`); the Wansbeek–Kapteyn
components EViews used are not among plm's methods.

```{r}
#| label: c05-4-book-twoway
gr_re2w <- plm::plm(fg, data = gr, model = "random", effect = "twoways",
                    random.method = "walhus")
gr_re2n <- plm::plm(fg, data = gr, model = "random", effect = "twoways",
                    random.method = "nerlove")
plm::phtest(gr_w2, gr_re2w)
plm::phtest(gr_w2, gr_re2n)
```

@croissant2019 [Example 5.9, p. 126] print 2.3 and 2.1 for the contrast
and the regression form on Grunfeld: reproduced.

@baltagi2021 [Table 7.6, p. 175] prints the Hausman–Taylor fit of the wage
equation: `ed` 0.137944 (0.0212485), constant 2.912726 (0.2836522),
`sigma_u` 0.9418, `sigma_e` 0.1518. The `c05-4-wages` chunk gives 0.13794
(0.02125), 2.91273 (0.28365), 0.94180 and 0.15180, reproduced. Table 7.5
[p. 174] prints the Amemiya–MaCurdy constant 2.927 and `ed` 0.137:
2.92734 and 0.13720, reproduced. Page 174 gives $\chi^2_9$ = 5075 for GLS
against within: 5075.3, reproduced. Page 175 gives $\chi^2_3$ = 5.26 for
HT against within: 5.2577, reproduced, with plm's df = 9; on 3 df the
p-value is

```{r}
#| label: c05-4-book-ht-df
pchisq(plm::phtest(wg_ht, wg_w)$statistic, df = 3, lower.tail = FALSE)
```

@croissant2019 [pp. 94–95] print `piest` 110 on 87 df (p = 0.03) and
`aneweytest` 140 (p = 2e-04) on the rice farms, 171 farms over six
seasons:

```{r}
#| label: c05-4-book-rice
data("RiceFarms", package = "plm")
rf <- plm::pdata.frame(RiceFarms, index = "id")
frf <- log(goutput) ~ log(seed) + log(totlabor) + log(size)
plm::piest(frf, data = rf)
plm::aneweytest(frf, data = rf)
```

113.72 (p = 0.029) and 141.89 (p = 0.00019): reproduced to the two
significant digits CM print.
:::
