## Heteroskedasticity {#sec-core-hetero}

**Question.** Do big firms' investment shocks have a bigger variance than
small firms', and large states' price shocks than small states'? If so,
which estimates can still be trusted, and which spatial estimator of the
[spatial chapter](../03-spatial/index.qmd#sec-sp-fe) survives?

**Intuition.** Unequal error variances leave the slopes consistent but not
efficient, and the classical standard errors biased, in either direction
[@baltagi2021, p. 109]. Two routes. Keep the estimator and fix the
variance: White's sandwich allows each observation its own variance
[@white1980], and Arellano's panel version allows each unit its own
variance *and* any serial correlation inside the unit [@arellano1987;
@baltagi2021, p. 113]. Or model the variance and reweight: **general
FGLS** estimates the whole $T \times T$ covariance of a unit's errors from
the cross-section, by averaging the outer products of the residuals over
units, and runs GLS with it [@parks1967; @kiefer1980; @wooldridge2010;
@croissant2019, sec. 5.2.1]. The second route is efficient when it is
right, and it is right only when $N$ is large against $T(T + 1)/2$, the
number of covariance parameters it estimates [@croissant2019, p. 128].

::: {.callout-tip appearance="simple"}
Heteroskedastic error components, the Mazodier–Trognon model, and FEGLS
with its rank-$(T - 1)$ caveat: [the error of a panel
regression](theory-panel-error.qmd).
:::

With $\tilde X_i$ and $\tilde u_i$ unit $i$'s within-transformed
regressors and residuals, Arellano's covariance is [@arellano1987;
@croissant2019, p. 121]

$$
\widehat{\operatorname{var}}(\hat\beta_{Within}) =
  (\tilde X'\tilde X)^{-1}
  \Big(\sum_{i=1}^{N} \tilde X_i'\tilde u_i\tilde u_i'\tilde X_i\Big)
  (\tilde X'\tilde X)^{-1},
$$

White's keeps only the diagonal of $\tilde u_i\tilde u_i'$. General FGLS
replaces the middle by a model [@croissant2019, p. 128]:

$$
\hat\Omega_T = \frac{1}{N}\sum_{i=1}^{N} \hat e_i\hat e_i',
\qquad
\hat\beta_{GGLS} = \big(X'(I_N \otimes \hat\Omega_T)^{-1}X\big)^{-1}
  X'(I_N \otimes \hat\Omega_T)^{-1}y .
$$

::: {.callout-important appearance="simple"}
**Limitation.** There is no panel test of homoskedasticity in R. Verbon's
LM test, where both variance components are the same function of
time-invariant unit characteristics, and Lejeune's joint LM tests of no
unit effect and homoskedasticity in the remainder error [@verbon1980;
@lejeune2006; @baltagi2021, sec. 5.1.1, pp. 113–114] have no
implementation here, nor does the modified Breusch–Pagan test for random
effects robust to heteroskedasticity of Li and Stengos [@baltagi2021, p.
113]. This step shows the evidence by picture and by the ordinary
Breusch–Pagan regression of squared residuals on the within-transformed
data, a substitute that is not a panel test and has no published panel
size.
:::

**The tests.** `vcovHC` through `coeftest` with `method = "arellano"`,
the default, and `method = "white1"`; `pggls` with `model = "pooling"`
and `model = "within"` [@croissant2019, sec. 5.2.1]; `phtest` on the two
`pggls` fits, the Hausman test between them [@croissant2019, p. 132].
`lmtest::bptest`, the Breusch–Pagan test of @breusch1980's companion
paper for heteroskedasticity, on an `lm` of the within-transformed
variables; null: constant variance; not the LM test for effects of the
third step.

**Run it, Produc.** The within fit of the public-capital equation, the
spec's canonical case: classical, White, Arellano.

```{r}
#| label: c05-6-pr-se
pr_w <- plm::plm(fp, data = pr, model = "within")
lmtest::coeftest(pr_w)
lmtest::coeftest(pr_w, vcov = function(x) plm::vcovHC(x, method = "white1"))
lmtest::coeftest(pr_w, vcov = plm::vcovHC)
```

Then the reweighted fits, pooled and within, and the Hausman test between
them:

```{r}
#| label: c05-6-pr-ggls
pr_ggls_p <- plm::pggls(fp, data = pr, model = "pooling")
summary(pr_ggls_p)
pr_ggls_w <- plm::pggls(fp, data = pr, model = "within")
summary(pr_ggls_w)
plm::phtest(pr_ggls_w, pr_ggls_p)
```

**Run it, Grunfeld.** The same six calls:

```{r}
#| label: c05-6-gr
lmtest::coeftest(gr_w)
lmtest::coeftest(gr_w, vcov = function(x) plm::vcovHC(x, method = "white1"))
lmtest::coeftest(gr_w, vcov = plm::vcovHC)
gr_ggls_p <- plm::pggls(fg, data = gr, model = "pooling")
coef(summary(gr_ggls_p))
gr_ggls_w <- plm::pggls(fg, data = gr, model = "within")
coef(summary(gr_ggls_w))
plm::phtest(gr_ggls_w, gr_ggls_p)
```

**Run it, house prices.**

```{r}
#| label: c05-6-hp
lmtest::coeftest(hp_w)
lmtest::coeftest(hp_w, vcov = function(x) plm::vcovHC(x, method = "white1"))
lmtest::coeftest(hp_w, vcov = plm::vcovHC)
hp_ggls_p <- plm::pggls(fh, data = hp, model = "pooling")
coef(summary(hp_ggls_p))
hp_ggls_w <- plm::pggls(fh, data = hp, model = "within")
coef(summary(hp_ggls_w))
plm::phtest(hp_ggls_w, hp_ggls_p)
```

The substitute diagnostic, on the within-transformed variables of each
panel:

```{r}
#| label: c05-6-bptest
lmtest::bptest(lm(plm::Within(log(pr$gsp)) ~ plm::Within(log(pr$pcap)) +
  plm::Within(log(pr$pc)) + plm::Within(log(pr$emp)) +
  plm::Within(pr$unemp)))
lmtest::bptest(lm(plm::Within(gr$inv) ~ plm::Within(gr$value) +
  plm::Within(gr$capital)))
lmtest::bptest(lm(plm::Within(hp$lp) ~ plm::Within(hp$li)))
```

The evidence, drawn: each unit's within-residual standard deviation
against its size, the fan; and the correlation matrix the pooled FGLS fit
estimated on Produc, whose off-diagonals do not die out.

```{r}
#| label: c05-6-picture
#| fig-width: 9
#| fig-cap: "Within-residual standard deviation per unit against unit
#|   size, Grunfeld (left, mean market value) and house prices (centre,
#|   mean income, log scale); and the 17 × 17 error correlation matrix
#|   estimated by pooled general FGLS on Produc (right), darker is higher."
#| fig-alt: "Three panels. Left, ten points that rise steeply with size:
#|   the two largest firms have residual standard deviations near 95, the
#|   smallest near 2. Centre, 49 points in a flat band between 0.06 and
#|   0.23 with no trend in income. Right, a grey-scale square that is
#|   darkest on the diagonal and stays dark for most of each row, fading
#|   only in the last columns."
unit_sd <- function(fit, size, main, xlab, log = "") {
  id <- plm::index(fit)[[1]]
  s <- tapply(as.numeric(residuals(fit)), id, sd)
  z <- tapply(as.numeric(size), id, mean)
  plot(z, s, pch = 16, log = log, main = main, xlab = xlab,
       ylab = "residual sd")
}
op <- graphics::par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
unit_sd(gr_w, gr$value, "Grunfeld", "mean market value")
unit_sd(hp_w, hp$income, "House prices", "mean income", log = "x")
r_pr <- cov2cor(pr_ggls_p$sigma)
graphics::image(1:17, 1:17, r_pr[, 17:1], col = grey(seq(1, 0, by = -0.05)),
                xlab = "year", ylab = "year", main = "Produc: pooled FGLS",
                yaxt = "n")
yrs <- levels(plm::index(pr)[[2]])
graphics::axis(2, at = c(1, 17), labels = c(yrs[17], yrs[1]))
graphics::par(op)
```

**Read it.**

- **Produc.** Arellano's errors are 2.1 to 2.7 times the classical ones;
  the within FGLS fit moves `log(pc)` from 0.29 to 0.17 and `log(emp)`
  from 0.77 to 0.84, and the Hausman test between the FGLS fits (32.1)
  keeps the within version. *So:* the classical errors overstate
  precision by more than half, and reweighting changes the point
  estimates, which is what inefficiency under heteroskedasticity and
  serial correlation looks like; the Breusch–Pagan substitute (18.7, p =
  0.0009) agrees.
- **Grunfeld.** The fan is steep: the two largest firms have residual
  standard deviations near 95, the smallest 1.5, and Arellano's errors
  are 1.2 (value) and 2.9 (capital) times the classical ones. *So:* the
  shocks scale with the firm; and `pggls` on Grunfeld, 210 covariance
  parameters from ten firms, is shown and read as unreliable, its
  Hausman test of 816 included.
- **House prices.** The fan is flat, the correlation of residual
  standard deviation with income 0.18 on the log scale; yet the
  substitute rejects (25.7) and Arellano's error is 3.3 times the
  classical. *So:* heteroskedasticity is there but not by size, and most
  of Arellano's inflation is the serial correlation of the previous step.
  The within FGLS slope is 0.316 against within's 0.345, the pooled FGLS
  slope 0.378.
- **Misleads (White's `"white1"` on Produc).** 1.1 to 1.3 times the
  classical errors. *So:* it allows the variance to differ but not the
  serial correlation, and understates the problem by half; on a panel the
  default of `vcovHC.plm` is Arellano for that reason [@croissant2019, p.
  113].

::: {.callout-note appearance="simple"}
**Next.** The menu of robust errors and which one fits which dependence
([robust errors](index.qmd#sec-core-robust-se)). For the spatial fits:
heteroskedasticity present, keep the GM fit, because spatial maximum
likelihood under homoskedastic innovations is inconsistent when the
innovations are not [@lin2010; @pesaran2015, p. 807] ([fixed
effects](../03-spatial/index.qmd#sec-sp-fe)).
:::

::: {.callout-note collapse="true"}
## Recipe

```r
# w: a within plm fit; f: its formula; d: a pdata.frame.
lmtest::coeftest(w, vcov = plm::vcovHC)                   # Arellano
lmtest::coeftest(w, vcov = function(x) plm::vcovHC(x, method = "white1"))
plm::pggls(f, data = d, model = "within")                 # FEGLS, N >> T
plm::phtest(plm::pggls(f, data = d, model = "within"),
            plm::pggls(f, data = d, model = "pooling"))
```
:::

::: {.callout-caution collapse="true"}
## Pitfalls

**Two `vcovHC`s.** `plm::vcovHC` on a panel fit defaults to Arellano,
clustered by unit; `sandwich::vcovHC` on an `lm` of the same data defaults
to White. The same call name gives different numbers, and only
`method = "white1"` with the `HC3` correction makes the `lm` numbers
reappear [@croissant2019, Example 5.1, p. 113].

**FGLS needs $N \gg T$.** The number of covariance parameters is
$T(T + 1)/2$; with few units the estimate is near-singular and the
standard errors biased downward [@croissant2019, p. 128]. Grunfeld's 210
from ten firms is not a model. The within version drops one period,
because the covariance of demeaned errors has rank $T - 1$
[@croissant2019, p. 131].

**`pggls(model = "fd")` carries an intercept.** In plm 2.6.7 the
first-difference model has one by default; CM's printed fit has none, and
the Book check shows both.

**Heteroskedastic errors break spatial ML.** The maximum-likelihood
spatial estimators of the spatial chapter assume homoskedastic
innovations and are inconsistent without them [@lin2010; @pesaran2015,
p. 807]; the GM fit survives.
:::

::: {.callout-tip collapse="true"}
## Book check

@croissant2019 [Example 5.1, p. 112] print `coeftest(plmmod, vcov =
vcovHC)` on pooled Produc with standard errors 0.06012, 0.04623, 0.06861,
0.00309 on the four slopes:

```{r}
#| label: c05-6-book-hc
pr_pool <- plm::plm(fp, data = pr, model = "pooling")
lmtest::coeftest(pr_pool, vcov = plm::vcovHC)
```

0.06012, 0.04623, 0.06861, 0.00309: reproduced.

@croissant2019 [Examples 5.11 to 5.14, pp. 129–132] print, for `EmplUK`,
pooled general FGLS 2.0235, −0.2323, 0.6105; first-difference FGLS
−0.3343 (0.0385), 0.3786 (0.0203); and the Hausman test between the within
and pooled FGLS fits, 1100:

```{r}
#| label: c05-6-book-ggls
data("EmplUK", package = "plm")
em <- plm::pdata.frame(EmplUK)
fe_em <- log(emp) ~ log(wage) + log(capital)
em_ggls_p <- plm::pggls(fe_em, data = em, model = "pooling")
coef(summary(em_ggls_p))
em_ggls_w <- plm::pggls(fe_em, data = em, model = "within")
plm::phtest(em_ggls_w, em_ggls_p)
em_ggls_fd <- plm::pggls(fe_em, data = em, model = "fd")
coef(summary(em_ggls_fd))
em_ggls_fd1 <- plm::pggls(update(fe_em, . ~ . - 1), data = em, model = "fd")
coef(summary(em_ggls_fd1))
```

Pooled 2.0235, −0.2323, 0.6105: reproduced. Hausman 1064.6 against 1100:
reproduced to CM's two digits. First differences: −0.3129 and 0.3712 with
plm's default intercept, not CM's numbers; −0.3343 (0.0385) and 0.3786
(0.0203) with `- 1` in the formula, reproduced. @baltagi2021 [sec. 5.1]
prints no number for this step.
:::
