---
title: "Core panel models"
---

In 1935 the largest of Grunfeld's ten firms had a market value
forty-three times that of the smallest, and invested a hundred and
twenty-five times as much. @grunfeld1958 followed the ten for twenty years
to ask one question: does a firm invest more when the market values it
more, and when it already owns more capital? The pooled answer is one line
through two hundred points. But the largest firm is not the smallest on a
larger scale. Each firm has a character of its own, management, product,
luck, that the regressors do not see and that moves every one of its twenty
observations the same way.

A **panel regression with unit effects** gives each unit its own intercept
and lets the slopes be shared. The intercept is the unit's character, and
the whole chapter turns on what to assume about it. Under **fixed effects**
it is a parameter, estimated from the unit's own variation over time, and
it may be correlated with the regressors. Under **random effects** it is a
draw from a distribution, uncorrelated with the regressors, and it stays in
the error term, where it makes a unit's errors correlated across years. Why
the two differ, and when they agree: [fixed and random
effects](theory-fixed-random.qmd); what the error of a panel regression
carries, and what to do about it: [the error of a panel
regression](theory-panel-error.qmd).

```{r}
#| label: c05-0-setup
devtools::load_all(quiet = TRUE)
# pooltest()'s formula method builds a call to plm(), so plm must be
# attached, not just namespaced.
library(plm)
data("Grunfeld", package = "plm")
data("Produc", package = "plm")
data("Wages", package = "plm")
data("HousePricesUS", package = "pder")
gr <- plm::pdata.frame(Grunfeld)
pr <- plm::pdata.frame(Produc)
wg <- plm::pdata.frame(Wages, index = 595)
hp <- plm::pdata.frame(HousePricesUS)
# phtest(method = "aux") cannot parse log() in a formula, so the logs are
# columns.
hp$lp <- log(hp$price)
hp$li <- log(hp$income)
fg <- inv ~ value + capital
fp <- log(gsp) ~ log(pcap) + log(pc) + log(emp) + unemp
fh <- lp ~ li
```

```{r}
#| label: c05-0-picture
#| fig-cap: "Investment against market value for Grunfeld's ten firms,
#|   1935–54. The thick line is pooled OLS; the thin lines are the
#|   fixed-effects fit, one slope and ten firm intercepts."
#| fig-alt: "A scatterplot of investment against firm value with two
#|   hundred points, most crowded near the origin and a few far out to the
#|   right. One thick line runs through the whole cloud. Ten thin parallel
#|   lines, each through the points of one firm, sit at different heights,
#|   the low firms' lines near the origin and the big firms' lines higher
#|   and further right."
gr_pool1 <- plm::plm(inv ~ value, data = gr, model = "pooling")
gr_w1 <- plm::plm(inv ~ value, data = gr, model = "within")
firm <- plm::index(gr)[[1]]
plot(as.numeric(gr$value), as.numeric(gr$inv), pch = 16,
     col = grey(0.2 + 0.6 * (as.integer(firm) - 1) / 9),
     xlab = "market value", ylab = "investment")
graphics::abline(coef(gr_pool1), lwd = 3)
for (a in plm::fixef(gr_w1)) graphics::abline(a, coef(gr_w1), lty = 3)
```

The thin lines are the one-way **error-component model**, the equation
that organises the chapter [@baltagi2021, eqs. 2.1–2.2]:

$$
y_{it} = x_{it}'\beta + \mu_i + \nu_{it},
\qquad i = 1, \dots, N;\; t = 1, \dots, T,
$$

where $\mu_i$ is the unit effect and $\nu_{it}$ the idiosyncratic error.
Fixed effects estimates $\beta$ from deviations around each unit's own mean,
so $\mu_i$ drops out. Random effects treats $\mu_i + \nu_{it}$ as one error
with a known shape and weights the within and between variation by how much
each is worth. The chapter's tests ask, in order, whether the slopes can be
shared at all, whether $\mu_i$ is there, whether it is a parameter or a
draw, what else $\nu_{it}$ carries, and which standard error is honest about
it.

**In this chapter.**

1. **Poolability.** Can one line serve every unit, or every unit with its
   own intercept ([poolability](index.qmd#sec-core-pool)).
2. **Slope heterogeneity.** How far apart the units' slopes are, and
   whether the average is still worth reporting
   ([slope heterogeneity](index.qmd#sec-core-slopes)).
3. **Effects tests.** Whether a unit effect, a year effect, or both are
   there at all ([effects tests](index.qmd#sec-core-effects)).
4. **Fixed or random.** Whether the unit effect moves with the regressors,
   and what to do when a regressor never moves over time
   ([fixed or random](index.qmd#sec-core-fe-re)).
5. **Serial correlation.** Whether this year's shock carries into next
   year's, and whether fixed effects or first differences is the better
   estimator ([serial correlation](index.qmd#sec-core-serial)).
6. **Heteroskedasticity.** Whether big units have bigger shocks, and which
   estimates survive it ([heteroskedasticity](index.qmd#sec-core-hetero)).
7. **Robust errors.** The menu of standard errors, and which dependence
   each one survives ([robust errors](index.qmd#sec-core-robust-se)).
8. **Summary** ([summary](index.qmd#sec-core-summary)).

Two panels run through every step. Grunfeld's investment panel carries the
book checks: @baltagi2021 prints Tables 4.2, 4.3, 4.4, 4.6 and 5.3 and the
poolability example of section 4.1.3 on it, so the book's numbers come out
of the calls in the text. House prices carry the story forward:
`HousePricesUS` [@holly2010], 49 states over 29 years, whose income
elasticity the [cross-dependence
chapter](../02-cross-dependence/index.qmd#sec-cce) estimated state by state
and averaged, and whose errors left the [spatial
chapter](../03-spatial/index.qmd#sec-sp-serial) with a serial correlation of
0.82. The public-capital panel `Produc` [@munnell1990], 48 states over 17
years, joins for the heteroskedasticity and robust-error steps, where
@croissant2019 print their tables on it; `Wages`, 595 workers over 7 years,
appears once, for the Hausman–Taylor estimator.

Every section has the shape of the [spatial
chapter](../03-spatial/index.qmd): **question**, **intuition**, the
**test** in words, **run it**, a picture, **read it**, a **next** box, and
collapsed **recipe**, **pitfalls** and **book check** boxes at the end.

{{< include 1-poolability.qmd >}}

{{< include 2-slopes.qmd >}}

{{< include 3-effects.qmd >}}

{{< include 4-fe-or-re.qmd >}}

{{< include 5-serial.qmd >}}

{{< include 6-hetero.qmd >}}

{{< include 7-robust-se.qmd >}}

{{< include 8-summary.qmd >}}
