### 0.1 Panel dimensions, balance, variation {#sec-panel-shape}

**Theory.** The shape of the panel decides which methods are usable. *Micro*
panels have large N (hundreds or thousands) and short T (rarely over 10–20),
so asymptotics are large N, fixed T. *Macro* panels have moderate N (7 to 200
countries) over 20–60 years, so asymptotics can be large N and T
[@baltagi2021, p. 1]. With N large (say over 100) and T small (under 50),
panel unit-root and cointegration analysis "can proceed only under restrictive
assumptions such as dynamic homogeneity and/or local cross-sectional
dependence as in spatial autoregressive or moving average models"
[@pesaran2015, p. 817].

A panel is *unbalanced* when units are observed over different periods. In the
one-way model the Within estimator carries over unchanged, while GLS becomes
weighted least squares with weights that vary with each unit's Tᵢ, and
variance components lose their optimal estimators [@baltagi2021, pp. 230–232].

**Data.** `TobinQ` (balanced) and `Tileries` (unbalanced), both in **pder**
[@croissant2019].

```{r}
#| label: s0-1-pdim
library(plm)
data("TobinQ", package = "pder")
data("Tileries", package = "pder")
pdim(TobinQ)
pdim(Tileries)
```

How unbalanced? Both measures equal 1 for a balanced panel and fall toward 0
as balance worsens [@ahrens1981]:

```{r}
#| label: s0-1-punbalancedness
punbalancedness(Tileries)
```

**Book check.** @croissant2019 prints `n = 188, T = 35, N = 6580` for TobinQ
[p. 32] and `n = 25, T = 12-22, N = 483` for Tileries [p. 62]. The book prints
no `punbalancedness` output.
