### State the alternative first {#sec-ur-alternative}

::: {.callout-tip appearance="simple"}
One null, three alternatives, and why the p-value alone does not say which one
you rejected in favour of:
[theory note 07](theory/07-panel-unit-roots.qmd).
:::

**Theory.** Every panel unit-root test in this stage shares the same null —
$\rho_i = 1$ for every *i*. They do not share an alternative, and the
alternative is the whole content of a rejection [@pesaran2015, pp. 818–820]:

| | Alternative | Tests that assume it |
|---|---|---|
| H₁ₐ | $\rho_i = \rho < 1$, the same root for every unit | LLC, Breitung, Harris–Tzavalis |
| H₁b | $\rho_i < 1$ for $N_1$ units, with $N_1/N \to \kappa > 0$ | IPS, Fisher-type |
| H₁c | $\rho_i < 1$ for a *fixed number* of units, so $\kappa = 0$ | none |

The difference between H₁b and H₁c is not a technicality. Both say "some series
are stationary"; only H₁b says the stationary units are a *share* of the panel.
A test built on averaging over *N* — and all of them average over *N* — has
power against a fixed share and loses it against a fixed count, because the
signal from $N_1$ units is diluted by the $N - N_1$ that carry none.

**Data.** Simulated, because the point is about the data-generating process and
no real panel tells you its own $\kappa$. Each series is an AR(1) with
$\rho_i \in \{0.5, 1\}$ and the count of stationary ones is set by hand:

```{r}
#| label: s1-1-dgp
ar1 <- function(t, rho, burn = 50) {
  y <- Reduce(function(a, e) rho * a + e, rnorm(t + burn),
              accumulate = TRUE)
  y[-seq_len(burn)]
}
panel <- function(n, t, k, rho = 0.5) {
  sapply(seq_len(n), function(i) ar1(t, if (i <= k) rho else 1))
}
```

**Power against a share, and against a count.** Rejection frequency of IPS at a
nominal 5%, over 200 panels of T = 30, as *N* grows. Column `none` is the size
of the test (no stationary series at all), `two` holds the *count* fixed at two
series, `share_40` holds the *fraction* fixed at 40%:

```{r}
#| label: s1-1-power
reject <- function(n, k, reps = 200, t = 30) {
  p <- replicate(reps, plm::purtest(panel(n, t, k), test = "ips",
                                    exo = "intercept",
                                    lags = 0)$statistic$p.value)
  mean(p < 0.05)
}
set.seed(1)
ns <- c(10, 30, 90)
data.frame(
  N = ns,
  none = sapply(ns, function(n) reject(n, k = 0)),
  two = sapply(ns, function(n) reject(n, k = 2)),
  share_40 = sapply(ns, function(n) reject(n, k = 0.4 * n))
)
```

Read the three columns against each other, never the middle one alone. Holding
the *share* at 40%, power goes 0.92 → 1.00 → 1.00: more units is more evidence.
Holding the *count* at two, it goes 0.40 → 0.30 → 0.25 *while the size of the
test rises from 0.085 to 0.150* — so at N = 90 the test is barely ten points
above its own false-positive rate. That is H₁c: the two stationary series are
still there, and the test has stopped being able to see them.

The size column is the other half of the lesson. These series are drawn
independently, with no cross-sectional dependence of any kind, and IPS still
rejects a true null 8–15% of the time at a nominal 5% because T = 30 is short.
Size comes before power, and it comes back in
[size under spatial dependence](panel-spat.qmd#sec-ur-spatial-size).

**Book check.** The spec records no numerical target for this section —
$H_{1a}/H_{1b}/H_{1c}$ is a taxonomy, not a statistic. What *can* be checked is
that the software does not carry the distinction. LLC assumes H₁ₐ and IPS
assumes H₁b, so the two rejections mean different things; both print the same
word:

```{r}
#| label: s1-1-book-check
set.seed(1)
m <- panel(n = 20, t = 30, k = 8)
llc <- plm::purtest(m, test = "levinlin", exo = "intercept", lags = 0)
ips <- plm::purtest(m, test = "ips", exo = "intercept", lags = 0)
c(llc = llc$statistic$alternative, ips = ips$statistic$alternative)
```

`"stationarity"` for both, which is true of neither: LLC rejects towards one
common stationary root, IPS towards a positive fraction of stationary roots.
Whichever test is run, the alternative has to be written into the text by hand.

Remark 8 of @pesaran2015 [p. 820] is the constraint on that text: at T ≈ 15,
panel unit-root results are informative only "in an average sense". That is a
limit on what may be claimed, not on what may be computed.
