### First-generation tests {#sec-ur-first-gen}

::: {.callout-tip appearance="simple"}
"First generation" is an assumption about cross-sectional dependence, not a
date: [theory note 07](theory/07-panel-unit-roots.qmd).
:::

**Theory.** Six tests, one null, and the family split of
[the alternatives](panel-spat.qmd#sec-ur-alternative). Every one of them
assumes the series are cross-sectionally independent once a common time effect
has been taken out:

| Test | Source | Statistic | H₁ | In `purtest` |
|---|---|---|---|---|
| Levin–Lin–Chu | [@levin2002; @baltagi2021, pp. 340–343] | pooled ADF *t*, bias-corrected | H₁ₐ | `"levinlin"` |
| Harris–Tzavalis | [@harris1999; @baltagi2021, p. 343] | pooled ρ̂ at fixed *T* | H₁ₐ | ✖ |
| Breitung | [@breitung2000; @baltagi2021, pp. 345–346] | pooled, bias-free by construction | H₁ₐ | ✖ |
| Im–Pesaran–Shin | [@im2003; @baltagi2021, pp. 344–345] | mean of individual ADF *t* | H₁b | `"ips"` |
| Fisher | [@maddala1999; @choi2001; @baltagi2021, pp. 346–348] | combines individual p-values | H₁b | `"madwu"`, `"Pm"`, `"invnormal"`, `"logit"` |
| Hadri | [@hadri2000; @baltagi2021, pp. 348–349] | LM; the **null is stationarity** | reversed | `"hadri"` |

The comparative evidence favours the two that are missing. @pesaran2015 [p. 819]
puts LLC's local power above IPS's, and [p. 838] reports that LLC and Breitung
carry the smallest size distortions, with Fisher at least as good as IPS — the
same ranking @baltagi2021 [p. 348] gives. Of the three tests that come out
best, **plm** implements one.

::: {.callout-important appearance="simple"}
**Gap.** Breitung and Harris–Tzavalis are not in `purtest` and have no verified
R implementation. If either is load-bearing, the benchmark is Stata's
`xtunitroot breitung` / `xtunitroot ht`, and the result has to be labelled as
coming from outside this book.
:::

**Data.** Grunfeld investment [@grunfeld1958], the panel `?purtest` runs its own
examples on, and `HousePricesUS` [@holly2010] for the book's run.

**One test, three interfaces.** `purtest` accepts a wide matrix, a formula with
`data` and `index`, or a `pseries`. The help page's own consistency claim is
that all three give the same number:

```{r}
#| label: s1-2-interfaces
data("Grunfeld", package = "plm")
wide <- data.frame(split(Grunfeld$inv, Grunfeld$firm))
pG <- plm::pdata.frame(Grunfeld, index = c("firm", "year"))
mw <- function(x, ...) {
  unname(plm::purtest(x, test = "madwu", pmax = 4, ...)$statistic$statistic)
}
c(matrix = mw(wide, exo = "intercept"),
  formula = mw(inv ~ 1, data = Grunfeld, index = "firm"),
  pseries = mw(pG$inv, exo = "intercept"))
```

**The interface trap.** The three are not interchangeable in how they take the
deterministic part. With a formula, the deterministics are the formula's right
side and `exo` is silently ignored; with a matrix or a `pseries`, `exo` is the
only thing that sets them:

```{r}
#| label: s1-2-exo-trap
c(formula_1_exo_trend = mw(inv ~ 1, data = Grunfeld, index = "firm",
                           exo = "trend"),
  formula_trend = mw(inv ~ trend, data = Grunfeld, index = "firm"),
  pseries_exo_trend = mw(pG$inv, exo = "trend"))
```

The first call asks for a trend and gets individual intercepts. Nothing warns;
the statistic just changes by a factor of five. Pick one interface per project
and keep to it.

**What refuses to run.** LLC and Hadri need a balanced panel; the Fisher and
IPS statistics do not, and run on the unbalanced `EmplUK` [@arellano1991]:

```{r}
#| label: s1-2-unbalanced
data("EmplUK", package = "plm")
pE <- plm::pdata.frame(EmplUK, index = c("firm", "year"))
unbalanced <- function(test) {
  tryCatch({
    plm::purtest(log(pE$emp), test = test, exo = "intercept", lags = 0)
    "runs"
  }, error = conditionMessage)
}
sapply(c("levinlin", "ips", "madwu", "hadri"), unbalanced)
```

IPS has a second statistic, `tbar`, which is exact for fixed *T* — and which is
unavailable with `lags > 0` and returns tabulated critical values instead of a
p-value:

```{r}
#| label: s1-2-tbar
plm::purtest(pG$inv, exo = "intercept", test = "ips", ips.stat = "tbar",
             lags = 0)
```

**Where the p-values come from.** A Fisher test is a combination of per-series
p-values and nothing more, and `purtest` keeps its working: `$idres` holds each
series' ADF *t* and the p-value read off the MacKinnon (1996) surface
[@mackinnon1996] by `urca::punitroot`. Maddala–Wu then combines them as
$\chi^2 = -2\sum_i \log p_i$ on 2N degrees of freedom:

```{r}
#| label: s1-2-fisher
mw_G <- plm::purtest(wide, test = "madwu", pmax = 4, exo = "intercept")
t_i <- sapply(mw_G$idres, function(x) unname(x$trho))
p_i <- sapply(mw_G$idres, function(x) unname(x$p.trho))
round(c(plm_p1 = p_i[[1]],
        urca_p1 = urca::punitroot(t_i[[1]], N = Inf, trend = "c"),
        fisher = -2 * sum(log(p_i)),
        printed = unname(mw_G$statistic$statistic),
        df = unname(mw_G$statistic$parameter)), 5)
```

Those same per-series p-values are the input
[the fraction of stationary series](panel-spat.qmd#sec-ur-fraction) works on.

Without **urca** installed, `padf` falls back to the MacKinnon (1994)
approximation instead. That is why **urca** is in the project renv, though on
this series the choice moves the fourth digit and nothing else:

```{r}
#| label: s1-2-padf
data("HousePricesUS", package = "pder")
php <- plm::pdata.frame(HousePricesUS)
mw_hp <- function(...) {
  unname(plm::purtest(log(php$price), test = "madwu", lags = 2,
                      exo = "trend", ...)$statistic$statistic)
}
c(MacKinnon1996 = mw_hp(), MacKinnon1994 = mw_hp(p.approx = "MacKinnon1994"))
```

**Book check.** @croissant2019 [§8.4.2] runs LLC, Maddala–Wu and IPS on
`HousePricesUS` log price with `lags = 2` and `exo = "trend"`;
`.docs/panel-spat-PLAN.md` records that run as this section's target. Hadri is
added here because its reversed null is the cheapest corroboration available:

```{r}
#| label: s1-2-book-check
first_gen <- function(test) {
  r <- plm::purtest(log(php$price), test = test, lags = 2, exo = "trend")
  c(statistic = unname(r$statistic$statistic),
    p.value = unname(r$statistic$p.value))
}
round(rbind(LLC = first_gen("levinlin"), MW = first_gen("madwu"),
            IPS = first_gen("ips"), Hadri = first_gen("hadri")), 4)
```

Three tests of "unit root in every series" fail to reject (p = 0.10, 0.43,
0.78), and the one test of "stationarity in every series" rejects flat
(p ≈ 0). All four point the same way: log house prices are I(1), which is the
reading @croissant2019 [§8.4.3] carries into the second-generation tests of
[the next section](panel-spat.qmd#sec-ur-second-gen).

The stronger check is the internal one above — the same Maddala–Wu statistic,
14.719, from three different interfaces on Grunfeld. That is the `?purtest`
example, and package examples are re-run on every CRAN check, so a silent
change in any one interface would have to break the others to survive.

And none of it is worth much yet: every test in this section assumes
cross-sectional independence, which
[testing for cross-sectional dependence](panel-spat.qmd#sec-csd-tests) rejects
on this very panel at z = 53. The numbers above are the ones to beat, not the
ones to report.
