### Second-generation tests {#sec-ur-second-gen}

::: {.callout-tip appearance="simple"}
What the cross-section averages are doing inside the ADF regression:
[theory note 07](theory/07-panel-unit-roots.qmd), and
[theory note 01](theory/01-cross-sectional-dependence.qmd) for why the
first-generation assumption fails here.
:::

**Theory.** The tests of
[the previous section](panel-spat.qmd#sec-ur-first-gen) assume the series are
independent across units. On `HousePricesUS`
[that assumption is rejected](panel-spat.qmd#sec-csd-tests) at z = 53, so the
whole battery has to be redone under dependence.

Pesaran's CADF regression adds the cross-section averages $\bar y_{t-1}$ and
$\Delta \bar y_t$ to each unit's ADF equation, which sweeps out a single common
factor whatever its loadings, and **CIPS** is the average of the resulting *t*
statistics [@pesaran2007; @baltagi2021, pp. 352–353]:

$$
CIPS = \frac{1}{N} \sum_{i=1}^{N} t_i(N, T).
$$

One factor. With several, the averages have to be augmented further, which is
@pesaran2013's CIPS and its companion CSB statistic [@pesaran2015, p. 836].
The other route defactors first and tests afterwards: Moon–Perron
[@moon2004; @baltagi2021, pp. 350–351], Bai–Ng's PANIC
[@bai2004; @baltagi2021, p. 352] and Phillips–Sul
[@phillips2003; @baltagi2021, pp. 351–352].

::: {.callout-important appearance="simple"}
**Gap.** Of that list, only CIPS is implemented. Moon–Perron, Bai–Ng, CSB and
Phillips–Sul have no R implementation here and no verification target, so a
multi-factor error structure cannot be tested in this book — only assumed away.
:::

**Data.** `HousePricesUS` [@holly2010] again, in levels, in first differences,
and as the residuals of the two CCE models of
[the CCE section](panel-spat.qmd#sec-cce).

```{r}
#| label: s1-3-setup
# pcce() builds a call to plm(), so plm must be attached, not just namespaced.
library(plm)
data("HousePricesUS", package = "pder")
php <- plm::pdata.frame(HousePricesUS)
ccemg <- plm::pcce(log(price) ~ log(income), data = HousePricesUS,
                   model = "mg")
ccep <- plm::pcce(log(price) ~ log(income), data = HousePricesUS,
                  model = "p")
cips <- function(x, type) {
  r <- plm::cipstest(x, type = type)
  c(statistic = unname(r$statistic), p.value = unname(r$p.value))
}
```

`cipstest` takes a `pseries` and nothing else — no formula interface, no
matrix. The deterministic part is `type`, and it is the one choice that has to
be made by hand: levels of a price series get a drift, first differences get
neither drift nor trend, and regression residuals get neither.

**Book check.** @croissant2019 [§8.4.3] runs exactly these four tests;
`.docs/panel-spat-PLAN.md` records its printed values as the target — CIPS
about −2 with p ≈ 0.1 in levels, about −1.8 with p ≈ 0.01 on the differenced
series, and rejections for both sets of CCE residuals:

```{r}
#| label: s1-3-book-check
round(rbind(
  level_drift = cips(log(php$price), "drift"),
  diff_none = cips(diff(log(php$price)), "none"),
  ccemg_residuals = cips(residuals(ccemg), "none"),
  ccep_residuals = cips(residuals(ccep), "none")
), 4)
```

−2.0342 against a target of about −2, −1.8199 against about −1.8, and both
residual tests rejecting. Four numbers, four matches.

**Reading the p-value.** `cipstest` has no exact p-values: the CIPS
distribution is non-standard, so **plm** returns the nearest tabulated level
and warns. `p.value = 0.10` above means *greater than* 0.10 and
`p.value = 0.01` means *less than* 0.01 — the warnings that say so are
suppressed in this book's chunk options, so the qualification has to travel in
the text. Never quote a CIPS p-value as if it were exact.

What the four rows say: log house prices carry a unit root that survives
controlling for a common factor (levels, not rejected); their growth rates do
not (differences, rejected); and the residuals of both CCE regressions are
stationary, so the levels regression of log price on log income is a
cointegrating one. That last row is the pragmatic substitute for a panel
cointegration test that Stage 2 of `.docs/panel-spat-PLAN.md` falls back on,
and it is verifiable, which nothing in Stage 2 is.

One caveat carries into the spatial stages. CIPS defactors, and defactoring is
the right move against *strong* dependence only. Under weak, spatially
generated dependence the factor correction is not just unnecessary but invalid
[@pesaran2015, p. 839] — which is the subject of
[size under spatial dependence](panel-spat.qmd#sec-ur-spatial-size).
