### The fraction of stationary series {#sec-ur-fraction}

::: {.callout-tip appearance="simple"}
Why a panel rejection is not a licence to pool:
[theory note 07](theory/07-panel-unit-roots.qmd).
:::

**Theory.** Every test so far answers one question — "do *all* series have a
unit root?" — and stops. Remark 8 of @pesaran2015 [p. 820] is the consequence:
at short *T* the answer is informative only in an average sense. @pesaran2015
[p. 832] gives the way out, and it is a multiple-testing device rather than a
new statistic.

Take the per-series p-values that a Fisher-type test already computes. Simes'
inequality [@simes1986] turns them into a global test of the intersection null
that stays valid under general patterns of cross-sectional dependence, and
Hommel's procedure [@hommel1988] controls the family-wise error rate across the
*N* individual hypotheses, so each series can be judged on its own.
@hanck2013 puts the two together for panel unit roots: instead of "some series
are stationary", a named list of which ones.

**Data.** `HousePricesUS` [@holly2010], through the Maddala–Wu object of
[the first-generation section](panel-spat.qmd#sec-ur-first-gen), and the
19 countries of @hanck2013 [Table 11] for the check.

```{r}
#| label: s1-4-hpus
data("HousePricesUS", package = "pder")
php <- plm::pdata.frame(HousePricesUS)
mw <- plm::purtest(log(php$price), test = "madwu", lags = 2, exo = "trend")
hp <- plm::phansitest(mw)
c(states = length(hp$p), rejected = hp$rejected.no)
```

Not one of the 49 states rejects its own unit root once the multiplicity of
testing 49 hypotheses is controlled for. The global tests agree, but not by the
same margin:

```{r}
#| label: s1-4-global
simes <- function(p) min(length(p) * sort(p) / seq_along(p))
round(c(simes = simes(hp$p), fisher = unname(mw$statistic$p.value)), 4)
```

0.12 against 0.43. Simes is much the sharper of the two here, and it is also
the one that survives the cross-sectional dependence this panel has, so it is
the number to quote — and it still does not reject at 5%.

**The evaluation line is wrong.** `phansitest` returns the individual verdicts
but never computes a global decision, and its print method says the same thing
either way. On the object above, where no individual null is rejected and
Simes' global p-value is 0.12:

```{r}
#| label: s1-4-print-trap
hp
```

"H0 rejected (globally)" is printed unconditionally in **plm** 2.6.7: the
`H0.not.rej.txt` string that belongs in that branch is defined in
`plm:::print.phansitest` and never used. Read `$rejected.no` and compute the
global p-value as above; do not read the sentence.

One input constraint: the test needs a panel test that produces a p-value per
series, so `madwu`, `Pm`, `invnormal`, `logit` and `ips` are usable and Hadri,
whose statistic is not built from individual p-values, is refused by name:

```{r}
#| label: s1-4-hadri
hadri <- plm::purtest(log(php$price), test = "hadri", lags = 2, exo = "trend")
tryCatch(plm::phansitest(hadri), error = conditionMessage)
```

**Book check.** The p-values of @hanck2013 [Table 11], for 19 countries' real
exchange rates, as `?phansitest` transcribes them. Hommel's procedure should
reject the individual unit-root null for the first ten and no others:

```{r}
#| label: s1-4-book-check
pvals <- c(0.0001, 0.0001, 0.0001, 0.0001, 0.0001, 0.0001, 0.0050, 0.0050,
           0.0050, 0.0050, 0.0175, 0.0175, 0.0200, 0.0250, 0.0400, 0.0500,
           0.0575, 0.2375, 0.2475)
names(pvals) <- c("Argentina", "Sweden", "Norway", "Mexico", "Italy",
                  "Finland", "France", "Germany", "Belgium", "U.K.", "Brazil",
                  "Australia", "Netherlands", "Portugal", "Canada", "Spain",
                  "Denmark", "Switzerland", "Japan")
h <- plm::phansitest(pvals)
c(rejected = h$rejected.no, of = length(pvals))
c(simes_global_p = round(simes(pvals), 5))
h$name[h$rejected]
```

Ten of nineteen, and the ten are the ten with the smallest p-values, which is
the split @hanck2013 reports. Note what the second line adds: Simes' global
p-value is 0.0003, so this panel rejects the intersection null *and* names the
units — the two answers Remark 8 asks for. Spain, at an individual p-value of
0.05, is not among them; testing 19 hypotheses at once is what costs it.
