### Size under spatial dependence {#sec-ur-spatial-size}

::: {.callout-tip appearance="simple"}
What W is and what row standardisation does to it:
[theory note 02](theory/02-spatial-weights.qmd). Why size comes before power:
[theory note 07](theory/07-panel-unit-roots.qmd).
:::

**Theory.** @baltagi2007 run 1600 Monte Carlo experiments on panel unit-root
tests under spatial error dependence. The distortions are largest under a
spatial autoregressive error and grow with the density of W
[@baltagi2021, pp. 353–354]. @pesaran2015 [p. 839] adds the two rules that
matter for a choice of test: Pesaran's CD-based approach is the most robust to
spatial alternatives, and factor-based methods are *invalid* under weak
dependence — they defactor something that is not a factor.

::: {.callout-important appearance="simple"}
**No target.** `.docs/panel-spat-PLAN.md` records this element as having no
canned verification: reproducing @baltagi2007's size tables needs their exact
design, and the deliverable instead is an empirical size table for this
project's own W, read against nominal 5%. Everything below is self-produced,
so it is evidence about *this* code on *this* W and nothing more.
:::

**Data.** Simulated. W is rook contiguity on a 7 × 7 grid — 49 units, the shape
of `HousePricesUS` — row-standardised. Under the null every series is a random
walk; the innovations are spatially correlated through
$(I - \lambda W)^{-1}$, so λ = 0 is independence and λ = 0.8 is strong spatial
dependence:

```{r}
#| label: s1-5-dgp
rook_w <- function(side) {
  cells <- expand.grid(row = seq_len(side), col = seq_len(side))
  neighbour <- as.matrix(dist(cells, method = "manhattan")) == 1
  neighbour / rowSums(neighbour)
}
W <- rook_w(7)
rw_panel <- function(lambda, t = 30) {
  spill <- solve(diag(nrow(W)) - lambda * W)
  eps <- matrix(rnorm(t * nrow(W)), t, nrow(W))
  apply(eps %*% t(spill), 2, cumsum)
}
as_pseries <- function(m) {
  long <- data.frame(id = rep(seq_len(ncol(m)), each = nrow(m)),
                     period = rep(seq_len(nrow(m)), ncol(m)),
                     y = as.vector(m))
  plm::pdata.frame(long, index = c("id", "period"))$y
}
```

**The size table.** Rejection frequency of a *true* null at nominal 5%, over
200 panels of T = 30, with one lagged difference in every test. A correctly
sized test prints 0.05:

```{r}
#| label: s1-5-size
battery <- function(m) {
  p <- function(...) unname(plm::purtest(m, lags = 1, ...)$statistic$p.value)
  c(LLC_none = p(test = "levinlin", exo = "none"),
    LLC_intercept = p(test = "levinlin", exo = "intercept"),
    IPS = p(test = "ips", exo = "intercept"),
    MW = p(test = "madwu", exo = "intercept"),
    CIPS = plm::cipstest(as_pseries(m), type = "drift", lags = 1)$p.value)
}
set.seed(1)
lambdas <- c(0, 0.4, 0.8)
size <- sapply(lambdas, function(l) {
  rowMeans(replicate(200, battery(rw_panel(l))) <= 0.05)
})
colnames(size) <- paste0("lambda_", lambdas)
size
```

Read the λ = 0 column before anything else. These series are independent —
there is no spatial dependence to distort anything — and four of the five tests
already reject a true null between two and a half and twelve times too often.
Only CIPS is where it should be, at 0.04.

Then read across. The spatial effect is real and it is the smaller of the two:
LLC without an intercept goes 0.13 → 0.31, IPS 0.18 → 0.28, CIPS 0.04 → 0.07.
Maddala–Wu and the intercept version of LLC are already so distorted that λ
moves them inside Monte Carlo noise — with 200 replications the standard error
is about 0.015 near 5% and 0.035 near 50%. The ranking matches @pesaran2015
[p. 839]: the CD-based test degrades least.

**Where the λ = 0 distortion comes from.** Not from W, and not from a bug: it
is the shape of the panel. LLC's asymptotics need *T* to grow faster than *N*
[@levin2002; @baltagi2021, pp. 340–343], and its size tracks N/T almost
mechanically on independent random walks:

```{r}
#| label: s1-5-shape
iid_rw <- function(n, t) apply(matrix(rnorm(t * n), t, n), 2, cumsum)
llc_size <- function(n, t, reps = 200) {
  p <- replicate(reps, unname(plm::purtest(iid_rw(n, t), test = "levinlin",
                                           exo = "intercept",
                                           lags = 1)$statistic$p.value))
  mean(p <= 0.05)
}
set.seed(1)
shape <- data.frame(N = c(10, 25, 49, 100), T = c(250, 100, 30, 30))
shape$N_over_T <- round(shape$N / shape$T, 2)
shape$size <- mapply(llc_size, shape$N, shape$T)
shape
```

At N/T = 0.04 — the regime LLC was designed for — size is 0.09 and the test is
usable. At N/T = 3.33 it is 0.855: the test rejects a true null five times in
six. The project's own target shape, N ≈ 170 over T ≈ 20, is N/T ≈ 8.5, off
the right-hand end of this table.

**Book check.** There is none, and that is the finding. No published number
covers this W, this shape and this code, so nothing here is verified against a
book — only against the 5% the tests themselves claim. What the table does
support is a decision rule, and it is a conservative one: at the shapes this
project works with, report CIPS, report it as a test whose size survives both
the panel shape and moderate spatial dependence, and treat every
first-generation p-value in
[the first-generation section](panel-spat.qmd#sec-ur-first-gen) as an upper
bound on the evidence rather than a measurement of it.

Two limits on that rule. The rejection rule for CIPS is "statistic beyond the
tabulated 5% critical value", because `cipstest` has no exact p-values, so its
row is coarser than the others. And CIPS defactors, which @pesaran2015 [p. 839]
warns is invalid under weak dependence — it holds its size here, at λ ≤ 0.8 on
a sparse rook W, which is evidence for this design and not a general licence.
