### Cross-unit cointegration {#sec-ur-cross-coint}

::: {.callout-tip appearance="simple"}
The assumption every second-generation test makes and none of them tests:
[theory note 07](theory/07-panel-unit-roots.qmd).
:::

**Theory.** Cross-*correlation* is units moving together. Cross-unit
*cointegration* is stronger: a linear combination of *different units'* series
is stationary, so the units share a stochastic trend rather than merely
correlated shocks. @pesaran2015 [p. 836] separates the two and warns that panel
unit-root tests may be "severely biased" under the second while tolerating the
first.

The tests of [the previous sections](panel-spat.qmd#sec-ur-second-gen) divide
on it. Moon–Perron and Pesaran's CADF assume it away — CADF allows every unit
to load on one common I(1) factor, which is what the cross-section averages
remove, and rules out further cointegrating relations among what is left
[@pesaran2015, p. 837]. Bai–Ng's PANIC allows it, but @pesaran2015
[pp. 837–838] records that the pooled version of those tests is asymptotically
invalid and that a later alternative was needed. Subsampling
[@choi2007] handles it directly and needs *N* small relative to *T*
[@pesaran2015, p. 838] — the opposite of the shape this project works at.

**Data.** Simulated pairs, because the distinction is a property of the
data-generating process and no diagnostic in this book reports it. Both pairs
below are built from I(1) series and both are strongly correlated; only the
first is cointegrated:

```{r}
#| label: s1-6-pairs
t <- 100
factor_pair <- function() {
  f <- cumsum(rnorm(t))
  cbind(a = 0.8 * f + rnorm(t), b = 1.3 * f + rnorm(t))
}
spatial_pair <- function() {
  shocks <- matrix(rnorm(2 * t), t, 2) %*% chol(matrix(c(1, .8, .8, 1), 2))
  m <- apply(shocks, 2, cumsum)
  colnames(m) <- c("a", "b")
  m
}
```

The first pair shares one stochastic trend, so $1.3 a_t - 0.8 b_t$ is
stationary. The second has two separate stochastic trends whose *increments*
are correlated at 0.8 — the spatial case — and no stationary combination
exists. Johansen's trace test tells them apart; correlation does not:

```{r}
#| label: s1-6-johansen
johansen <- function(dgp, reps = 200) {
  out <- replicate(reps, {
    x <- dgp()
    z <- urca::ca.jo(x, type = "trace", ecdet = "none", K = 2)
    c(unname(z@teststat[2]) > z@cval[2, "5pct"], abs(cor(x)[1, 2]))
  })
  c(reject_r0 = mean(out[1, ]), mean_abs_cor = mean(out[2, ]))
}
set.seed(1)
round(rbind(common_factor = johansen(factor_pair),
            correlated_increments = johansen(spatial_pair)), 3)
```

The rank test rejects "no cointegration" every time on the first pair and 5% of
the time on the second, which is exactly its nominal size. The correlations are
0.90 and 0.77 — close enough that no correlation-based diagnostic in this book,
CD or α included, could tell the two panels apart. That is the whole problem:
the quantity that breaks the tests is invisible to the quantities that get
reported.

::: {.callout-important appearance="simple"}
**Gap.** There is no R implementation of a *panel* test for cross-unit
cointegration and no verification target for one, so this has to travel as a
stated limitation rather than a result. The pairwise route above does not
substitute for it: at N = 49 it is 1176 rank tests, at the project's N ≈ 170 it
is 14 365, and this book has no aggregation rule for them that any of its
sources validates.
:::

**Book check.** None, and none is possible: nothing in @baltagi2021,
@pesaran2015 or @croissant2019 prints a number for an implementation that does
not exist. What the simulation above verifies is only its own arithmetic — a
rank test separates two DGPs that a correlation cannot — and `urca::ca.jo`
reproduces Johansen's published money-demand examples on its own help page,
which is the credit the function is used on here.

So the limitation, stated once, in the form it has to be carried into every
later stage: the CIPS results of
[the second-generation section](panel-spat.qmd#sec-ur-second-gen) hold under
the assumption that no cointegrating relation links the idiosyncratic
components of different states. That assumption is untested here, it is not
testable with the packages in this project, and the panel it is made on is one
where a common factor is strong enough to move the income elasticity from 0.30
to 1.14.
