## Spatial weights {#sec-sp-weights}

**Question.** Who counts as Arizona's neighbour, and what does that choice
commit the model to?

**Intuition.** W is an assumption, not data. Row *i* of W lists the states
whose outcome enters state *i*'s equation, and the weight of each. When the
rows are standardised to sum to one, $\sum_j w_{ij} y_{jt}$ is the
neighbours' average, and ρ reads as "how much of the neighbours' average".
Three common choices encode three beliefs: states that share a border; the
*k* nearest states; every state within some distance, the closer ones
counting more. The estimates inherit whichever belief is chosen, so a
spatial parameter is only comparable across fits that share a W
[@anselin1988, ch. 3].

::: {.callout-tip appearance="simple"}
Why the parameter space of ρ depends on W, why a SAR needs weights that are
not too spread out, and why a sparse W still links every unit to every other:
[spatial weights](theory-spatial-weights.qmd).
:::

Four checks come before any model, all cheap [@pesaran2015, sec. 30.3]:

1. **Parameter space.** $I - \rho W$ must be invertible. It is whenever
   $|\rho| < 1/\tau^*$ with $\tau^* = \min(\lVert W\rVert_1, \lVert
   W\rVert_\infty)$, the larger of the column sums against the larger of the
   row sums [@kelejian2010; @pesaran2015, p. 799]. The exact interval is
   $(1/\omega_{\min}, 1/\omega_{\max})$, with ω the real eigenvalues of W.
2. **Granularity.** A spatial lag is endogenous only when some weights stay
   large as N grows. If every $w_{ij}$ were of order $1/N$, the lag would
   be a cross-section average and nothing more [@pesaran2015, p. 800].
3. **Sparsity.** W has few non-zero cells. $(I - \rho W)^{-1}$, which a
   spatial lag model puts between every unit and every other, has almost
   none [@pesaran2015, p. 801].
4. **Standardisation.** Row sums equal one, or the parameter ρ changes
   meaning from row to row.

**Run it.** `usaw49` [@croissant2019] is the binary contiguity of the 48
contiguous states and the District of Columbia, row-standardised. Arizona's
row, and how many neighbours each state has:

```{r}
#| label: c03-1-rows
dim(usaw49)
usaw49["ARIZONA", usaw49["ARIZONA", ] > 0]
table(neighbours = rowSums(usaw49 > 0))
```

The four checks, one line each:

```{r}
#| label: c03-1-checks
c(norm_1 = max(colSums(usaw49)), norm_inf = max(rowSums(usaw49)))
omega <- Re(eigen(usaw49, only.values = TRUE)$values)
c(rho_min = 1 / min(omega), rho_max = 1 / max(omega))
rownames(usaw49)[apply(usaw49, 1, max) == 1]
S <- solve(diag(49) - 0.5 * usaw49)
c(nonzero_w = mean(usaw49 > 0), nonzero_inverse = mean(abs(S) > 1e-8))
```

Two alternatives built from the state centroids with **spdep**
[@bivand2013]: the four nearest states, and every state within 800 km.
`w_checks()` runs the same four checks on each:

```{r}
#| label: c03-1-alternatives
xy <- sf::st_coordinates(sf::st_centroid(sf::st_geometry(us)))
nb_knn <- spdep::knn2nb(spdep::knearneigh(xy, k = 4, longlat = TRUE))
w_knn <- spdep::nb2mat(nb_knn, style = "W")
nb_dist <- spdep::dnearneigh(xy, 0, 800, longlat = TRUE)
w_dist <- spdep::nb2mat(nb_dist, style = "W")
round(rbind(contiguity = w_checks(usaw49), nearest_4 = w_checks(w_knn),
            within_800km = w_checks(w_dist)), 3)
```

```{r}
#| label: c03-1-picture
#| fig-height: 3
#| fig-cap: "Three maps of who is whose neighbour: shared borders (left),
#|   the four nearest state centroids (middle), every centroid within 800
#|   km (right). Each line is one non-zero cell of W."
#| fig-alt: "Three outline maps of the contiguous United States with lines
#|   drawn between state centres. The left map has a sparse web following
#|   borders. The middle map has a web of similar density that does not
#|   always follow borders, with long lines in the West. The right map is
#|   dense in the East, where states are small, and sparse in the West."
op <- graphics::par(mfrow = c(1, 3), mar = c(0, 0, 2, 0))
plot(sf::st_geometry(us), col = "grey95", main = "Shared border")
draw_links(sf::st_geometry(us), usaw49, col = "firebrick")
plot(sf::st_geometry(us), col = "grey95", main = "Four nearest")
draw_links(sf::st_geometry(us), w_knn, col = "steelblue")
plot(sf::st_geometry(us), col = "grey95", main = "Within 800 km")
draw_links(sf::st_geometry(us), w_dist, col = "darkorange")
graphics::par(op)
```

```{r}
#| label: c03-1-sparsity
#| fig-cap: "The contiguity matrix W (left) and the matrix (I − 0.5 W)⁻¹ a
#|   spatial lag model works through (right), states in alphabetical order
#|   on both axes, darker for larger. W has 9% non-zero cells; its inverse
#|   has none equal to zero."
#| fig-alt: "Two square grids of 49 by 49 cells. The left one is almost
#|   all white, with a scatter of dark cells near the diagonal and a few
#|   away from it. The right one is shaded everywhere, dark on the diagonal
#|   and fading away from it, with no white cells."
op <- graphics::par(mfrow = c(1, 2), mar = c(1, 1, 2, 1))
graphics::image(t(usaw49[49:1, ]), col = grey(seq(1, 0.1, length.out = 16)),
                axes = FALSE, main = "W")
graphics::image(t(abs(S)[49:1, ]), col = grey(seq(1, 0.1, length.out = 16)),
                axes = FALSE, main = expression((I - 0.5 * W)^-1))
graphics::par(op)
```

**Read it.**

- **Parameter space.** The row sums are 1 and the largest column sum is
  1.7, so $\tau^* = 1$ and any $|\rho| < 1$ is safe. The eigenvalues widen
  the interval on the negative side only, to −1.39. *So:* a ρ̂ printed
  above 1 is a broken fit, not a strong spillover.
- **Granularity.** Maine's row has one non-zero cell, with weight 1: its
  "neighbours' average" is New Hampshire alone. Most states average two to
  eight neighbours. *So:* the lag is no cross-section average, and a SAR
  model will treat it as endogenous, which is right.
- **Sparsity.** 9% of the cells of W are non-zero; 99.7% of the cells of
  $(I - 0.5W)^{-1}$ are. *So:* a shock in California reaches Maine, through
  a chain of neighbours, in any spatial lag model.
- **Alternatives.** The four-nearest W caps every weight at 0.25 and keeps
  links in the West that no border justifies; the 800 km W has twice the
  links, almost all of them in the East. *So:* the chapter keeps `usaw49`
  and reports every ρ̂ as a number *for this W*.

::: {.callout-note appearance="simple"}
**Next.** With W on the table, ask whether the dependence the [previous
chapter](../02-cross-dependence/index.qmd#sec-csd-exponent) left behind
follows it ([does it follow the map](index.qmd#sec-sp-map)).
:::

::: {.callout-note collapse="true"}
## Recipe

```r
# w: an N x N matrix, units in the order of the panel's unit index.
w <- w / rowSums(w)                              # row standardise
spdep::mat2listw(w, style = "W")                 # what spml() takes
w_checks(w)                                      # the four checks
xy <- sf::st_coordinates(sf::st_centroid(geometry))
spdep::knn2nb(spdep::knearneigh(xy, k = 4, longlat = TRUE))
spdep::dnearneigh(xy, 0, 800, longlat = TRUE)
```
:::

::: {.callout-caution collapse="true"}
## Pitfalls

**`mat2listw()` without `style`.** The object then carries style `M`,
"missing", and methods that need to know the rows sum to one refuse it
with `Only row-standardised weights supported`. Pass `style = "W"`.

**A disconnected W.** `riceww` calls two farms neighbours when they share
a village, so it is six blocks with nothing between them. **spdep** says
so:

```{r}
#| label: c03-1-subgraphs
#| warning: true
ricelw <- spdep::mat2listw(riceww, style = "W")
```

The fit is legal. What changes is the meaning of "neighbour": a shock in
one village never reaches another, however large λ is.

**Row standardisation breaks symmetry.** `riceww` is symmetric; `usaw49`
is not, since Maine gives New Hampshire weight 1 and gets 1/3 back. Methods
that want a symmetric W, such as the `"moments"` option of `impacts()`,
need its similar symmetric form [@bivand2018].

**Units in a different order from W.** Nothing checks that row *i* of W is
the *i*-th unit of the panel. `usaw49` follows the alphabetical order of
`HousePricesUS`; `us_states_sf()` reorders the polygons to match.
:::

::: {.callout-tip collapse="true"}
## Book check

None: the checks verify themselves. A row-standardised W must have
$\lVert W\rVert_\infty = 1$ and largest eigenvalue 1, and the bound
$1/\tau^*$ must lie inside the eigenvalue interval. All three hold above.
`.docs/panel-spat-PLAN.md` records no published target for section 5.1.
:::
