## Does it follow the map? {#sec-sp-map}

**Question.** After the national force is out, do neighbouring states'
house prices still move together more than random pairs of states do?

**Intuition.** The CD test of the [previous
chapter](../02-cross-dependence/index.qmd#sec-csd-tests) averages all
1 176 pairwise correlations. The **local CD** test averages only the 130
pairs that `usaw49` names as neighbours [@pesaran2004; @croissant2019, sec.
10.1.2.1]. A large value says neighbours move together. It does not say
why: a national factor moves every pair, neighbours included, and then the
neighbour pairs are a sample drawn from a population of correlated pairs,
which reads as geography when it is not [@croissant2019, sec. 10.1.2.2].
The **randomised-W test** of @millo2017 settles it. Shuffle the map, so that
each state is handed a random set of "neighbours", and compute the local
statistic again; do that 999 times. If the states are exchangeable, the
true map is nothing special and its statistic sits inside the shuffled
pile. If the dependence follows the map, the true statistic is the most
extreme of the thousand.

::: {.callout-tip appearance="simple"}
Why plain CD cannot answer this question, why local CD alone over-reads a
factor as geography, and what the shuffled maps control for:
[testing for spatial dependence](theory-spatial-dependence-tests.qmd).
:::

With $\hat\rho_{ij}$ the correlation of units *i* and *j* over T periods and
$w_{ij} = 1$ for neighbours and 0 otherwise [@croissant2019, sec. 10.1.2.1]:

$$
CD(p) = \sqrt{\frac{T}{\sum_{i<j} w_{ij}}}\ \sum_{i<j} w_{ij}\,\hat\rho_{ij}
  \;\xrightarrow{d}\; N(0,1).
$$

The randomised test takes a statistic τ of this kind, computes it under B
shuffled maps, $\tau_1, \dots, \tau_B$, and reports the share of shuffles
at least as extreme as the true map [@croissant2019, eq. 10.1]:

$$
p = \frac{1 + \sum_{h=1}^{B} \mathbf{1}[\,|\tau_h| \ge |\tau|\,]}{B + 1}.
$$

**The tests.** Local CD [@pesaran2004]. Null: no correlation between
neighbours. A rejection means neighbours move together, for whatever
reason. Randomised W [@millo2017]. Null: the units are exchangeable, so the
map does not matter. A rejection means the co-movement follows this map
and not merely the panel.

**Run it.** The four calls of @croissant2019 [sec. 10.1.2.2], in their
order: the raw price index, then the residuals once the national force is
out. `ccemg` and its residuals come from the setup chunk; `mg` is the same
regression without the cross-state averages.

```{r}
#| label: c03-2-local-prices
plm::pcdtest(hp$price, w = usaw49)
```

```{r}
#| label: c03-2-rw-prices
splm::rwtest(hp$price, w = usaw49, replications = 999, seed = 1)
```

```{r}
#| label: c03-2-local-cce
plm::pcdtest(residuals(ccemg), w = usaw49)
```

```{r}
#| label: c03-2-rw-mg
mg <- plm::pmg(log(price) ~ log(income), data = HousePricesUS, model = "mg")
splm::rwtest(residuals(mg), w = usaw49, replications = 999, seed = 1)
```

How large is what is left? @holly2010 fit a spatial lag model with no
regressors to the defactored residuals, so that ρ measures the strength of
the local co-movement on its own. `spreml` with `errors = "ols"` is that
model; the pooled regression of the residual on its spatial lag, "spatial
OLS", is the shortcut that the endogeneity of $Wy$ biases:

```{r}
#| label: c03-2-sar
e_dat <- data.frame(state = hp_idx[[1]], year = hp_idx[[2]], e = hp$e_cce)
sar_e <- splm::spreml(e ~ 1, data = e_dat, w = usaw49, lag = TRUE,
                      errors = "ols")
summary(sar_e)$ARCoefTable
lmtest::coeftest(plm::plm(e ~ splm::slag(e, listw = usalw) - 1, data = e_dat,
                          model = "pooling"))
```

The 1 176 correlations behind the local statistic, split by whether the
pair are neighbours:

```{r}
#| label: c03-2-pairs
#| fig-cap: "Pairwise correlations of the CCEMG residuals over 29 years,
#|   for the 130 neighbour pairs (left) and the 1 046 other pairs (right).
#|   Thick line: the mean of each pile. Dashed line at zero."
#| fig-alt: "Two histograms of correlations between minus one and one. The
#|   neighbour pile sits to the right, with its mean near 0.5. The
#|   non-neighbour pile is centred close to zero with a long spread either
#|   side."
e_mat <- tapply(hp$e_cce, list(hp_idx[[1]], hp_idx[[2]]), identity)
r <- stats::cor(t(e_mat))
is_pair <- lower.tri(r)
is_nb <- usaw49 > 0
r_nb <- r[is_pair & is_nb]
r_far <- r[is_pair & !is_nb]
op <- graphics::par(mfrow = c(1, 2))
graphics::hist(r_nb, breaks = 20, xlim = c(-1, 1), main = "Neighbours",
               xlab = "pairwise correlation")
graphics::abline(v = c(0, mean(r_nb)), lty = c(2, 1), lwd = c(1, 3))
graphics::hist(r_far, breaks = 30, xlim = c(-1, 1), main = "Not neighbours",
               xlab = "pairwise correlation")
graphics::abline(v = c(0, mean(r_far)), lty = c(2, 1), lwd = c(1, 3))
graphics::par(op)
```

And the shuffle itself. `rwtest` keeps only its p-value, so the picture
rebuilds it: 999 random orderings of the states, the local CD of the CCEMG
residuals under each, and the value under the true map:

```{r}
#| label: c03-2-shuffle
#| fig-cap: "Local CD of the CCEMG residuals under 999 shuffled maps, and
#|   under the true map (vertical line). The shuffled values are what
#|   'no spatial structure' looks like on this panel."
#| fig-alt: "A histogram of statistics concentrated between about minus
#|   three and three, and a single vertical line far to the right of it
#|   near 28."
local_cd <- function(w) unname(plm::pcdtest(hp$e_cce, w = w)$statistic)
set.seed(1)
shuffled <- replicate(999, {
  s <- sample(49)
  local_cd(usaw49[s, s])
})
graphics::hist(shuffled, breaks = 30, xlim = range(c(shuffled, 30)),
               main = "", xlab = "local CD under a shuffled map")
graphics::abline(v = local_cd(usaw49), lwd = 3)
```

**Read it.**

- **Rejects (price index).** Local CD z = 37; the randomised test p =
  0.002, the smallest value 999 shuffles can give. *So:* the index moves
  with its neighbours, but on the raw series that is also what a national
  force looks like.
- **Rejects (residuals, national force out).** Local CD on the CCEMG
  residuals z = 28; the randomised test on the MG residuals p = 0.002. The
  neighbour pile sits well to the right of the other, and the true map's
  statistic is nowhere near the shuffled pile. *So:* the local co-movement
  is not a shadow of the factor. It follows the map.
- **How much.** ρ = 0.65 with a standard error of 0.02 on the defactored
  residuals. The spatial OLS shortcut says 0.88. *So:* a state's residual
  carries about two thirds of its neighbours' average residual, and the
  shortcut overstates that by a third, which is the bias of regressing on
  an endogenous $Wy$.
- **Misleads (local CD on the raw index).** z = 37 on its own would have
  passed for spatial evidence, and most of it is the national force that
  the [previous chapter](../02-cross-dependence/index.qmd#sec-csd-exponent)
  measured at α = 0.83. *So:* never read local CD without the shuffle
  beside it.

::: {.callout-note appearance="simple"}
**Next.** The dependence follows the map: decide which of the two
equations it belongs in ([lag or error](index.qmd#sec-sp-lag-or-error)).
It does not: a spatial model buys nothing here; report standard errors
robust to whatever dependence is left ([core
chapter](../05-core/index.qmd)).
:::

::: {.callout-note collapse="true"}
## Recipe

```r
# y: a pseries, or the residuals of a plm, pmg or pcce fit;
# w: an N x N matrix with rows in the order of the unit index.
plm::pcdtest(y, w = w)                                    # local CD
splm::rwtest(y, w = w, replications = 999, seed = 1)      # shuffled maps
splm::rwtest(y, w = w, replications = 999, seed = 1, order = 2)
```
:::

::: {.callout-caution collapse="true"}
## Pitfalls

**Both tests read W as yes or no.** `pcdtest(w = )` and `rwtest(w = )` pass
W through `as.logical()`, so a distance-weighted W is tested as "within the
cutoff or not". They test the neighbourhood *structure*, never the weights.
Do not describe either as using an inverse-distance W.

**99 shuffles is the default.** The smallest p-value 99 shuffles can
return is 0.01, and the pile is coarse. Use at least 999, and pass `seed`
so the number can be reproduced.

**Global CD on residuals with period effects.** Two-way fixed effects and
CCE both force the residuals to cancel across units in every period, which
kills the global CD test by construction, as the [previous
chapter](../02-cross-dependence/index.qmd#sec-csd-tests) showed. Local CD
and the randomised test survive it, because they never average over all
pairs [@croissant2019, sec. 10.1.2.2].

**`rwtest` keeps no statistic.** The object holds a p-value and nothing
else; the shuffled pile has to be rebuilt by hand, as above, for a picture
or for a different statistic.
:::

::: {.callout-tip collapse="true"}
## Book check

@croissant2019 [sec. 10.1.2.2, Example 10.2] print z = 37 and p = 0.002 for
the price index and z = 28 and p = 0.002 for the residuals; the four calls
above are those runs, and `.docs/panel-spat-PLAN.md` names this
four-number pattern the most useful verification target in the spatial
stage. @croissant2019 [sec. 10.2.2.2, Example 10.4] print λ = 0.6498 with
standard error 0.02037 for the pure spatial lag on the CCEMG residuals, and
0.8831 with 0.0251 for spatial OLS; the `c03-2-sar` chunk above is that run.
:::
