## Lag or error? {#sec-sp-lag-or-error}

**Question.** Does a neighbour's house price move mine, or do we only share
the shocks? And on a rice farm: does a neighbour's harvest move mine, or do
we share the weather?

**Intuition.** Fit the regression with no spatial term and look at its
residuals twice. If they move with the neighbours' *outcome*, a spatial lag
was left out. If they move with the neighbours' *residuals*, a spatial
error was left out. The trouble is that each plain test also fires on the
other's problem: an omitted lag leaves a spatially correlated error behind,
and vice versa. @anselin1996 made each test robust to a small amount of the
other effect, so the robust lag test asks "is there a lag, allowing for
some error?" and the robust error test the reverse. When one plain test
rejects and its robust version does not, the signal was the other effect's
shadow.

::: {.callout-tip appearance="simple"}
What each specification claims about the world, what an omitted lag does
to the estimates that an omitted error does not, and when the encompassing
model is the honest answer: [spatial model
choice](theory-spatial-model-choice.qmd).
:::

The plain error test is a Lagrange multiplier statistic on the residuals
*e* of the non-spatial fit [@burridge1980; @anselin1996, eq. 5]:

$$
LM_\lambda =
  \frac{\left(e'We / \hat\sigma^2\right)^2}{\operatorname{tr}(W'W + W^2)}
  \;\sim\; \chi^2_1,
$$

the squared spatial correlation of the residuals, scaled. The lag test
replaces $e'We$ by $e'Wy$. The robust versions subtract from each
numerator the part that the other effect would produce, so that a small
λ does not inflate the lag test and a small ρ does not inflate the error
test [@anselin1996, eqs. 10 and 13].

**The tests.** `slmtest` [@anselin1996; @croissant2019, sec. 10.3.4.2]:
`lml` and `lme`, the plain lag and error tests; `rlml` and `rlme`, their
robust versions. Null: no lag (ρ = 0), or no error (λ = 0). A rejection of
a *robust* test means that effect is there in its own right.

**Run it, house prices.** The spatial model has to be fitted to the panel
with the national force taken out, or it reads the factor as geography.
This is the two-stage approach of @bailey2016spatio. Stage one is the CCE
transformation of the [previous
chapter](../02-cross-dependence/index.qmd#sec-cce): regress each state's
log price, and its log income, on the cross-state averages of both, and
keep the residuals. Pooled OLS on those residuals *is* the CCEP estimator,
which is the check that stage one did what it should:

```{r}
#| label: c03-3-partial
lp <- log(as.numeric(hp$price))
li <- log(as.numeric(hp$income))
avg <- cbind(lp_bar = ave(lp, hp_idx[[2]]), li_bar = ave(li, hp_idx[[2]]))
hp$lp_t <- cce_partial(lp, avg, hp_idx[[1]])
hp$li_t <- cce_partial(li, avg, hp_idx[[1]])
coef(plm::plm(lp_t ~ li_t - 1, data = hp, model = "pooling"))
```

1.199, the CCEP slope of the previous chapter to the third decimal. Stage
two asks the four questions of the transformed panel:

```{r}
#| label: c03-3-hp
splm::slmtest(lp_t ~ li_t, data = hp, listw = usalw, model = "within",
              test = "lml")
splm::slmtest(lp_t ~ li_t, data = hp, listw = usalw, model = "within",
              test = "lme")
splm::slmtest(lp_t ~ li_t, data = hp, listw = usalw, model = "within",
              test = "rlml")
splm::slmtest(lp_t ~ li_t, data = hp, listw = usalw, model = "within",
              test = "rlme")
```

**Run it, rice farms.** Output on seed, labour and land, all in logs, with
village and season dummies, as @croissant2019 [sec. 10.3.4.2] specify it:

```{r}
#| label: c03-3-rice
rice_f <- log(goutput) ~ log(seed) + log(totlabor) + log(size) + region + time
splm::slmtest(rice_f, data = rice, listw = ricelw, test = "lml")
splm::slmtest(rice_f, data = rice, listw = ricelw, test = "lme")
splm::slmtest(rice_f, data = rice, listw = ricelw, test = "rlml")
splm::slmtest(rice_f, data = rice, listw = ricelw, test = "rlme")
```

The error signal, drawn: each unit's residual from the non-spatial
fixed-effects fit against its neighbours' average residual. The slope of
the line is Moran's *I*, the correlation that $LM_\lambda$ squares:

```{r}
#| label: c03-3-picture
#| fig-cap: "Moran scatterplots of the fixed-effects residuals: each unit's
#|   residual against the average residual of its neighbours, with the
#|   least-squares line, for the transformed house-price panel (left) and
#|   the rice panel (right)."
#| fig-alt: "Two scatterplots with a fitted line through the origin. On
#|   the left, 1 421 points in a tight upward-sloping cloud with a line of
#|   slope about 0.6. On the right, 1 026 points in a looser cloud, also
#|   sloping up, with a line of slope about 0.5."
rice_f0 <- log(goutput) ~ log(seed) + log(totlabor) + log(size)
fe_hp <- plm::plm(lp_t ~ li_t, data = hp, model = "within")
fe_rice <- plm::plm(rice_f0, data = rice, model = "within")
moran_plot <- function(e, listw, main) {
  we <- splm::slag(e, listw = listw)
  plot(as.numeric(e), as.numeric(we), pch = 16, col = "grey55",
       xlab = "own residual", ylab = "neighbours' average residual",
       main = main)
  graphics::abline(lm(as.numeric(we) ~ as.numeric(e)), lwd = 2)
  graphics::abline(h = 0, v = 0, lty = 3)
}
op <- graphics::par(mfrow = c(1, 2))
moran_plot(residuals(fe_hp), usalw, "House prices")
moran_plot(residuals(fe_rice), ricelw, "Rice farms")
graphics::par(op)
```

**Read it.**

- **Rice: lag rejected only by the plain test.** `lml` = 39 rejects;
  `rlml` = 0.17 does not. `lme` = 245 and `rlme` = 206 both reject. *So:*
  the lag signal was the error's shadow; the farms share shocks, and no
  farm's harvest moves its neighbours'. A spatial error model.
- **House prices: both robust tests reject.** `rlml` = 32 and `rlme` = 28.
  *So:* neither effect can be ruled out by the other, and the tests have
  reached the edge of what they are built for. Fit both equations, and the
  model that holds both, and let the fits decide
  ([fixed effects](index.qmd#sec-sp-fe)).
- **Misleads (the plain lag test on rice).** `lml` = 39 with p ≈ 0 would
  have put a neighbour's harvest into the equation, and that coefficient
  would have been read as a spillover. *So:* report the robust pair, never
  a plain test alone.

::: {.callout-note appearance="simple"}
**Next.** Error, or both: fit the model with unit effects
([fixed effects](index.qmd#sec-sp-fe)). Both robust tests reject strongly:
fit the encompassing SARAR there as well and read its two parameters.
:::

::: {.callout-note collapse="true"}
## Recipe

```r
# f: a formula; d: a pdata.frame; listw: from spdep::mat2listw(w, "W").
splm::slmtest(f, data = d, listw = listw, model = "within", test = "rlml")
splm::slmtest(f, data = d, listw = listw, model = "within", test = "rlme")
```
:::

::: {.callout-caution collapse="true"}
## Pitfalls

**`slmtest` pools by default.** `model = "pooling"` is the default, and the
tests were derived for a cross-section, so unit effects are ignored unless
`model = "within"` is passed or the variables are demeaned in the formula
with `plm::Within()`. The rice run above pools because @croissant2019 do;
the box below repeats it within.

**Robust means locally robust.** Each robust test tolerates a *small*
amount of the other effect. When both robust tests reject strongly, as
on house prices, the tests "can behave suboptimally": estimate the
encompassing model instead of choosing between them.

**Only one kind of error.** `spatial.error` in **splm** is a spatial
autoregressive error. The moving-average and error-component forms of
@pesaran2015 [eqs. 30.5–30.6], under which the covariance stays sparse,
are not implemented anywhere in this book's tools.
:::

::: {.callout-tip collapse="true"}
## Book check

@croissant2019 [sec. 10.3.4.2, Example 10.11] print 39.28, 244.8, 0.1654
and 205.7 for `lml`, `lme`, `rlml` and `rlme` on the pooled rice model;
the `c03-3-rice` chunk above is that run. They then demean the variables
with `Within()` and print 125.2, 604.3, 1.538 and 480.6:

```{r}
#| label: c03-3-book-check
rice_w <- plm::Within(log(goutput)) ~ plm::Within(log(seed)) +
  plm::Within(log(totlabor)) + plm::Within(log(size)) + region + time
splm::slmtest(rice_w, data = rice, listw = ricelw, test = "lml")
splm::slmtest(rice_w, data = rice, listw = ricelw, test = "lme")
splm::slmtest(rice_w, data = rice, listw = ricelw, test = "rlml")
splm::slmtest(rice_w, data = rice, listw = ricelw, test = "rlme")
```
:::
