## Fixed effects {#sec-sp-fe}

**Question.** With everything fixed about a state, or a farm, taken out:
how far does a shock spread, and what happens to the slope when the model
allows it to?

**Intuition.** Take each unit's mean out of every variable, as the
within estimator does, then fit the spatial model by maximum likelihood.
The likelihood carries one term a plain regression does not, the
log-determinant of $I - \lambda W$ (or $I - \rho W$), which is the price
of a model in which every unit's error, or outcome, depends on every
other's. A spatial error buys *precision*: the within estimator was
consistent already, and the spatial fit reweights the observations the
way GLS does. A spatial lag buys *consistency*: a within estimator that
leaves $Wy$ out is biased [@pesaran2015, p. 801]. The generalised moments
estimator skips the determinant and the normality assumption, at the
price of no standard error for λ.

::: {.callout-tip appearance="simple"}
The likelihood with its Jacobian, what the within transformation does to
it, the moment conditions behind the GM fit, and the naming trap in the
printed output: [fitting spatial panels](theory-spatial-estimation.qmd).
:::

For the spatial error model with unit effects removed, the log-likelihood
is [@anselin1988, ch. 6; @croissant2019, sec. 10.2.3]

$$
\ln L = -\frac{NT}{2}\ln(2\pi\sigma^2) + T \ln|I_N - \lambda W|
  - \frac{1}{2\sigma^2}\, u' \big(I_T \otimes B'B\big)\, u,
\qquad B = I_N - \lambda W,
$$

with $u = \tilde y - \tilde X\beta$ the within residuals. The spatial lag
model puts $B$ on the outcome instead, $B\tilde y = \tilde X\beta + u$,
and the two together give the SARAR model.

**The estimators.** `spml(model = "within")` [@millo2012; @croissant2019,
sec. 10.3.1]: `spatial.error = "b"` for the error, `lag = TRUE` for the
lag, both for SARAR. `spgm(model = "within", spatial.error = TRUE)` for
the GM fit [@kapoor2007; @mutl2011]. In the printed output of a within
fit the *lag* parameter is called `lambda` and the *error* parameter
`rho`, the reverse of this book's naming; the text below translates.

**Run it, house prices.** On the transformed panel of the [previous
section](index.qmd#sec-sp-lag-or-error). The plain within fit first, for
the slope the spatial fits are compared with:

```{r}
#| label: c03-4-hp-fe
summary(fe_hp)$coefficients
```

```{r}
#| label: c03-4-hp-sem
sem_hp <- splm::spml(lp_t ~ li_t, data = hp, listw = usalw,
                     model = "within", lag = FALSE, spatial.error = "b")
summary(sem_hp)$CoefTable
```

```{r}
#| label: c03-4-hp-sar
sar_hp <- splm::spml(lp_t ~ li_t, data = hp, listw = usalw,
                     model = "within", lag = TRUE, spatial.error = "none")
summary(sar_hp)$CoefTable
```

```{r}
#| label: c03-4-hp-sarar
sarar_hp <- splm::spml(lp_t ~ li_t, data = hp, listw = usalw,
                       model = "within", lag = TRUE, spatial.error = "b")
summary(sarar_hp)$CoefTable
```

```{r}
#| label: c03-4-hp-gm
gm_hp <- splm::spgm(lp_t ~ li_t, data = hp, listw = usaw49,
                    model = "within", lag = FALSE, spatial.error = TRUE)
coef(gm_hp)
gm_hp$rho
```

Then the check that closes the loop. A spatial error model's residuals
$u = \tilde y - \tilde X\hat\beta$ are correlated across neighbours *by
construction*; that is what the model says about them. What it claims to
be free of spatial correlation is the innovation $\varepsilon = (I -
\hat\lambda W)u$, period by period. Test both:

```{r}
#| label: c03-4-hp-check
hp$u_sem <- splm_residuals(sem_hp, hp_idx)
hp$eps_sem <- spatial_filter(hp$u_sem, usaw49, coef(sem_hp)[["rho"]], hp_idx)
plm::pcdtest(hp$u_sem, w = usaw49)
splm::rwtest(hp$u_sem, w = usaw49, replications = 999, seed = 1)
plm::pcdtest(hp$eps_sem, w = usaw49)
splm::rwtest(hp$eps_sem, w = usaw49, replications = 999, seed = 1)
```

**Run it, rice farms.** The same five fits on output, seed, labour and
land, farm effects removed, as @croissant2019 [sec. 10.3.1] fit them:

```{r}
#| label: c03-4-rice-fe
summary(fe_rice)$coefficients
```

```{r}
#| label: c03-4-rice-sem
sem_rice <- splm::spml(rice_f0, data = rice, listw = ricelw,
                       model = "within", lag = FALSE, spatial.error = "b")
summary(sem_rice)$CoefTable
```

```{r}
#| label: c03-4-rice-sar
sar_rice <- splm::spml(rice_f0, data = rice, listw = ricelw,
                       model = "within", lag = TRUE, spatial.error = "none")
summary(sar_rice)$CoefTable
```

```{r}
#| label: c03-4-rice-sarar
sarar_rice <- splm::spml(rice_f0, data = rice, listw = ricelw,
                         model = "within", lag = TRUE, spatial.error = "b")
summary(sarar_rice)$CoefTable
```

```{r}
#| label: c03-4-rice-gm
gm_rice <- splm::spgm(rice_f0, data = rice, listw = ricelw,
                      model = "within", lag = FALSE, spatial.error = TRUE)
summary(gm_rice)$CoefTable
gm_rice$rho
```

```{r}
#| label: c03-4-rice-check
rice$u_sem <- splm_residuals(sem_rice, rice_idx)
rice$eps_sem <- spatial_filter(rice$u_sem, riceww, coef(sem_rice)[["rho"]],
                               rice_idx)
plm::pcdtest(rice$u_sem, w = riceww)
plm::pcdtest(rice$eps_sem, w = riceww)
splm::rwtest(rice$eps_sem, w = riceww, replications = 999, seed = 1)
```

The income elasticity of house prices under every treatment of the
national force and of space, and the three rice elasticities under the
five fits:

```{r}
#| label: c03-4-picture
#| fig-cap: "Left: the income elasticity of house prices with a 95%
#|   interval, under fixed effects on the raw panel, two-way fixed
#|   effects, a spatial lag on the raw panel, and the four fits on the
#|   CCE-transformed panel. Right: the rice elasticities of seed, labour
#|   and land under the within, SEM, SAR, SARAR and GM fits."
#| fig-alt: "Two dot-and-whisker charts. On the left, seven estimates of
#|   one slope: the raw fixed-effects and raw spatial-lag values sit low,
#|   near 0.15 and 0.2, the two-way value and the transformed fits sit
#|   between 0.7 and 1.2. On the right, three groups of five dots, one
#|   group per input: within each group the dots lie close together, the
#|   spatial-lag dot a little lower than the others."
ci <- function(b, se) c(b, b - 1.96 * se, b + 1.96 * se)
hp$lp <- lp
hp$li <- li
fe_raw <- plm::plm(lp ~ li, data = hp, model = "within")
fe_2w <- plm::plm(lp ~ li, data = hp, model = "within", effect = "twoways")
sar_raw <- splm::spml(lp ~ li, data = hp, listw = usalw, model = "within",
                      lag = TRUE, spatial.error = "none")
# splm's vcov() carries no names; the summary tables do.
se_table <- function(fit) {
  s <- summary(fit)
  if (is.null(s$CoefTable)) s$coefficients else s$CoefTable
}
slope <- function(fit, name) ci(se_table(fit)[name, 1], se_table(fit)[name, 2])
hp_est <- rbind(
  "FE, raw" = slope(fe_raw, "li"),
  "two-way FE, raw" = slope(fe_2w, "li"),
  "SAR-FE, raw" = slope(sar_raw, "li"),
  "FE, transformed" = slope(fe_hp, "li_t"),
  "SEM-FE" = slope(sem_hp, "li_t"),
  "SAR-FE" = slope(sar_hp, "li_t"),
  "SARAR-FE" = slope(sarar_hp, "li_t")
)
rice_fits <- list(within = fe_rice, SEM = sem_rice, SAR = sar_rice,
                  SARAR = sarar_rice, GM = gm_rice)
inputs <- c("log(seed)", "log(totlabor)", "log(size)")
rice_est <- do.call(rbind, lapply(inputs, function(v) {
  t(sapply(rice_fits, slope, name = v))
}))
op <- graphics::par(mfrow = c(1, 2), mar = c(4, 8, 2, 1))
k <- nrow(hp_est)
plot(hp_est[, 1], k:1, xlim = range(hp_est), yaxt = "n", pch = 16,
     xlab = "income elasticity", ylab = "", main = "House prices")
graphics::segments(hp_est[, 2], k:1, hp_est[, 3], k:1)
graphics::axis(2, at = k:1, labels = rownames(hp_est), las = 1, cex.axis = 0.8)
m <- nrow(rice_est)
cols <- rep(c("black", "firebrick", "steelblue"), each = 5)
plot(rice_est[, 1], m:1, xlim = range(rice_est), yaxt = "n", pch = 16,
     col = cols, xlab = "elasticity", ylab = "", main = "Rice farms")
graphics::segments(rice_est[, 2], m:1, rice_est[, 3], m:1, col = cols)
graphics::axis(2, at = m:1, labels = paste(rep(c("seed", "labour", "land"),
                                                each = 5), names(rice_fits)),
               las = 1, cex.axis = 0.7)
graphics::par(op)
```

**Read it.**

- **House prices, error.** λ = 0.69 with a standard error of 0.02; the
  income slope is 0.86 (0.06), down from the CCEP 1.20. *So:* a state's
  shock is two thirds shared with its neighbours, and once that sharing
  is weighted for, a 10% rise in income goes with a 9% rise in prices
  rather than 12%.
- **House prices, lag.** ρ = 0.65, slope 0.72. *So:* if the mechanism is
  that neighbours' prices move prices, the own-income slope is smaller
  still, and the [effects section](index.qmd#sec-sp-impacts) says what it
  adds up to.
- **House prices, both.** SARAR returns a lag of −0.52 and an error of
  0.87, opposite signs on the same W. *So:* with one W for both, the two
  parameters are barely separable [@elhorst2014, ch. 2], and the
  encompassing model does not settle the question the tests left open.
  The chapter reports the SEM and the SAR side by side.
- **House prices, the residual check.** The SEM residuals *u* still
  reject, z = 32 and p = 0.002, as they must. The filtered innovations do
  not, z = 0.5 and p = 0.54. *So:* one λ on this W accounts for all the
  spatial correlation the shuffle could find.
- **Rice, error.** λ = 0.79 (0.02); slopes 0.13, 0.25 and 0.54 against the
  within fit's 0.21, 0.29 and 0.50, and tighter. The filtered innovations
  give a local CD of −0.5 and a shuffle p of 0.04. *So:* with four fifths
  of a shock shared by the village, the GLS weighting moves the seed
  slope by a third; the within fit was consistent but far from efficient,
  and one λ captures the village shock with only a trace left.
- **Rice, lag and both.** SAR gives ρ = 0.51 and shrinks the slopes; SARAR
  puts 0.24 on the lag and 0.71 on the error, and the LR test of the lag
  given the error is in the [random-effects section](index.qmd#sec-sp-re).
  *So:* the error carries the dependence, as the robust tests said.
- **Rice, GM.** λ = 0.78 and slopes within 0.001 of the ML ones, with no
  standard error on λ. *So:* the two estimators agree on this panel, and
  GM is the one to keep when the errors are not normal or not
  homoskedastic.
- **Misleads (the raw panel).** A spatial lag fitted to raw log prices
  with state effects gives ρ = 0.68 and an income slope of 0.15; with
  two-way effects, 0.60 and 0.70. *So:* the national force reads as
  geography, and the slope swings by a factor of five with how that force
  is handled. Space comes *after* the factor, never instead of it.

::: {.callout-note appearance="simple"}
**Next.** Can the unit effect be left in the error, with the precision
that buys ([random effects](index.qmd#sec-sp-re))? On the house-price
panel the question is moot, and that section says why.
:::

::: {.callout-note collapse="true"}
## Recipe

```r
# f: a formula; d: a pdata.frame; listw: from spdep::mat2listw(w, "W").
splm::spml(f, data = d, listw = listw, model = "within",
           lag = FALSE, spatial.error = "b")        # SEM
splm::spml(f, data = d, listw = listw, model = "within",
           lag = TRUE, spatial.error = "none")      # SAR
splm::spml(f, data = d, listw = listw, model = "within",
           lag = TRUE, spatial.error = "b")         # SARAR
splm::spgm(f, data = d, listw = w, model = "within",
           lag = FALSE, spatial.error = TRUE)       # SEM by GM
u <- splm_residuals(fit, plm::index(d))             # back in panel order
e <- spatial_filter(u, w, coef(fit)[["rho"]], plm::index(d))
plm::pcdtest(e, w = w)                              # the residual check
```
:::

::: {.callout-caution collapse="true"}
## Pitfalls

**The names are the wrong way round.** In a `spml(model = "within")` fit,
`coef()` and the printed table call the spatial *lag* parameter `lambda`
and the spatial *error* parameter `rho`. @pesaran2015 [eqs. 30.1, 30.4]
and this book use ρ for the lag and λ for the error. Random-effects fits
name them the same way but store the lag in `$arcoef` and the error in
`$errcomp`. @millo2012 printed the reverse of today's package. Read a
parameter by what it does, never by its label.

**Residuals come back in the wrong order.** `spml` and `spreml` stack the
panel by period, then unit; the `pdata.frame` is stacked by unit, then
period. `splm_residuals()` puts them back; without it, every residual
test above would run on a scrambled panel.

**The within fit is a correction, not the transformation.** `spml` demeans
and then corrects the error variance after the fact, Lee and Yu's second
route [@lee2010, sec. 3.2; @croissant2019, sec. 10.3.1]. The orthonormal
transformation of their first route, whose estimator is consistent when
either N or T grows, is not what runs here.

**No standard error for the GM λ.** `spgm` estimates λ from moment
conditions and reports no dispersion for it, so no significance test is
possible [@croissant2019, sec. 10.3.3.3]. With one regressor its
`summary()` also fails, which is why the house-price GM fit prints
`coef()` and `$rho` instead.

**Heteroskedastic errors.** Spatial maximum likelihood is inconsistent
under heteroskedasticity of unknown form [@lin2010; @pesaran2015, p. 807].
Test for it as the core chapter does
([heteroskedasticity](../05-core/index.qmd#sec-core-hetero)) and, if it is
there, keep the GM fit.

**A spatial Durbin model.** $WX$ terms are added through `splm::slag()`
in the formula of `spml`, or with `Durbin = TRUE` in `spgm`. They change
the [effects](index.qmd#sec-sp-impacts), not the estimation.
:::

::: {.callout-tip collapse="true"}
## Book check

@croissant2019 [sec. 10.3.1, Example 10.6] print λ = 0.7913 with standard
error 0.0249 and slopes 0.1342, 0.2505 and 0.5419 for the SEM fixed-effects
fit of the rice panel; @croissant2019 [sec. 10.3.3.3, Example 10.9] print
λ = 0.7807, σ²ν = 0.0801 and slopes 0.1346, 0.2508 and 0.5418 for the GM
fit, whose header reads "fixed effects GM model" because `model =
"within"` is the default. The `c03-4-rice-sem` and `c03-4-rice-gm` chunks
above are those runs. @millo2012 [sec. 5.2] print 0.5574 for the spatial
error of a SEM-FE fit on `Produc` with `usaww`, and 0.4553 and 0.0886 for
the SARAR-FE fit, where their `rho` is the *lag*:

```{r}
#| label: c03-4-book-check
data("Produc", package = "plm")
data("usaww", package = "splm")
prod_f <- log(gsp) ~ log(pcap) + log(pc) + log(emp) + unemp
prod_lw <- spdep::mat2listw(usaww, style = "W")
coef(splm::spml(prod_f, data = Produc, listw = prod_lw, model = "within",
                lag = FALSE, spatial.error = "b"))
coef(splm::spml(prod_f, data = Produc, listw = prod_lw, model = "within",
                lag = TRUE, spatial.error = "b"))
```
:::
