## Serial correlation too {#sec-sp-serial}

**Question.** Does a farm's shock carry over to its next season, once the
farm effect and the village shock are allowed for? And does a state's?

**Intuition.** Three things can sit in one error: a unit effect, a shock
shared with the neighbours, and a shock that lingers from last period, the
**serial correlation** that the [time chapter](../04-time/index.qmd) takes
up on its own. @baltagi2007serial show that testing for any one of the
three while ignoring the other two "leads to misleading results", and give
a joint test and three conditional ones, each for one symptom allowing
the other two. The encompassing model then estimates all three at once,
and its *t* statistics mimic the conditional tests with the magnitudes
added [@millo2014].

::: {.callout-tip appearance="simple"}
How the three error components combine into one covariance, and which of
the twelve error structures of `spreml` holds which: [fitting spatial
panels](theory-spatial-estimation.qmd).
:::

The full error of @baltagi2007serial [eq. 2], with ψ the serial
parameter:

$$
u_{it} = \mu_i + v_{it}, \qquad
v_{\cdot t} = \lambda W v_{\cdot t} + \varepsilon_{\cdot t}, \qquad
\varepsilon_{it} = \psi\, \varepsilon_{i,t-1} + \eta_{it}.
$$

**The tests and estimator.** `bsjktest` [@baltagi2007serial]: `J`, the
joint test; `C.1`, spatial error given the other two; `C.2`, serial
correlation given the other two; `C.3`, random effects given the other
two. `spreml(errors = "semsrre")` fits all three; `"semsr"` leaves the
random effect out.

**Run it, rice farms.** The four tests on the model with village and
season dummies, then the full fit, as @croissant2019 [sec. 10.4.2.1] run
them:

```{r}
#| label: c03-6-rice-j
splm::bsjktest(rice_f, data = RiceFarms, index = "id", listw = ricelw,
               test = "J")
```

```{r}
#| label: c03-6-rice-c1
splm::bsjktest(rice_f, data = RiceFarms, index = "id", listw = ricelw,
               test = "C.1")
```

```{r}
#| label: c03-6-rice-c2
splm::bsjktest(rice_f, data = RiceFarms, index = "id", listw = ricelw,
               test = "C.2")
```

```{r}
#| label: c03-6-rice-c3
splm::bsjktest(rice_f, data = RiceFarms, index = "id", listw = ricelw,
               test = "C.3")
```

```{r}
#| label: c03-6-rice-semsrre
semsrre_rice <- splm::spreml(rice_f, data = rice, w = riceww, lag = FALSE,
                             errors = "semsrre")
round(summary(semsrre_rice)$ErrCompTable, 4)
```

**Run it, house prices.** On the transformed panel, with the random
effect left out of the fit for the reason the [previous
section](index.qmd#sec-sp-re) gave:

```{r}
#| label: c03-6-hp-tests
hp_df <- data.frame(state = hp_idx[[1]], year = hp_idx[[2]],
                    lp_t = hp$lp_t, li_t = hp$li_t)
splm::bsjktest(lp_t ~ li_t, data = hp_df, index = c("state", "year"),
               listw = usalw, test = "C.1")
splm::bsjktest(lp_t ~ li_t, data = hp_df, index = c("state", "year"),
               listw = usalw, test = "C.2")
```

```{r}
#| label: c03-6-hp-semsr
semsr_hp <- splm::spreml(lp_t ~ li_t, data = hp, w = usaw49, lag = FALSE,
                         errors = "semsr")
round(summary(semsr_hp)$ErrCompTable, 4)
```

What ψ measures, unit by unit: the correlation of this period's filtered
innovation with last period's, for every state and every farm:

```{r}
#| label: c03-6-picture
#| fig-cap: "Lag-one autocorrelation of each unit's innovations from the
#|   SEM fixed-effects fit, states (left) and farms (right). Thick bar: the
#|   ψ of the encompassing fit. Dashed line at zero."
#| fig-alt: "Two strip charts of points on a vertical axis from minus one
#|   to one. The state points cluster between 0.5 and 0.9 with the bar
#|   near 0.8. The farm points scatter widely around zero, with the bar a
#|   little above zero near 0.13."
lag1 <- function(e, index) {
  m <- tapply(e, list(index[[1]], index[[2]]), identity)
  apply(m, 1, function(x) stats::cor(x[-1], x[-length(x)]))
}
ac <- list(states = lag1(hp$eps_sem, hp_idx),
           farms = lag1(rice$eps_sem, rice_idx))
psi <- c(semsr_hp$errcomp[["psi"]], semsrre_rice$errcomp[["psi"]])
graphics::stripchart(ac, vertical = TRUE, method = "jitter", pch = 16,
                     col = "grey50", ylim = c(-1, 1),
                     ylab = "lag-one autocorrelation, by unit")
graphics::segments(c(0.8, 1.8), psi, c(1.2, 2.2), psi, lwd = 4)
graphics::abline(h = 0, lty = 2)
```

**Read it.**

- **Rice, each given the other two.** C.1 = 372 (spatial) and C.3 = 76
  (farm effect) reject hard; C.2 = 11.9 with p = 0.0006 (serial) rejects
  less hard. The full fit gives φ = 0.25, ψ = 0.13 and λ = 0.61, each with
  a *t* statistic that echoes its test. *So:* the village shock is the
  strongest thing in the error; a farm carries about an eighth of last
  season's shock into this one, real but small.
- **House prices.** C.2 = 216; ψ = 0.82 with a standard error of 0.015,
  and almost every state's innovations correlate above 0.5 with last
  year's. *So:* the static two-stage model is incomplete in time. A state's
  shock is four fifths still there a year later, which is what @holly2010
  model with lagged prices, and what the [time
  chapter](../04-time/index.qmd) takes up for this panel.
- **Misleads (the joint test).** J = 320 on rice. *So:* it rejects
  whenever any of the three is present and says nothing about which; read
  it last, if at all.

::: {.callout-note appearance="simple"}
**Next.** Read the spatial lag fits as effects on the map rather than as
coefficients ([direct and indirect effects](index.qmd#sec-sp-impacts)).
Serial correlation of this size in house prices sends the panel to
[time dependence](../04-time/index.qmd) before any slope is reported.
:::

::: {.callout-note collapse="true"}
## Recipe

```r
# f: a formula; df: a data.frame with the two index columns; d: a
# pdata.frame; w: the matrix; listw: spdep::mat2listw(w, "W").
splm::bsjktest(f, data = df, index = c("id", "t"), listw = listw, test = "C.2")
splm::spreml(f, data = d, w = w, lag = FALSE, errors = "semsrre")
splm::spreml(f, data = d, w = w, lag = FALSE, errors = "semsr")
```
:::

::: {.callout-caution collapse="true"}
## Pitfalls

**Two interfaces, two spellings.** `bsjktest` takes a plain `data.frame`
with `index =`; `spreml` takes the matrix as `w`, where `spml` takes a
`listw`. Mixing them up fails late and unhelpfully.

**Time.** The full rice fit takes about six seconds and the C.2 test about
four; on a panel of 170 units they take longer. Quarto's `freeze` keeps
them from running on every render.

**ψ is a nuisance parameter here, not a model.** A serial correlation of
0.8 in the errors is a sign of omitted dynamics, not a feature to keep.
The [time chapter](../04-time/index.qmd) owns that question.
:::

::: {.callout-tip collapse="true"}
## Book check

@croissant2019 [sec. 10.4.2.1, Example 10.15] print J = 319.5, C.1 =
371.5, C.2 = 11.894431 with p = 0.000563 and C.3 = 75.8, and
[Example 10.16] φ = 0.2500, ψ = 0.1250 and λ = 0.6136 for the full fit.
The rice chunks above are those runs.
:::
