## Direct and indirect effects {#sec-sp-impacts}

**Question.** Income rises 10% in Ohio and nowhere else. How much do
Ohio's house prices rise, how much do its neighbours' rise, and how much of
Ohio's rise came back through them?

**Intuition.** With $\rho W y$ on the right-hand side, a change in Ohio's
income moves Ohio's price by β, which moves Pennsylvania's price by ρ
times its share of Pennsylvania's neighbours, which moves Ohio's price
again. The chain never ends, and the model sums it: the reduced form is
$y = (I - \rho W)^{-1}(X\beta + u)$, so the effect of regressor *k* on
every unit's outcome is a whole N × N matrix. @lesage2009 [ch. 2] read it
in three numbers. The average of its diagonal is the **direct** effect, a
unit's own response, feedback included. The average of its row sums is
the **total** effect of raising $x_k$ everywhere. The difference is the
**indirect** effect, what spills to other units. β is none of the three.

::: {.callout-tip appearance="simple"}
Why the coefficient of a spatial lag model is not a marginal effect, how
the three averages are built, and why a spatial error model has no
indirect effects: [direct and indirect
effects](theory-direct-indirect-effects.qmd).
:::

For regressor *k* with slope $\beta_k$ [@lesage2009, sec. 2.7]:

$$
S_k = (I - \rho W)^{-1} \beta_k, \qquad
\text{direct} = \frac{1}{N}\operatorname{tr}(S_k), \qquad
\text{total} = \frac{1}{N}\, \iota' S_k\, \iota,
$$

and with a row-standardised W the total effect collapses to
$\beta_k / (1 - \rho)$, the spatial multiplier.

**The computation.** `sar_impacts()` builds $S_k$ and the three averages;
`sar_impacts_mc()` draws ρ and β from the fit's asymptotic normal
distribution and recomputes them 200 times for a standard error
[@lesage2009, sec. 5.2]. `splm::impacts()` does the same, when it runs;
see Pitfalls.

**Run it.** On the spatial lag fits of the [fixed-effects
section](index.qmd#sec-sp-fe). The lag parameter is `coef()[["lambda"]]`
in **splm**'s naming:

```{r}
#| label: c03-7-hp
rho_hp <- coef(sar_hp)[["lambda"]]
sar_impacts(rho_hp, coef(sar_hp)["li_t"], usaw49)
coef(sar_hp)[["li_t"]] / (1 - rho_hp)
```

```{r}
#| label: c03-7-hp-mc
set.seed(1)
mc_hp <- sar_impacts_mc(sar_hp, usaw49, reps = 200)
round(mc_hp$se, 3)
```

```{r}
#| label: c03-7-rice
rho_rice <- coef(sar_rice)[["lambda"]]
sar_impacts(rho_rice, coef(sar_rice)[inputs], riceww)
set.seed(1)
mc_rice <- sar_impacts_mc(sar_rice, riceww, reps = 200)
round(mc_rice$se, 3)
```

Ohio's column of $S$, drawn on the map, and the three effects of every
regressor with their intervals:

```{r}
#| label: c03-7-picture
#| fig-cap: "Left: the effect of a rise in Ohio's log income on each
#|   state's log house price, the Ohio column of (I − ρW)⁻¹β, darker for
#|   larger. Right: direct, indirect and total effects of income on house
#|   prices and of seed, labour and land on rice output, with 95% Monte
#|   Carlo intervals."
#| fig-alt: "A map of the contiguous United States with Ohio darkest, its
#|   neighbours dark grey, their neighbours lighter, and the West almost
#|   white. Beside it, four rows of three dots with whiskers: for income
#|   the total dot sits far to the right of the direct one; for the three
#|   rice inputs the indirect dot sits level with the direct one."
S_hp <- solve(diag(49) - rho_hp * usaw49) * coef(sar_hp)[["li_t"]]
est <- rbind(income = mc_hp$estimate, mc_rice$estimate)
se <- rbind(mc_hp$se, mc_rice$se)
rownames(est) <- rownames(se) <- c("income", "seed", "labour", "land")
graphics::layout(matrix(1:2, 1), widths = c(1.3, 1))
graphics::par(mar = c(0, 0, 2, 0))
plot(sf::st_geometry(us), col = shade(S_hp[, "OHIO"]),
     main = "Effect of Ohio's income")
graphics::par(mar = c(4, 6, 2, 1))
y <- rep(4:1, each = 3) + rep(c(0.2, 0, -0.2), 4)
cols <- rep(c("black", "firebrick", "steelblue"), 4)
plot(t(est), y, xlim = range(c(est - 1.96 * se, est + 1.96 * se)),
     yaxt = "n", pch = 16, col = cols, xlab = "effect", ylab = "",
     main = "Direct, indirect, total")
graphics::segments(t(est - 1.96 * se), y, t(est + 1.96 * se), y, col = cols)
graphics::axis(2, at = 4:1, labels = rownames(est), las = 1)
graphics::legend("bottomright", c("direct", "indirect", "total"),
                 col = c("black", "firebrick", "steelblue"), pch = 16,
                 bty = "n")
graphics::layout(1)
```

**Read it.**

- **House prices.** β = 0.72; direct 0.83, indirect 1.19, total 2.01,
  which is $0.72 / (1 - 0.65)$. *So:* a 10% rise in every state's income
  raises prices by 20%, and more than half of that arrives through the
  neighbours. Ohio's own rise is 8.3%, of which 1.1 points came back
  through Pennsylvania, Michigan and the rest.
- **Rice.** For land, β = 0.50; direct 0.51, indirect 0.52, total 1.03.
  *So:* a farm's own land raises its output by half a per cent per per
  cent, with feedback adding 2% of that, and as much again spills over the
  village.
- **Misleads (β as the elasticity).** Reporting 0.72 as the income
  elasticity of house prices misses two thirds of the total effect; for
  rice land it halves it. *So:* a paper that fits a spatial lag and reads
  β has not reported an effect at all.
- **A caveat on both.** These are readings of the *lag* model. For rice
  the tests chose the error model, under which β is the whole effect and
  the indirect effects above do not exist; for house prices the choice
  between lag and error stayed open. The effects are what a lag model
  claims, not what the data proved.

::: {.callout-note appearance="simple"}
**Next.** What the chapter can only state, not test
([limits](index.qmd#sec-sp-limits)).
:::

::: {.callout-note collapse="true"}
## Recipe

```r
# fit: splm::spml(..., model = "within", lag = TRUE); w: the matrix.
rho <- coef(fit)[["lambda"]]               # the lag, in splm's naming
sar_impacts(rho, coef(fit)[-1], w)
sar_impacts_mc(fit, w, reps = 200)         # Monte Carlo standard errors
```
:::

::: {.callout-caution collapse="true"}
## Pitfalls

**`splm::impacts()` does not run.** With splm 1.6-5 and spatialreg 1.4-3
it fails on every within fit, for every `type`:

```{r}
#| label: c03-7-impacts-fail
#| error: true
splm::impacts(sar_rice, listw = ricelw, time = 6)
```

The hand computation above is the same arithmetic, and the Book check
compares it with the cross-sectional machinery that does run.

**N × N inverses.** `type = "mult"` inverts a dense N × N matrix; at a few
thousand units that swaps. The trace approximations of `spatialreg::trW`
with `type = "MC"` or `"moments"` avoid it; `"moments"` needs a symmetric
W, or the similar symmetric form of a row-standardised one.

**No indirect effects in a spatial error model.** With the spatial term in
the error, $\partial y / \partial x_k' = \beta_k I$: the slope is the
effect, and nothing spills.

**A Durbin term changes the effects.** With $WX$ in the model the direct
and indirect effects of $x_k$ involve both $\beta_k$ and the coefficient
on $Wx_k$, and the shortcut $\beta_k / (1 - \rho)$ no longer holds
[@lesage2009, sec. 2.7].
:::

::: {.callout-tip collapse="true"}
## Book check

No published number: `?spml` shows its `impacts()` call on `Produc`
commented out, and @croissant2019 predate the method in **splm**. The hand
computation is checked instead against `spatialreg::impacts()` on a
cross-sectional spatial lag fit, one growing season of the rice panel,
where that package's method is documented and runs:

```{r}
#| label: c03-7-book-check
season1 <- RiceFarms[RiceFarms$time == 1, ]
cs <- spatialreg::lagsarlm(rice_f0, data = season1, listw = ricelw)
spatialreg::impacts(cs, listw = ricelw)
sar_impacts(cs$rho, coef(cs)[inputs], riceww)
```

Three numbers per input, identical to the printed digits.
:::
