## Limits {#sec-sp-limits}

Three questions this chapter cannot answer with a published number. The
first gets a simulation, as the [time
chapter](../04-time/index.qmd#sec-ur-traps) does: evidence about this code
on one design, and nothing more. The other two get a statement of what is
missing and what to write instead.

### Heterogeneous slopes {#sec-sp-heterogeneous}

**Question.** If every state has its own income elasticity, is the pooled
fixed-effects slope still right, and is its standard error?

**Intuition.** Let the slope vary, $\beta_i = \beta + \eta_i$, and keep the
spatial error. @pesaran2011 show that both the fixed-effects slope and the
mean-group slope, the average of the N unit-by-unit slopes met in the
[CCE section](../02-cross-dependence/index.qmd#sec-cce), stay consistent
for β: spatial dependence is weak, and weak dependence averages out. What
does not survive is the fixed-effects standard error. The part of the
error that is $\eta_i x_{it}$ is heteroskedastic and, with the spatial
shocks, correlated across units; the conventional formula assumes neither.
The mean-group standard error, the dispersion of the N slopes, is built
for the first problem. Driscoll–Kraay standard errors, met in the core
chapter ([robust errors](../05-core/index.qmd#sec-core-robust-se)), are
built for the second, but only as T
grows: with T small, "the problem of consistent nonparametric covariance
matrix estimation is much less tractable" [@baltagi2021, p. 394].

**Run it.** 49 units on a 7 × 7 grid with `rook_w(7)`, 20 periods, slopes
drawn around 1 with standard deviation 0.3, and shocks that spill over
with strength λ. First one draw at λ = 0.8:

```{r}
#| label: c03-8-one-draw
W <- rook_w(7)
set.seed(1)
d <- sim_spatial_slope_panel(W, lambda = 0.8)
fe <- plm::plm(y ~ x, data = d, model = "within")
mg <- plm::pmg(y ~ x, data = d, model = "mg")
c(FE = coef(fe)[["x"]], MG = coef(mg)[["x"]])
c(FE_conventional = sqrt(vcov(fe)["x", "x"]),
  FE_DriscollKraay = sqrt(plm::vcovSCC(fe)["x", "x"]),
  MG = sqrt(vcov(mg)["x", "x"]))
```

Then 200 draws at three values of λ, counting how often each 95%
interval covers the true mean slope of 1:

```{r}
#| label: c03-8-coverage
# One replication: the three intervals above, kept for whether they cover 1.
covers <- function(d) {
  fe <- plm::plm(y ~ x, data = d, model = "within")
  mg <- plm::pmg(y ~ x, data = d, model = "mg")
  inside <- function(b, se) abs(b - 1) <= 1.96 * se
  c(FE_conventional = inside(coef(fe)[["x"]], sqrt(vcov(fe)["x", "x"])),
    FE_DriscollKraay = inside(coef(fe)[["x"]],
                              sqrt(plm::vcovSCC(fe)["x", "x"])),
    MG = inside(coef(mg)[["x"]], sqrt(vcov(mg)["x", "x"])))
}
set.seed(1)
lambdas <- c(0, 0.4, 0.8)
coverage <- sapply(lambdas, function(l) {
  rowMeans(replicate(200, covers(sim_spatial_slope_panel(W, l))))
})
colnames(coverage) <- paste0("lambda_", lambdas)
coverage
```

```{r}
#| label: c03-8-picture
#| fig-cap: "Coverage of the nominal 95% interval for the mean slope, by
#|   estimator and spillover strength λ, from the 200 draws above. Dashed:
#|   the nominal 0.95."
#| fig-alt: "Three lines over lambda equal to 0, 0.4 and 0.8. The
#|   mean-group line sits on the dashed 0.95 line at the first two values
#|   and dips to 0.91 at the third. The conventional fixed-effects line
#|   runs between 0.75 and 0.85, and the Driscoll-Kraay line below it,
#|   between 0.68 and 0.78."
cols <- c("firebrick", "darkorange", "black")
graphics::matplot(lambdas, t(coverage), type = "b", pch = 16, lty = 1,
                  col = cols, xlab = "spillover strength, lambda",
                  ylab = "coverage of the 95% interval", ylim = c(0.5, 1))
graphics::abline(h = 0.95, lty = 2)
graphics::legend("bottomleft", rownames(coverage), col = cols, pch = 16,
                 lty = 1, bty = "n")
```

**Read it.**

- **Both slopes are centred.** FE and MG land within 0.02 of each other
  in the draw above, and the coverage of the mean-group interval does not
  drift with λ until the strongest spillover. *So:* @pesaran2011 hold on
  this design: spatial errors do not bias either estimator of the mean
  slope.
- **The conventional interval is too short.** Its coverage is 0.80, 0.76
  and 0.85 against a nominal 0.95, and λ does not drive it. *So:* the
  damage is the slope heterogeneity, not the spatial shocks; the plain
  fixed-effects standard error overstates what is known at every λ.
- **Driscoll–Kraay is worse at T = 20.** Coverage 0.68 to 0.78. *So:* the
  estimator that is robust to spatial and serial dependence as T → ∞ is
  not robust at twenty periods; @baltagi2021 [p. 394] warns of exactly
  this, and the house-price panel's 29 years sit close to the edge.
- **The mean-group interval holds, until λ = 0.8.** Coverage 0.97 and
  0.97, then 0.91. *So:* when slopes may differ, report the mean-group
  slope with its own standard error; under strong spillover, the N slopes
  stop being N independent pieces of evidence and even that interval
  shortens.

::: {.callout-note appearance="simple"}
**Next.** With 29 years per state the mean-group estimator runs; with 6
seasons per farm it does not (four parameters from six observations).
Rerun this table at your own N, T and W before trusting any of the three,
and read the 200-draw rates with their noise of about ±0.03.
:::

### Dynamic spatial panels {#sec-sp-dynamic}

**Question.** Last year's price moves this year's, and last year's
neighbours' prices too. Can one model hold the time lag and the spatial
lag at once?

**What the theory says.** The model is [@pesaran2015, eq. 30.27]

$$
y_{\cdot t} = \alpha + \gamma\, y_{\cdot,t-1} + \rho\, W y_{\cdot t}
  + \lambda\, W y_{\cdot,t-1} + X_{\cdot t}\beta + u_{\cdot t},
$$

with the third term the **space-time diffusion**: a neighbour's shock
arrives a period late. It is stable when $|\gamma| + |\rho| + |\lambda| <
1$ for a W that is row *and* column standardised [@pesaran2015, p. 810].
@yu2008 give three regimes for its maximum-likelihood estimator: with T
large relative to N it is centred; with N proportional to T it is not,
and needs a bias correction; with N large relative to T it converges at
rate T only [@baltagi2021, pp. 406–407]. A panel of 170 regions over 20
years is in the third regime at best.

**What runs.** Nothing, in R. The estimator is in Stata's `xsmle`
[@belotti2017] and in Elhorst's MATLAB code [@elhorst2014]. What a reader
might try instead, a lagged outcome fed to `spml`, is shown here so that
it is not mistaken for the real thing:

```{r}
#| label: c03-8-naive-dynamic
hp_dyn <- data.frame(state = hp_idx[[1]], year = hp_idx[[2]],
                     lp_t = as.numeric(hp$lp_t),
                     lp_t_lag = as.numeric(plm::lag(hp$lp_t)),
                     li_t = as.numeric(hp$li_t))
hp_dyn <- plm::pdata.frame(stats::na.omit(hp_dyn), index = c("state", "year"))
dyn <- splm::spml(lp_t ~ lp_t_lag + li_t, data = hp_dyn, listw = usalw,
                  model = "within", lag = TRUE, spatial.error = "none")
summary(dyn)$CoefTable
sum(abs(coef(dyn)[c("lambda", "lp_t_lag")]))
```

**Read it.** The time lag and the spatial lag together stay inside the
stability bound, which is the one check that can be made. The coefficient
on the lagged outcome is biased downward by about $(1 + \gamma) / T$
[@baltagi2021, p. 188], the Nickell bias that the [dynamic
chapter](../06-dynamic/index.qmd#sec-dyn-bias) takes up, and the spatial
lag inherits
the regime problem above with no correction available. *So:* write "a
dynamic spatial specification was not estimated; no R implementation with
a verified target exists", and give the stability check as the only
number.

### Forecasting {#sec-sp-forecast}

**Question.** Does the spatial model predict next year's prices better
than the plain fixed-effects one?

**What the theory says.** @baltagi2006 derive the best linear unbiased
predictor for a random-effects panel with spatial errors [@baltagi2021,
eq. 13.27]: the forecast is $x'\hat\beta$ plus a term that borrows from
the unit's own past residuals *and* its neighbours', through
$(I - \lambda W)^{-1}$. With the unit effect fixed, or its variance zero,
that term drops out entirely [@baltagi2021, p. 408], and the spatial
model's forecast is the fixed-effects forecast with a better $\hat\beta$.
The two anchor texts then disagree about whether the spatial terms help
out of sample: on cigarette demand they gave the best root mean squared
error [@baltagi2021, p. 408]; on liquor demand they did not, except one
year ahead [@pesaran2015, Table 30.2].

**What runs.** No `predict` method for `splm` objects; `effects()`
extracts the fixed effects, and the spatial predictor has to be assembled
by hand from the equation. The question is answered on one's own panel,
by a rolling-origin comparison of root mean squared errors, not by a
citation.

::: {.callout-caution collapse="true"}
## Pitfalls

**Not implemented anywhere in this book's tools.** The joint, marginal and
conditional LM and LR tests of @debarsy2010 for the fixed-effects case;
the Mundlak-style test of the fixed-effects assumption; the
heteroskedasticity-robust GM estimators of @lin2010 and their panel
extension, since **sphet** [@piras2010] is cross-sectional only; the
moving-average and error-component forms of spatial error [@pesaran2015,
eqs. 30.5–30.6]; eigenvector filtering as a check on the SAR functional
form [@tiefelsdorf2007], which **spatialreg** implements for a
cross-section only. Each is a sentence in a paper's limitations, not a
substitute presented as the real thing.
:::

::: {.callout-tip collapse="true"}
## Book check

None, for any subsection. `.docs/panel-spat-PLAN.md` records no published
target for the Pesaran–Tosetti variance, and the dynamic and forecasting
targets live in Stata output on data that **R** does not ship.
:::
