---
title: "Spatial dependence"
---

When house prices jump in Nevada, Arizona's follow within the year, and
Maine's do not. The [previous chapter](../02-cross-dependence/index.qmd)
found that the 49 states share a national force, took it out with
cross-state averages, and measured what was left: an exponent α of 0.51,
dependence so weak that averaging over states washes it out the way it
washes out noise ([the exponent
α](../02-cross-dependence/index.qmd#sec-csd-exponent)). Weak is not gone.
The co-movement that survives can be *local*: each state with its
neighbours, and with nobody else.

**Spatial dependence** is correlation that follows a map. A unit moves with
the units next to it, and the link fades with distance. The map enters the
model as a **spatial weights matrix** W, with one row per state that says
who its neighbours are and how much each counts. The pictures below put the
two halves of the story side by side: the 2000 level of the house-price
index, which carries the national force, and the residual of the CCE
regression of the previous chapter, which does not. Clusters survive in
both.

```{r}
#| label: c03-0-setup
devtools::load_all(quiet = TRUE)
# pmg() and pcce() build a call to plm(), so plm must be attached, not just
# namespaced.
library(plm)
data("HousePricesUS", package = "pder")
data("usaw49", package = "pder")
hp <- plm::pdata.frame(HousePricesUS)
hp_idx <- plm::index(hp)
# style = "W" records that the rows already sum to one; without it the
# object carries style "M" and some methods refuse it.
usalw <- spdep::mat2listw(usaw49, style = "W")
us <- us_states_sf()
data("RiceFarms", package = "splm")
data("riceww", package = "splm")
rice <- plm::pdata.frame(RiceFarms, index = c("id", "time"))
rice_idx <- plm::index(rice)
ricelw <- spdep::mat2listw(riceww, style = "W")
```

```{r}
#| label: c03-0-maps
#| fig-height: 3.5
#| fig-cap: "The 49 states shaded by the house-price index in 2000 (left,
#|   1980 = 100) and by the residual of the CCEMG regression of log price on
#|   log income in 2000 (right). Darker is higher. The national force is in
#|   the left map and out of the right one; neighbours still look alike in
#|   both."
#| fig-alt: "Two maps of the contiguous United States. On the left, the
#|   Pacific coast and the Northeast are dark and the interior is light. On
#|   the right the shading is patchier, but dark and light states still come
#|   in clusters of neighbours rather than scattered at random."
ccemg <- plm::pmg(log(price) ~ log(income), data = HousePricesUS,
                  model = "cmg")
hp$e_cce <- as.numeric(residuals(ccemg))
shade <- function(v) grey(1 - 0.85 * (v - min(v)) / diff(range(v)))
in_2000 <- hp_idx[[2]] == "2000"
op <- graphics::par(mfrow = c(1, 2), mar = c(0, 0, 2, 0))
plot(sf::st_geometry(us), col = shade(as.numeric(hp$price[in_2000])),
     main = "Price index in 2000")
plot(sf::st_geometry(us), col = shade(hp$e_cce[in_2000]),
     main = "CCEMG residual in 2000")
graphics::par(op)
```

Two equations organise the chapter. A **spatial lag** puts the neighbours'
outcome into a unit's own equation; a **spatial error** says only that the
shocks are shared:

$$
y_{it} = \rho \sum_{j} w_{ij}\, y_{jt} + x_{it}'\beta + u_{it},
\qquad
u_{it} = \lambda \sum_{j} w_{ij}\, u_{jt} + \varepsilon_{it}.
$$

Here $w_{ij}$ is the weight of state *j* in state *i*'s row of W, ρ is the
lag parameter and λ the error parameter. This book keeps that naming
throughout; the software does not, and the chapter says so where it bites.
The two claims are different claims about the world, not two fits of one
curve: under a spatial error, the fixed-effects estimator of the core
chapter ([fixed or random](../05-core/index.qmd#sec-core-fe-re)) stays
consistent and loses precision; under a spatial lag it is biased
[@pesaran2015, p. 801].

**In this chapter.**

1. **Spatial weights.** Who counts as a neighbour, and the four checks a W
   must pass before any model uses it
   ([spatial weights](index.qmd#sec-sp-weights); theory in
   [spatial weights](theory-spatial-weights.qmd)).
2. **Does it follow the map?** Whether the dependence left after the
   national force is out sits between neighbours, and how to tell that from
   a factor in disguise ([does it follow the map](index.qmd#sec-sp-map);
   theory in [testing for spatial
   dependence](theory-spatial-dependence-tests.qmd)).
3. **Lag or error?** Which of the two equations the data ask for
   ([lag or error](index.qmd#sec-sp-lag-or-error); theory in
   [spatial model choice](theory-spatial-model-choice.qmd)).
4. **Fixed effects.** Fit both, read the spatial parameter, check the
   residuals ([fixed effects](index.qmd#sec-sp-fe); theory in
   [fitting spatial panels](theory-spatial-estimation.qmd)).
5. **Random effects.** When the unit effect can stay in the error, and the
   tests that say so ([random effects](index.qmd#sec-sp-re)).
6. **Serial correlation too.** The third thing an error can carry, and the
   model that holds all three
   ([serial correlation too](index.qmd#sec-sp-serial)).
7. **Direct and indirect effects.** Why a spatial lag coefficient is not a
   marginal effect ([direct and indirect
   effects](index.qmd#sec-sp-impacts); theory in
   [direct and indirect effects](theory-direct-indirect-effects.qmd)).
8. **Limits.** Heterogeneous slopes, dynamic spatial panels and forecasts:
   what this chapter can check and what it can only state
   ([limits](index.qmd#sec-sp-limits)).
9. **Summary** ([summary](index.qmd#sec-sp-summary)).

Two panels run through every step. House prices carry the story:
`HousePricesUS` [@holly2010] with the contiguity matrix `usaw49`, both in
**pder** [@croissant2019]. Rice farms carry the book checks: `RiceFarms`
[@druska2004], 171 farms in six villages of West Java over six growing
seasons, with the matrix `riceww` that calls two farms neighbours when they
share a village, both in **splm** [@millo2012]. Every estimator and test in
@croissant2019 [ch. 10] is printed on the rice panel, and with 171 units
over 6 periods it is the one canonical panel shaped like the ones this book
worries about: many units, few periods.

Every section has the shape of the [time chapter](../04-time/index.qmd):
**question**, **intuition**, the **test** or estimator in words, **run it**,
a picture, **read it**, a **next** box, and collapsed **recipe**,
**pitfalls** and **book check** boxes at the end.

{{< include 1-weights.qmd >}}

{{< include 2-map.qmd >}}

{{< include 3-lag-or-error.qmd >}}

{{< include 4-fe.qmd >}}

{{< include 5-re.qmd >}}

{{< include 6-serial.qmd >}}

{{< include 7-impacts.qmd >}}

{{< include 8-limits.qmd >}}

{{< include 9-summary.qmd >}}
