<!-- Test section. One test (or one family) per section. Header: a few plain
words. Code calls stay in the open. Recipe, pitfalls and book check sit in
collapsed boxes at the end of the section, so a first reading can skip them.
-->

## {{Short header}} {#sec-{{chapter}}-{{what}}}

**Question.** {{The question, in the running example's words, in one
sentence.}}

**Intuition.** {{One paragraph: how the test works, in plain words, before
any equation. Bold each new term where it is defined. Gloss any term from
another chapter in one line and link back by chapter and anchor.}}

<!-- Point to the theory note right after the intuition, e.g. a callout-tip
"Why this test behaves this way: [{{note name}}](theory-{{what}}.qmd)."
Keep the derivation there; this step keeps only the math a reader needs to
run and read the call. Several equations are fine when each one earns its
place, e.g. the test regression and the statistic built from it. -->

**The test.** {{Name}} [@key]. Null: {{in words}}. A rejection means
{{in words}}.

**Run it.** {{One sentence on the data.}}

```{r}
#| label: c{{NN}}-{{n}}-run
pkg::test(...)
```

<!-- Show it: a picture of the evidence behind the result: the data, a
simulated contrast, a distribution against its critical value, a threshold.
Vary the kind of picture across sections. Every figure has fig-cap and
fig-alt. -->

```{r}
#| label: c{{NN}}-{{n}}-picture
#| fig-cap: "{{What the picture shows, in one line.}}"
#| fig-alt: "{{What a reader would see, for a screen reader.}}"
```

**Read it.**

- **Rejects ({{case}}).** {{Number}}. *So:* {{implication in one intuitive
  sentence}}.
- **Does not reject ({{case}}).** {{Number}}. *So:* {{implication in one
  intuitive sentence}}.
- **Misleads ({{case}}, only if there is one).** {{Number}}. *So:* {{why,
  in one sentence}}.

::: {.callout-note appearance="simple"}
**Next.** {{Rejects: where to go. Does not reject: where to go.}}
:::

::: {.callout-note collapse="true"}
## Recipe

```r
# {{What the input is, e.g. y: a pseries}}
pkg::test(y, ...)
```
:::

::: {.callout-caution collapse="true"}
## Pitfalls

**{{Trap in bold}}.** {{What goes wrong and how to see it; a chunk when the
trap can be shown.}}
:::

::: {.callout-tip collapse="true"}
## Book check

{{The source and page, the numbers it prints, and the chunk that reproduces
them. "None" plus the reason when `.docs/panel-spat-PLAN.md` records no
target.}}
:::
