<!-- Test section. One test (or one family) per section. Header: a few plain
words. Code calls stay in the open; pitfalls and book checks go to the
chapter's "Under the hood" section, not here. -->

## {{Short header}} {#sec-{{chapter}}-{{what}}}

<!-- The question, in the running example's words, in one sentence. -->

<!-- How the test works: one paragraph of intuition, then the math that
tells the story. Several equations are fine when each one earns its place,
e.g. the test regression and the statistic built from it. Bold each new term
where it is defined. Gloss any term from another chapter in one line and
link back. -->

**The test.** {{Name}} [@key]. Null: {{in words}}. A rejection means
{{in words}}.

**Run it.** {{One sentence on the data.}}

```{r}
#| label: s{{n}}-{{m}}-run
pkg::test(...)
```

<!-- Show it: a picture of the evidence behind the result: the data, a
simulated contrast, a distribution against its critical value, a threshold.
Vary the kind of picture across sections. Every figure has fig-cap and
fig-alt. -->

```{r}
#| label: s{{n}}-{{m}}-picture
#| fig-cap: "{{What the picture shows, in one line.}}"
#| fig-alt: "{{What a reader would see, for a screen reader.}}"
```

**Read it.**

- **{{Case where it rejects}}.** {{Number}}. *So:* {{implication in one
  intuitive sentence}}.
- **{{Case where it does not reject}}.** {{Number}}. *So:* {{implication in
  one intuitive sentence}}.
- **{{Case where it misleads, only if there is one}}.** {{Number}}. *So:*
  {{why, in one sentence}}.

**Next.** {{Rejects: where to go. Does not reject: where to go.}}
